% Chapter 6 — State: The Three Homes (KLEPPMANN). Budget 5 pp. Discharges §7.
\section{State: The Three Homes}\label{ch:homes}

This chapter discharges constitution clauses C-7.1--C-7.5 (the three homes, the one-writer discipline, and sufficiency).

A unit's state is not one thing. Some of it is a property of the unit alone --- the
settlement level a future has printed, the fixings a variance swap has accrued, whether a
note's barrier has knocked --- the same fact for every holder. Some of it is a joint
property of the unit and a particular wallet --- the variation margin this holder applied
yesterday, the dividend that holder is owed. And some of it is the contractual text
itself --- a multiplier, a strike, a fixing schedule --- versioned and append-only. Three
kinds of fact, and the state model gives each its own home: \textbf{UnitStatus},
\textbf{PositionState}, and \textbf{ProductTerms}. With the balances of Chapter~3
these are the four components of the ledger's state, and the four kinds of change a
transaction bundles (Chapter~3) write them one for one.

\subsection{The three homes}

\textbf{ProductTerms} holds the versioned, append-only contractual terms of an
instrument: the future's multiplier, the option's barrier and payout, the variance swap's
strike and 252-fixing schedule. A new version is appended, never an edit; the current
terms are the latest version, and any past state is read against the version in force
then. Its facts are properties of the instrument's terms, and a corporate action that
changes those terms writes a new version (the split below).

\textbf{UnitStatus} holds the facts that are properties of the unit as a whole --- the
same for every holder: the last recorded settlement level, the running accrued sum of
squared returns, the live / triggered / expired lifecycle node --- including, for a
settlement-obligation unit (Chapter~\ref{ch:settlement}), its instructed / settled /
failed node, so that an instructed-but-unsettled trade keeps its settlement level in a
home like any other unit and needs no fourth home invented for it. It is a materialised
projection of the unit-state changes on the log, written by the instrument's contract as
its life unfolds.

\textbf{PositionState} holds the facts that are joint properties of a unit and a wallet:
this holder's last applied settlement level, this holder's recorded dividend entitlement.
A fact lands here exactly when it depends on which wallet holds the unit --- when one
unit-level event yields a different value to each holder.

None of the three holds a balance --- that is the coordinate vector of
Chapter~3, the transaction's fourth kind of change --- and none holds a
view. The keys distinguish them: ProductTerms and UnitStatus are keyed by the unit,
PositionState by the wallet-and-unit pair.

\begin{lstlisting}
data Homes = Homes
  { productTerms  :: Unit           -> [TermsVersion]  -- append-only; latest is current
  , unitStatus    :: Unit           -> UnitFacts       -- properties of the unit alone
  , positionState :: (Wallet, Unit) -> PosFacts        -- joint facts, a collection (below)
  }
-- each field is a fold of the log; discard and replay reconstructs it.
\end{lstlisting}

\medskip\noindent\textbf{PositionState holds a collection, not a slot.} One
\texttt{PosFacts} value per (wallet, unit) pair would be too few, because two ordinary
situations need more than one live fact on the pair at once.

\emph{Two dividends in flight.} ACME declares a dividend on 2026-06-02, payable 2026-06-23,
and a second on 2026-06-09, payable 2026-06-30 --- before the first has paid. On the two
record dates W-ALPHA's holding earns two entitlements, say \USD{15{,}000} and \USD{12{,}000},
and both are owed at the same time, each waiting for its own payment date. A holder carrying
several unpaid entitlements on one stock is a routine Tuesday, not an exotic case. The pair
(W-ALPHA, ACME) must therefore hold two live facts, and each must retire on its own payment
date, not the other's.

\emph{A futures lot traded today.} W-ALPHA holds a futures position marked to yesterday's
settlement level and today trades one more lot at a new price, before today's print. Today's
variation margin owes the old lots their margin from yesterday's level and the new lot its
margin from today's trade price. The pair must carry both references at once.

So the slot is not one fact. \texttt{PosFacts} is a \emph{collection} of a pair's live facts,
each keyed by the cause-derived identifier of Chapter~\ref{ch:machines} --- the identifier of
the firing that wrote the fact. Two firings on the same pair write two distinct entries, and
neither overwrites the other.

\begin{lstlisting}
type Txid     = Hash             -- the cause-derived identifier of Chapter 4 (writing firing)
type PosFacts = Map Txid PosFact -- a pair's live facts, ONE entry per writing firing
data PosFact
  = Owed         Money      -- a determined entitlement; retires when its payment firing pays it
  | SettledValue Integer    -- a futures mark: the aggregate  sum q_i * level_i , ONE per pair
-- retirement: an entry leaves the collection exactly when its obligation discharges.
\end{lstlisting}

\emph{The retirement rule.} An entry is live for precisely the interval between the firing
that determines it and the firing that discharges it. The record-date firing of the first
dividend writes an \texttt{Owed} entry under its own identifier; the payment-date firing pays
it and retires that entry, while the second dividend's entry --- written under a different
identifier --- stays live until its own payment date. The collection at any moment is exactly
the pair's outstanding obligations.

\emph{The futures specialisation.} The multi-lot case does not need one entry per lot. The
margined lots collapse into a single \texttt{SettledValue}: the aggregate settled value
$\sum_i q_i \cdot \text{level}_i$ carried as one integer, from which day~$t$'s variation
margin is $\text{settle}_t \times Q - \text{storedSettledValue}$ for total quantity
$Q=\sum_i q_i$ (Chapter~\ref{ch:contracts}). A new lot is folded in at its trade price, so the
old lots still mark from yesterday's level and the new lot from today's price --- the
multi-lot case falls out of one integer, which is what a clearing house actually keeps. This
is why the event-keyed collection is the primitive and the aggregate rides inside it: two
dividends in flight cannot be added into one number, because each discharges on its own date
and must retire alone, so the collection must be keyed per firing; the futures lots all
discharge against the same daily print, so they may be summed, and their sum is just one entry
in that same collection.

Two disciplines make the homes trustworthy, and both are checkable.

\begin{principle}[One fact, one home]\label{prin:one-home}
Every non-balance fact has exactly one home, determined by what the fact is a property of:
the unit alone (UnitStatus), the unit-and-wallet pair (PositionState), or the contract's
terms (ProductTerms). No fact is kept in two homes, so no two homes can disagree about it.
\end{principle}

\begin{invariant}[One legal writer]\label{inv:one-writer}
Every non-balance fact has exactly one legal writer --- the smart contract of the
instrument the fact belongs to --- and no other contract may write it. The Transaction
Executor enforces this at the door: a transaction that writes a fact outside its writer's
authority is refused. The invariant joins the structural catalogue of
Chapter~14 with constitution~\S 11's writer discipline.
\end{invariant}

\subsection{One fact, one home: the future exercises all three}

The daily life of a cash-settled future carries all three kinds of fact, so it visits all
three homes. FUT-IDX is a future on index IDX, multiplier~10, USD, settled-to-market;
W-ALPHA holds one contract. On day~1 the official close prints $1{,}000.00$, the Event
Monitor emits, and the future's contract fires. Three facts are in play, and each has one
home.

\begin{center}
\begin{tabular}{llll}
\toprule
Fact & Property of & Home & Legal writer \\
\midrule
Settlement print $1{,}000.00$        & the unit alone         & UnitStatus    & the print firing \\
W-ALPHA's applied level; VM $\USD{50}$ & the unit--wallet pair & PositionState & the print firing \\
Multiplier $10$                      & the contract's terms   & ProductTerms  & registration (v.\,1) \\
\bottomrule
\end{tabular}
\end{center}

The print is one number for every holder: it is a UnitStatus fact. The variation margin
depends on the level W-ALPHA last applied, a fact of the wallet-and-unit pair: it is a
PositionState fact, and the same print yields a different margin to a holder who entered
at a different level. The multiplier was fixed at registration in ProductTerms and changes
only by a new version. The print firing writes the first two facts in one transaction ---
two facts, two homes, one writer --- and reads yesterday's applied level back from
PositionState rather than holding it (Chapter~5).

\emph{Each fact has exactly one home, fixed by what it is a property of; the future,
carrying three kinds of fact, exercises all three homes.}

\subsection{The homes suffice}

The homes are not bookkeeping conveniences. Their reason for existing is a sufficiency
requirement, and it is the load the rest of the architecture rests on.

\begin{principle}[The homes suffice]\label{prin:sufficiency}
The three homes, together with the balances, carry all the information the smart contracts
require in order to execute. Every fact a firing must carry forward to a later firing is
written into a home by an earlier firing, so the later firing finds its entire context on
the record; if a contract would ever need to remember something, that something has a
home.
\end{principle}

This is the precise reason a contract is stateless and never remembers
(Chapter~5): statelessness is affordable only because every fact a firing
would want to remember already has a home the firing can read. The sufficiency is what
turns ``never remembers'' from an aspiration into a consequence.

Minimality is its other half. A home holds a \emph{fact} --- an input a later firing reads
--- never a \emph{view}. NAV, profit and loss, availability, and encumbrance are
projections, computed on demand from the homes and the log and never materialised into one
(Chapters~7 and~12). The homes carry every fact a contract
must read and no figure a valuation or a report recomputes --- which is exactly why no
store of ``realised PnL'' exists anywhere to be reconciled against the position record.

The variance swap exercises the requirement at full stretch.

\medskip\noindent\textbf{Episode --- the variance swap (T5): sufficiency, exercised.}
\emph{Trigger.} The 127th scheduled IDX close is recorded and the Event Monitor emits
VS-1's 127th fixing. VS-1 is a one-year OTC variance swap on IDX, 252 scheduled fixings
against the recorded IDX official closes --- the same series FUT-IDX settles on.
\emph{Contract.} VS-1's contract fires. To extend the realised variance it needs the whole
path so far: the previous close and the running sum of squared returns over fixings
$1,\dots,126$. It holds neither, and it is forbidden to.
\emph{What it finds on the record.} The previous close is a recorded observation. The
accrued statistic is a UnitStatus fact of VS-1 --- a property of the unit alone, the
realised path as the swap has seen it --- written by the 126th firing for exactly this
reader. At the mid-life checkpoint (fixing~126, half of the 252 fixings elapsed) it stands
at $A_{126}=1{,}805{,}000$, an integer at scale $10^{-8}$: the 126 accrued squared returns
compressed into one number. The firing reads $A_{126}$ and the previous close, forms the
127th return, adds its square, and writes $A_{127}$ into UnitStatus for the 128th firing.
\emph{The coupling.} The IDX close at fixing~126 is $1{,}005.00$ --- the same recorded
close FUT-IDX takes as its day-3 final settlement; fixings $124,125,126$ are the future's
three settlement days, closes $1{,}000$, $990$, $1{,}005$. One recorded observation, two
consumers: the future writes it as a settlement level into FUT-IDX's UnitStatus, the swap
folds it into VS-1's accrued statistic in VS-1's UnitStatus, and neither reads the other's
home.
\emph{Design point.} \emph{Firing~127 finds its entire context on the record --- 126
fixings compressed into one UnitStatus integer $A_{126}=1{,}805{,}000$ --- so a contract
that spans a year of history remembers nothing between firings. The homes suffice.}

\subsection{Two more facts finding their homes}

The dividend and the split add one home each: the entitlement is per-holder, the split is a
terms change.

\medskip\noindent\textbf{Episode --- the dividend (T3): an entitlement in PositionState.}
On 2026-05-15, the record date, ACME's contract fires and reads the owned positions as the
ledger then stands: W-ALPHA holds $10{,}000$ ACME. It records the entitlement
$10{,}000\times\USD{1.50}=\USD{15{,}000.00}$. A dividend entitlement is a property of the
unit-and-wallet pair --- the same $\USD{1.50}$ per share yields a different amount to each
holder --- so it lands in PositionState. The transaction is moveless: no cash moves on the
record date, but the position state now carries what W-ALPHA is owed, for the payment-date
firing to discharge (Chapter~5).
\emph{Design point.} \emph{An entitlement owed but unpaid is a fact with a home; the record
date writes it into PositionState movelessly, and the payment date reads it back.}

\medskip\noindent\textbf{Episode --- the split (T4): a new version in ProductTerms.}
On 2026-06-01 ACME's 2-for-1 split becomes effective. In one atomic transaction the holding
moves from $10{,}000$ to $20{,}000$ shares (a paired-leg move, Chapter~3) and
ACME's terms gain a new version: ProductTerms ACME version~2 carries the split, and the
recorded price frame reads $100.00$ as $50.00$ after it (the market data operator's work,
Chapter~8). A split is a property of the contract's terms, so
it lands in ProductTerms and nowhere else. The terms are not edited; a version is appended.
\emph{Design point.} \emph{Terms change only by appending a version; the split is one
ProductTerms fact, atomic with the move it accompanies, and the append-only discipline is
what lets any past state be read against the terms in force then.}

\subsection{Rebuildable by replay}

None of the three homes is an independent store. Each is a deterministic projection of the
immutable log, so it can be discarded and rebuilt at any time without loss.

\begin{principle}[Homes are projections]\label{prin:rebuildable}
Each home is a deterministic projection of the log --- ProductTerms of the terms changes,
UnitStatus of the unit-state changes, PositionState of the position-state changes. Any
home may be discarded and rebuilt by replaying the log, without loss of information, so no
home is a store that can drift from the record.
\end{principle}

This is the precise sense in which the homes are \emph{materialised projections} and not
second copies. A second copy is dangerous because two stores of one fact, maintained by
independent logic, eventually disagree --- the failure the framework exists to remove
(Chapter~1). A home cannot play that role: it has exactly one legal
writer (the one-writer invariant above), it holds each fact in exactly one place
(one fact, one home), and it is a pure fold of the one log everything else also
folds. Materialising the accrued statistic into UnitStatus is therefore not the
reintroduction of a store to reconcile; it is a cached read of the log, provably equal to
the log it came from. What is materialised is a fact a contract must read; what is never
materialised is a view a projection can recompute. Between them the homes carry everything
the contracts need and nothing they do not.

The accrued statistic is order-dependent: each firing reads the previous accrual and
extends it, so its projection is defined by an order, and that order is fixed. The
immutable log's fold order for a unit's fixings is the recorded observation index~$k$ ---
the fixing's valid-time coordinate, distinct from its transaction-time position, the point
at which its print was appended to the log (Chapter~\ref{ch:picture},
Section~\ref{sec:bitemporal}) --- not the order in which prints arrive; replay folds the
accrual in valid-time order and is deterministic. A late or out-of-order print is inserted
at its observation position~$k$, and the accumulator tail $A_{k},A_{k+1},\dots$ is
recomputed from $k$ forward --- identical to timely arrival because the fold is over the
ordered valid-time sequence,
while the downstream daily CCP-cash figures at positions after~$k$ --- the future's
variation margin on the same recorded closes, not anything the once-settled variance swap
pays --- are corrected by compensating transactions, not left as posted. Rebuilding a home is therefore a re-fold in
observation-index order, never a replay in arrival order.
