\section{Conclusion}\label{sec:conclusion}

\subsection{The thesis, now shown}

One claim carried this document: the ledger is the orchestrator of the output of
many smart contracts. The contracts run outside its trust boundary. Each is
triggered by an event, reads the declared data and stamped observations it is
allowed to see, works out what must change, and produces transactions. The ledger
runs none of that product-specific logic. It is four things and nothing more --- a
single door that admits or refuses each proposed transaction, an immutable log that
records what the door admits, a family of views computed from the log, and one
generic operator application at the read seam. Everything a product needs beyond
those four lives in its contract, on the far side of the boundary.

That was asserted at the start. The tour discharged it. Eight economically distinct
products crossed the same door: a cash equity through a split, a dividend forecast,
a cleared future's daily margin, a structured note, a special dividend, a spin-off,
an elective merger, and a securities loan. No two share a payoff, a lifecycle, or a
settlement convention. Yet none required product-specific logic inside the ledger;
the two generalisations the tour did force --- the aggregate weld and the custody
coordinate --- are product-agnostic widenings of an existing check and an existing
coordinate. Each differed only in what its contract read and what it
produced --- the moves, terms versions, and status writes it emitted --- and every
one of those crossed the identical door, landed in the identical log, and was read
back by the identical views. A structured note is not a special case of the machine;
it is the same machine handed a different contract. Adding the ninth product is
writing a ninth contract, not extending the ledger. The claim that the ledger holds
nothing product-specific is now a demonstrated fact, because eight products passed
through and the only generalisations they forced were product-agnostic.

Each stop also closed one design question, by showing the mechanism absorb a real
product with real numbers rather than by argument. The eight closes are the load the
thesis actually bears, and they are worth collecting in one place.

\subsection{Each stop, and the choice it settled}

\begin{center}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{@{}p{3.5cm}p{10.5cm}@{}}
\toprule
\textbf{Tour stop} & \textbf{Design choice it substantiated} \\
\midrule
Cash equity \& split
& Single-basis consumption composed in declared order: a quantity and a price meet
in one frame only, fixed by the declared boundary and not by arrival order. An
adjustment is a declared operator applied at read, never a restatement of stored
shares. \\

Dividend forecast
& A forecast enters as a stamped observation of a registered kind, one dimension per
component --- never as a model number the ledger must trust. The kind's invariance
witness catches a mis-declared component before it can mark the position. \\

Listed futures
& Variation margin is ordinary conserved transactions keyed per settlement day. There
is no privileged variation-margin engine; the door that carries a share purchase
carries a future's entire margin life, and initial margin sits outside as posted
collateral. \\

Structured note
& The model-agnostic wall: the pricing that decided the note's worth is never opened.
Two layers protect the record --- the door catches transactions that are broken,
recomputation from the log catches the ones that are merely wrong. \\

Special dividend
& The adjustment schedule's three layers --- class default, declared override,
designated discretion --- behind a totality gate that fails closed. Bespokeness is a
higher-precedence layer of declared data, not a branch in code. \\

Spin-off
& Cross-instrument transport is one atomic multi-unit transaction, and a mark is
carried only at the level the data supports --- the pair while only the pair is
stamped --- never fabricated per component from a price the notice could not contain. \\

Elective merger
& Holder choice enters as logged data. The per-holder weld cannot apply when holders
elect differently, so the door checks the aggregate weld instead; the contract stays
deterministic and replays to the same per-holder transactions. \\

Securities lending
& Generalise the coordinate, not the architecture. The same door, the same
conservation, and views as folds of the same log carry a loan's position and
entitlement mechanics, collapsing back to a scalar when ownership and possession
coincide. \\
\bottomrule
\end{tabular}
\end{center}

Read down the right column and one pattern is unmistakable. Every choice pushes
product-specific knowledge out of the ledger and into declared data a contract reads:
the split's frame, the forecast's dimensions, the special dividend's override, the
merger holder's election. The ledger's own conduct never varies across the eight. It
admits, records, and reads back --- the same three acts, on every product.

\subsection{What the whole design rests on}

The reason eight products could share one machine is that a small set of invariants
holds by construction rather than by care, and every product inherits them for free.
An unconserved transaction cannot be built, because a move is a transfer and has no
way to express a quantity appearing from nowhere; conservation is therefore not
checked but guaranteed. A transaction applies in full or not at all, so no product
can leave the log in a half-written state. The log is append-only, so a recorded fact
is never overwritten --- an adjustment is a new entry read through a declared operator,
never a restatement of an old one. There is one door, so there is one place where
well-formedness is decided and one place a proposal can be refused. Defaults fail
closed: an unregistered datum kind is unconsumable, a stale datum met with no declared
conversion is a typed refusal, an incomplete adjustment schedule blocks registration.
And the model-agnostic wall keeps every pricing decision on the contract's side of the
boundary, entering the ledger only as a stamped observation the door never interprets.
These are the properties that made the tour uneventful. Each is enforced once, at the
door or in the types, and no contract --- careful, careless, or hostile --- can reach
around them.

What remains is to state these commitments in the compact, formal register they
deserve, drawing the guarantees together as one synthesis.

\subsection{The formal synthesis}

Nothing here is new; each statement compresses a guarantee the preceding
sections established in plain terms.

\paragraph{Conservation is inherited from the generator.} A move subtracts a
quantity from one wallet and adds the same quantity to another: its net effect
on a unit's total across all wallets is zero, and a transaction, a finite list
of moves, nets a sum of zeros. Formally, net change per unit is a homomorphism
from move sequences under concatenation to quantities under addition, fixed by
its values on the generators.

\begin{proposition}\label{prop:syn-conservation}
Every move nets to zero, so the net-change map is the zero homomorphism --- it
factors through the trivial group --- and no expressible transaction can create
or destroy quantity.
\end{proposition}

\paragraph{The log is free, and every view is a fold out of it.} State holds no
information the event sequence does not: replay to any point rebuilds the exact
state that held then, and every view --- balance, valuation, profit and loss,
report --- is a fold of the one log, determined by what it does with a single
transaction. Two folds of one log cannot disagree about a fact recorded once;
the reconciliation guarantee of Section~\ref{sec:thesis} is this uniqueness.

\begin{proposition}\label{prop:syn-fold}
A view is the unique homomorphism from sequences of admitted transactions under
concatenation to state transformations under composition, extending its
per-transaction action; replay is therefore deterministic, independent of
where it is cut, and re-reads history rather than re-executing any producer.
\end{proposition}

\paragraph{Conversion is an arrow that exists or does not.} A datum crosses a
basis boundary only by the operator the boundary declared, from a closed menu;
omission declares nothing, and the crossing is refused.

\begin{proposition}\label{prop:syn-operators}
For each datum kind, between two bases on one chain there is at most one
conversion --- the composite of the declared operators between them, in
effective order --- and a datum crosses exactly when that composite exists;
fail-closed is the absence of the arrow.
\end{proposition}

\paragraph{A contract is a recomputable function.} A contract is a
deterministic, total function from the committed log prefix and its trigger to
one result --- a proposed transaction or a typed refusal. Structural soundness,
once admitted, is Proposition~\ref{prop:syn-conservation} plus the door's
checks; economic soundness is checkable because everything the function read
is in the log: two evaluations of one function on one input are equal.

\begin{proposition}\label{prop:syn-witness}
Economic correctness is the agreement of the recorded transaction with an
independent re-evaluation of the producing function on the same committed
prefix, compared in canonical form: a discrepancy locates a producer defect
exactly, and the bound is eventual detection and recorded repair, never
admission-time verification.
\end{proposition}

\paragraph{One arrow, three names.} A trade, a lifecycle event, and a
corporate action differ in what fires them and in what they read --- never in
shape. Each is one arrow from log prefix and trigger to a proposed transaction
or a typed refusal; every guarantee above is stated over that arrow's inputs
and output, never over which instance fired, and so holds for every instance,
including those yet to be written.

One record, freely generated by what the door admits; conservation the zero
homomorphism on its generators; every view a fold out of it; every conversion
the one arrow its chain declares; every producer a recomputable function into
it. Not five disciplines maintained in parallel: one structure, read five
ways.
