\section{The Spin-Off}\label{sec:tour-spinoff}

A spin-off moves value across instrument lines: one holding becomes two. This stop
shows the ledger doing that in a single transaction, and --- the harder part --- being
honest about a price it does not yet have. In November, ACME spins off a subsidiary.
Write the parent as $A$ and the two resulting lines as $A_1$ (the continuing ACME) and
$A_2$ (the new company). The book holds 1{,}000 $A$, marked \EUR{100}, so \EUR{100{,}000}.

\subsection{One transaction across two lines}

The confirmed notice says each share of $A$ entitles the holder to one $A_1$ and one
$A_2$. The corporate-action contract reads it and submits a single transaction that
issues 1{,}000 $A_1$ and 1{,}000 $A_2$ into the book's wallet against the 1{,}000 $A$
it resolves --- $A_1$ and $A_2$ each registered as units at their origins, with their
own datum kinds. Because it is one transaction, the two new lines appear together or
not at all; no reader can catch the book holding $A_1$ but not yet $A_2$, and
per-unit conservation holds on each line separately, each new share issued by a move
from a virtual wallet. This is cross-instrument transport, and it is one atomic
multi-unit transaction, not a sequence that some interleaving could observe half-done.

The transaction carries the entitlement --- the quantities. What it cannot carry is the
children's prices, and this is the point of the stop. No notice can contain a price
formed after it. In practice both lines often trade \emph{when-issued} in the days
before the distribution settles, so per-component prices may well exist ahead of the
first regular-way session; but a when-issued print is market data, not terms, and the
notice that issues the shares still does not contain it. Say the lines settle near
$A_1 = \EUR{80}$ and $A_2 = \EUR{20}$ once they trade regular-way. Whether the
continuing parent is modelled as a fresh unit $A_1$ at all is itself a choice --- ACME's
listing and ISIN usually survive the spin-off, so a deployment may keep the parent line
and register only the new child; this stop registers both to work the general two-line
case.

\subsection{A mark carried only as far as the data supports}

So on the morning after the spin-off, before the children have printed, what is the
holding worth? The stale \EUR{100} parent mark still exists, stamped in the parent's
basis. The declaration at the boundary is \texttt{Recompose} of $A$ onto the basket
$\{A_1 \mapsto 1,\ A_2 \mapsto 1\}$, and that basket datum was jointly stamped at the
successors' origins. It is consumable, but only jointly. It licenses a mark on the
\emph{pair}: the combined 2{,}000 shares --- 1{,}000 $A_1$ and 1{,}000 $A_2$ together
--- carry \EUR{100{,}000}, which is exactly the pre-spin value, phantom-free, so the
overnight risk of the combined position is readable at the open.

What it refuses is a mark on either child alone. To place a number on $A_1$ by itself,
one would have to divide the \EUR{100{,}000} between the two lines in some proportion
--- and that proportion is exactly the decomposition the notice never contained. Its
absence is what single-basis consumption refuses on: the per-component kinds stay
pending until fresh per-component observations arrive to supply it. So the combined pair
can be valued and its overnight risk read, but the two lines cannot yet be hedged or
margined against each other, because separating their risk needs the very split that is
still pending. That split is not refused forever, only until it is observed: the
issuer's cost-basis allocation, or the first regular-way session's volume-weighted
prices, later supplies it --- the allocation as attested declared data, the prices as
stamped observations of their own. The basket step is a bridge, not
a standing conversion; it retires the moment the children carry their own fresh stamps,
after which valuation proceeds component by component.

The discipline earns its keep against a tempting error. Suppose someone carries the
stale \EUR{100} onto $A_1$ as though it were $A_1$'s price --- treating the parent
number as passing through unchanged --- and then adds the fresh $A_2$ print:
\[
  \underbrace{1{,}000 \times 100}_{A_1,\ \text{stale, assumed}} +
  \underbrace{1{,}000 \times 20}_{A_2,\ \text{fresh}}
  = \EUR{120{,}000}.
\]
Against a true \EUR{100{,}000}, that is a phantom \EUR{20{,}000} --- and it is exactly
$A_2$'s value counted twice, once embedded in the stale parent price and once as the
fresh $A_2$ print. The ledger refuses it twice over. Identity is a declared operator,
never assumed, so there is no conversion carrying $A$ onto $A_1$ alone for the stale
mark to ride; and single-basis consumption refuses mixing a stale pair value beside a
fresh single-component print. Once both children have printed --- $A_1$ at \EUR{80},
$A_2$ at \EUR{20} --- the basket bridge retires and the mark is component-wise:
$1{,}000 \times 80 + 1{,}000 \times 20 = \EUR{100{,}000}$, each on its own chain. The
value crossed the boundary as a whole and came out a sum of parts only when the parts
had prices of their own.

\emph{Cross-instrument transport is one atomic multi-unit transaction, and a mark is
carried at the level the data actually supports --- the pair, while only the pair is
stamped --- never fabricated per component from a price the notice could not have
contained.}
