\section{An Elective Merger}\label{sec:tour-merger}

The hardest corporate action for a deterministic system is one where the holders
choose. Different holders of the same security take different consideration, so the
same event produces different moves for different wallets --- and yet the security has
one price series that every holder must read the same way. This stop, the last on the
ACME line, shows both halves handled by holder choice entering as ordinary data. The
following March, an acquirer $Q$ takes over ACME; write the target line as $T$, held
1{,}000 shares, marked \EUR{100}, split 500 to holder $A$ and 500 to holder $B$.

\subsection{Nothing moves until confirmation}

The announcement opens a messy window: terms proposed, elections open, proration
unknown. The design's answer to the window is to declare no operator for it. With no
operator, nothing adjusts --- no quote is transported, no strike moves, no number
changes --- and any attempt to consume a datum across the still-unresolved boundary
fails closed. The fail-closed reaches only consumption that would \emph{cross} the
announced boundary; it does not freeze the $T$ line. A fresh $T$ print taken in the
standing pre-merger basis and read in that same basis is no boundary crossing, so $T$
marks normally all through the arbitrage window --- only a datum trying to convert across
the unresolved terms is stopped. The world has said a merger is coming; the log records
that it is pending and computes nothing across it. This is the opposite of guessing, and
it is what lets the window be messy without corrupting anything.

Confirmation resolves it. The final terms are \EUR{102} cash or 2.04 $Q$ shares per
$T$ share, under a 60\% aggregate cash cap that here does not bind. $A$ elects cash,
$B$ elects stock, so half the book elects cash --- comfortably under the cap, which is
why nothing is prorated. The elections are not inferred; each holder's choice is a logged
instruction, read by the contract as declared data.

\subsection{Different moves, one fate, and a weld in aggregate}

On the entitlement plane the two holders get different moves, because they chose
differently: $A$'s 500 shares become $500 \times 102 = \EUR{51{,}000}$ cash, and $B$'s
500 shares become $500 \times 2.04 = 1{,}020$ $Q$ shares. On the quotation plane the
$T$ series has one fate, read identically by every holder: the confirmed blend of
\EUR{51} cash and 1.02 $Q$ per $T$ share, declared as the composite
$\texttt{Subst}(Q, 1.02) \circ \texttt{Shift}(-51)$ --- no new case: the composite is
one declared conversion, the pair $(f, c) = (1.02, 51)$. A stale \EUR{100} $T$ mark transports across it to
$(100 - 51)/1.02 = 2450/51$ per $Q$ share, an exact rational.

No single holder's moves match the blend --- that is what an election means --- so the
door cannot check the merger holder by holder against a uniform rule. It checks the
legs in aggregate at the confirmed $(f, c) = (1.02, 51)$: the quantity legs must sum
to $f$ times the total, $1{,}020 = 1{,}000 \times 1.02$, and the cash legs to $c$ times
the total, $\EUR{51{,}000} = 1{,}000 \times 51$. Both hold. This aggregate weld is a
distinct check reserved for a confirmed elective structure; it does not replace the
strict per-holder weld that governs a uniform action, which stands at full strength for
every non-elective boundary. Each holder's own allocation is then economic detail,
checked by recomputation against that holder's logged election and the confirmed
proration.

The value is continuous in aggregate to the minor unit. Before: $1{,}000 \times 100 =
\EUR{100{,}000}$. After: $1{,}020 \times 2450/51 + 51{,}000 = 49{,}000 + 51{,}000 =
\EUR{100{,}000}$, with no rounding residue, since $1{,}020/51 = 20$ exactly. The
per-holder differences are each measured against the one transported mark of $2450/51$
per $Q$ share, and each is exactly that holder's own elected consideration seen against
that mark: $A$'s \EUR{51{,}000} of cash stands \EUR{1{,}000} above the \EUR{50{,}000}
its 500 shares carried before, and $B$'s $1{,}020$ $Q$ shares, valued at the mark, come
to \EUR{49{,}000}, \EUR{1{,}000} below. The two differences sum to zero, but this is a
frame identity, not a transfer between the holders: no value passes from $B$ to $A$.
Each holder's consideration settles against the acquirer $Q$ --- $Q$ pays $A$'s cash and
issues $B$'s shares --- and because the cash cap does not bind, both elections are
satisfied in full, nothing prorated and nothing exchanged between the two. The
$\pm\,\EUR{1{,}000}$ is simply what each election is worth when both are marked at the
single blended price the $T$ line resolves to.

And because the boundary transported only what it observed, a fresh $Q$ price is
ordinary marking profit and loss, not a phantom. If $Q$'s next print is \EUR{50}, the
transported mark $2450/51$ per share leaves a gap of $50 - 2450/51 = 100/51$ per
share, and over 1{,}020 shares that is $1{,}020 \times 100/51 = \EUR{2{,}000}$ exactly
--- old information meeting a fresh mark, shown as market movement, never absorbed into
the merger operator. \emph{Because holders elect differently, no single holder's moves
match the confirmed blend, so the per-holder weld that governs a uniform action cannot
apply; what the door checks instead is the aggregate weld --- the quantity legs summing
to $f$ times the total, $1{,}020 = 1{,}000 \times 1.02$, and the cash legs to $c$ times
the total, $\EUR{51{,}000} = 1{,}000 \times 51$ --- the distinct check an elective
structure forces. That the contract stays deterministic and replays to the same
per-holder transactions holds here too, but it is the aggregate weld this stop turns
on.}
