% =====================================================================
%  Part I --- Merged Systemic Review (peer merge of three parallel reviews)
%  Reviewers: TuringAward (TA), MINSKY (MK), FORMALIS (FL).
%  Subjects: Constitution v1.41 + Market Data Manifesto 1.3 + Valuation
%  Manifesto 1.0 (incl. Part B, PARK-1), reviewed as a COMPOSED SYSTEM.
%  LaTeX FRAGMENT (section-level; no preamble). Assembly supplies
%  documentclass, amsmath, amssymb, geometry, hyperref.
%  Severity CRITICAL/HIGH/MED/LOW; TA's MAJOR->HIGH, MINOR->LOW.
%  Attribution in brackets; divergences preserved, never averaged.
% =====================================================================

\section*{Part I --- Merged Systemic Review}

\noindent\emph{Three reviewers read the three manifestos independently as one system (TA:
architecture/algorithms; MK: type discipline, illegal states; FL: cross-manifesto formal
coherence). Findings are attributed; divergences are kept as positions, never averaged.
Severity CRITICAL/HIGH/MED/LOW --- TA's MAJOR$\to$HIGH, MINOR$\to$LOW.}

\subsection*{1. Systemic verdict}

The three documents compose \emph{without a vicious logical cycle}: definitions and discharges
each flow one way [FL]. But they share one fault line --- the boundary at which a deterministic
function of the record becomes a \emph{stored} artifact --- and all three reviewers, apart,
ranked the collision there the gravest finding. A \textbf{version-layering inversion} compounds
it: the parent clause a child leans on (MD-16, Market Data 1.3) postdates and is uncited by that
child (Valuation 1.0, subordinate to Market Data 1.1) [FL]. The recurring defect beyond the
headline is \textbf{drift-by-relabeling at the seams between authors and versions} --- a
guarantee kept reading true by renaming its residue (``event-outcome,'' ``forbidden,''
``$\mathcal{D}\equiv$~VM-5,'' ``structurally impossible''). Correctness holds where one author
states a thing once (VM-8 extends MD-14 without re-proving; MD-13 defers the sandwich to VM-9);
it frays where documents meet. These are composition gaps and relabelings, not thesis errors.

\subsection*{2. Headline finding (unanimous, gravest)}

\noindent\textbf{H1 [all three --- gravest]. A ledger-computable number is stored on one side of
the framework and parked as unstorable on the other.} MD-16 records a gate decision's
\emph{computed functionals and percentiles} --- each a map from a derived state $m^\ast$ to a
number, a \emph{projection} by C-4.11's own definition --- as stored fields, while the Valuation
Manifesto \emph{parks} the identical act (storing $\sigma_{\mathrm{prod}}^2$, PARK-1) because ``a
projection stores nothing.'' MD-16 declares ``no collision with C-4.11 \dots\ an event-outcome,
not a projection.'' Three angles converged: \textbf{[TA]} a \emph{certified} article (Market
Data 1.3) silently depends on an unresolved sibling park --- rule C-4.11 literal and it breaks
retroactively; \textbf{[MK]} ``event-outcome'' \emph{is} the reclassification PARK-1 refused
without the owner, the CLAUDE.md\,\S1 worse failure (a guarantee kept true by renaming its
residue, eroding the reconciliation root C-1.2--1.3); \textbf{[FL]} silent non-conformance with
a blast radius the parking index hides --- and, surgically, recording the pass/fail \emph{bit}
is fine, only the numeric \emph{fields} materialise a projection.
\emph{Divergence preserved (fix):} FL's remedy needs \emph{no} amendment --- ground the
recording on MD-11/C-14.15 (a simulated path must capture its outputs) and drop ``not a
projection''; else fold the fields under PARK-1. TA/MK treat the whole recording as
materialisation, FL only the fields. \emph{Complementary [TA]:} C-2.8 (``the seed \dots\ every
path replays exactly'') itself overclaims against MD-11 (``a path that recorded only the seed
could not replay''); amend C-2.8 to match MD-11.

\subsection*{3. Advice --- ordered, merged (highest leverage first)}

\begin{enumerate}\setlength{\itemsep}{0pt}
  \item \textbf{[all three] Rule PARK-1 with \emph{both} documents on the table} --- its ruling
  governs VM PE-6 \emph{and} MD-16's recorded gate fields; ``nothing depends on storage'' is
  true only inside the Valuation Manifesto. Precedes any other ratification.
  \item \textbf{[all three] Legislate ``may a deterministic function of the record be stored?''
  once} --- one clause over C-4.11, PARK-1, MD-16 (PARK-1's three-condition materialised
  projection is the candidate), so no document decides it by private nomenclature.
  \item \textbf{[FL, TA] Freeze the parent text the children consumed, then re-rate} --- C-2.7
  and C-12.6 (reworded by v1.4/v1.41, both ratification-pending) carry MD-4/MD-5/MD-13/VM-2/VM-9;
  fix the stale headers (VM 1.0 cites parent files that do not exist).
  \item \textbf{[all three] Decide prevention-vs-detection at the Constitution} --- add a
  forward-pointer on MD-9 to MD-16's carve-out; align ``forbidden'' language to the actual
  mechanism (detection, recorded-and-flagged).
  \item \textbf{[FL, MK] Tighten the A5 discharge} --- make Gate 1 an interior-membership gate
  over the move's neighbourhood, or downgrade ``grounds'' to ``guarantees the endpoint''; wire
  VM-10's finite-shift states to MD-16's gates.
  \item \textbf{[MK] Complete the typed arrow algebra (C-3.6)} --- name the producer of the
  observation-recording and re-entered-observation arrows, or ratify observable-is-a-unit.
  \item \textbf{[FL, TA] Add a firing witness to every new invariant} (Gate 1/2, VM-6, VM-7) and
  state a \emph{liveness} commitment: every stale flag, broken chain, undecidable gate carries a
  deadline, owner, and fallback.
\end{enumerate}

\subsection*{4. Criticisms --- wrong or weakly argued today}

\begin{itemize}\setlength{\itemsep}{0pt}
  \item \textbf{C1 [MK, HIGH]. The recording arrow has no typed producer.} C-3.6 types only
  $\mathit{contract}:(\mathrm{Event},\mathrm{Ledger})\to\mathrm{Transaction}$; the
  observation-recording and re-entered-observation transactions have none --- MD-1 attributes
  recording to the Events Executor, which C-5.4 says ``never writes the ledger.'' The arrow
  crossing the trust boundary has unspecified producer, input type, and validation.
  \item \textbf{C2 [all three, HIGH$/$MED]. Prevention vs detection: two polarities, one
  unannotated.} MD-9 ``by detection, not admission'' vs MD-16 Gate 1 ``prevention, not
  detection,'' reconciled only inside MD-16 [MK C3, FL C6]. \emph{Preserved:} TA goes further ---
  Gate 1 prevents on an \emph{economic} property \emph{outside} the Transaction Executor, a
  third correctness category C-13 never sanctions (HIGH); MK/FL rate it MED and read MD-16 as
  honest (``a consumer over-trusts'').
  \item \textbf{C3 [FL, HIGH; corr.\ MK].\ Gate 1 discharges VM PE-1(A5) incorrectly.} Gate 1
  guarantees one endpoint $m^\ast\in\Theta_{\mathrm{AF}}$; A5 needs $\Sigma(\cdot;S)\in\mathcal{A}$
  on a \emph{neighbourhood} with \emph{interiority} --- gated at neither end (the origin surface
  is admissible only by MD-9 detection). \emph{Corr.\ [MK W5]:} a finite bump (VM-10, spot
  $+20\%$) falls between the infinitesimal discharge and the finite gate. \emph{Mitigant [FL]:}
  poles are recorded diagnostics, so nothing breaks silently; only the \emph{discharge claim} is
  false.
  \item \textbf{C4 [MK, HIGH]. NAV carries two types.} C-8.1/8.2 type NAV as a bare cash number;
  VM-2 says a valuation without coordinates ``is not one'' and may be stale/broken --- different
  guarantees in different documents (see W1).
  \item \textbf{C5 [FL, MED-HIGH]. MD-4's ``every observation bears three times'' fails for VM's
  own observations.} By C-2.7 an emitted event bears no monitor time; a re-entered observation
  (VM-1: every model-priced leg) is emitted, not observed. Fix: ``bears \emph{up to} three
  times.''
  \item \textbf{C6 [TA, HIGH]. C-2.2 is narrowed in the children without amending the parent.}
  Unconditional ``to the last minor unit'' is made conditional by MD-6/VM-2 (re-deriving a model
  mark ``needs the retained model and numerical environment \dots\ outside this scope''); every
  model-mark's bit-for-bit claim (VM-8, VM-11) rests on it. C-Auth.3 requires amending the
  Constitution first.
  \item \textbf{C7 [TA, HIGH]. ``Look-ahead is structurally impossible'' (MD-4) is not
  structural.} ``By default'' concedes a non-default: VM-2's legal decoupled query (position
  as-at yesterday, market cut today) inside a VM-11 trajectory reintroduces look-ahead ---
  a discipline to remember. \emph{Adjacent:} VM-11's ``valuation at every step'' vs
  gate-\emph{undecidable} holes [FL C5]; MD-4 fixes the as-of/as-at indexing C-12.1 left to the
  spec [MK C5].
  \item \textbf{C8 [all three, MED]. The world-map family is asserted, not constrained.} [TA]
  three names for one concept (\emph{dynamic}/\emph{shift}/$\mathcal{D}$) stress C-Auth.4; [FL]
  MD-16's ``VM-5~$\equiv\mathcal{D}$'' is containment not identity, dropping the correlation
  channel; [MK] a surface's $\mathcal{D}$ and its vols' per-datum dynamics must agree
  \emph{nowhere}. Also stale subordination: A5's only discharge (MD-16) postdates and is uncited
  by VM 1.0 [FL C9].
\end{itemize}

\subsection*{5. Potential weaknesses --- what breaks later, and under what condition}

\begin{itemize}\setlength{\itemsep}{0pt}
  \item \textbf{W1 [MK, HIGH]. Stale marks under refold surface a bare NAV.} An execution-time-early
  observation lands; projections refold the quantities but re-entered marks go stale (VM-7); NAV
  computes fresh quantities $\times$ stale marks and surfaces a bare number --- C-8.1's NAV type
  carries no staleness flag. Ties C4.
  \item \textbf{W2 [FL, HIGH; with TA]. New invariants have no firing witness, and no liveness.}
  Gate 1/2, VM-6, VM-7 map to no property test (CLAUDE.md\,\S3: ``zero firings is a defect'').
  \emph{[FL]} Gate 2 passes \emph{vacuously} where thin history never generates its precondition;
  a loose VM-6 bound marks every move ``explained.'' \emph{[TA]} detect-and-flag has no liveness
  --- a bulk retraction stale-flags much of the book (MD-8 propagates forward) with no deadline
  or owner.
  \item \textbf{W3 [TA, HIGH]. Gate 2 blinds risk where it is most needed.} Thin-history names
  (IPO, post-restructuring) yield permanent ``undecidable'' --- no risk report for the riskiest
  names; under regime shift (2008/COVID/dividend cut) unprecedented states are rejected as
  ``unrealistic,'' suppressing the tail. \emph{Divergence preserved:} TA reads thin history as
  runtime \emph{refusal}, FL (W2) as the test's \emph{vacuous pass} --- the refuse-behaviour must
  be \emph{shown to fire}.
  \item \textbf{W4 [TA, HIGH; with FL]. $\sigma_{\mathrm{prod}}$ poles force structural
  chain-breaks the alarm cannot distinguish.} At spot near $\Gamma_{\mathrm{adj}}=0$ an ordinary
  vanilla reconciles at no real volatility, residual forced $O(1)$, VM-7 calls the chain
  ``broken'' through no data fault. \emph{Divergence preserved:} FL reads PE-6's pole treatment
  as an honest diagnostic; TA reads the clash with VM-7's ``forbidden'' as alarm fatigue ---
  reconciled by exempting degenerate-carry poles from the broken-chain class.
  \item \textbf{W5 [TA, HIGH]. Write amplification meets the single writer and $O(n)$ replay.}
  VM-1 re-enters every model-priced mark; VM-10/MD-11 re-enter every mark of every scenario and
  MC path to be replayable. A real book funnels the whole valuation/risk firehose through one
  writer, and time-travel folds a prefix growing at pricing cadence, not event cadence. No
  checkpoint is fixed.
  \item \textbf{W6 [FL, MED]. Two correction species collide at the TA-union seam.} A vendor
  ``correction'' that also restates the timestamp is both an order-correction (reorders the fold,
  C-2.7) and a value-correction (MD-10); never distinguished --- undefined behaviour.
  \item \textbf{W7 [MK, MED]. Idempotence is conditional on an unnamed assumption.} C-11.5 lists
  it with no caveat; MD-1 admits two distinct prints identical under every field collide and one
  is silently dropped. Name the identifier-grain trust assumption beside the invariant.
\end{itemize}
