% Chapter 14 — The Invariant Catalogue (NOETHER). Budget 6 pp. Discharges §11 and §12.
\section{The Invariant Catalogue}\label{ch:invariants}

This chapter discharges constitution clauses C-11.1--C-11.5 (the structural invariants) and C-12.1--C-12.3 (time travel, replay, and idempotence).

The chapters before this one \emph{use} invariants; this chapter states each one
once, as an executable property, and discloses the inner ground for it --- the
reason it holds, not merely the fact that it does. The catalogue divides in two.
A \textbf{law} holds by construction: no check runs, because the property is a
theorem about the fold, and a state violating it is unrepresentable. A
\textbf{guard} is enforced at the door: the Transaction Executor computes it on
every admission, and a violation is refused, never written. Each property below
is stated in the form a test consumes; the executable-check regime of Chapter~15
(Testability and the Executable-Check Regime) exercises all of them over
generated products, events, and histories, together with the product-graph
consistency invariant of Chapter~\ref{ch:objects} --- restated there, not here.

\subsection{Conservation per (unit, coordinate, agreement)}

The central law is a conservation law, and it is stated at the finest grain the coordinate
vector has: per (unit, coordinate, agreement), not merely per unit or per coordinate. For a
wallet $w$, a unit $u$, a coordinate $c\in\{\text{owned, lent, posted}\}$, and an
agreement~$G$ governing the non-owned coordinates, write $\mathrm{bal}_{c,G}(w,u)$ for the
balance obtained by folding the log (Section~\ref{sec:coordinate-vector};
Chapter~\ref{ch:picture}). Owned is unqualified, so its agreement is null, and there the
finer grain and the per-coordinate grain coincide; on the marker coordinates $G$ is the
reference every marker move carries (Chapter~\ref{ch:collateral}).

\begin{theorem}[Conservation]\label{thm:conservation}
For every unit $u$, every coordinate $c$, every agreement $G$, and every position in the
log,
\[
\sum_{w}\ \mathrm{bal}_{c,G}(w,u)\ =\ 0,
\]
the sum running over all wallets --- virtual counterparty and issuer wallets
included, because the system is closed (Chapter~\ref{ch:objects}).
\end{theorem}

Issuance is the first paired-leg owned move out of the issuing wallet; retirement is the
log position after which every wallet's vector for $u$ is the zero vector, every $(c,G)$
coordinate zero. Two coarser laws follow as corollaries, never the primitive. Summing over
agreements gives conservation per (unit, coordinate), $\sum_w \mathrm{bal}_c(w,u)=0$;
summing further over coordinates gives the per-unit law $\sum_w\sum_c \mathrm{bal}_c(w,u)=0$.
The finest grain is primitive for a reason the encumbrance report makes concrete: a wallet
posting 100 under $G_1$ and receiving 100 under $G_2$ has
$\mathrm{bal}_{\text{posted},G_1}=+100$ and $\mathrm{bal}_{\text{posted},G_2}=-100$, two
separately conserved coordinates whose aggregate $\mathrm{bal}_{\text{posted}}$ is $0$.
Summation is many-to-one: the fully encumbered state and a wholly unencumbered one share
that image, so a law stated only on the image cannot tell them apart, and the encumbrance
projection built on the image would read zero for a fully encumbered wallet
(Section~\ref{sec:coordinate-vector}; Chapter~\ref{ch:reporting}). The finer law,
and the finer state it certifies, keeps them distinct --- that is the exact sense in which
it is strictly stronger. The per-coordinate grain in turn forbids a cross-coordinate
rebooking that fabricates custody on the posted plane while conserving the gross sum; each
grain forbids one class of fabrication the coarser one admits. This theorem is the
discharge of the constitution's third admission test for the coordinate schema ---
conservation at the finest declared grain (C-4.10) --- for the (coordinate, agreement)
schema Section~\ref{sec:coordinate-vector} declares; the earlier PARK-3 conflict is
closed by the owner's ruling in v1.2 (Chapter~\ref{ch:scope}).

\begin{proof}
Induction on the log. \emph{Base:} the empty log gives every $(u,c,G)$ coordinate a sum of
zero. \emph{Step:} the only operation that writes a balance is an atomic move, which debits
one wallet and credits another in the same coordinate of the same unit under the same
agreement by the same positive quantity $q$ --- a marker move carries exactly one agreement
reference and an owned move carries none, writing the null-agreement coordinate
(Chapter~\ref{ch:collateral}) --- so it adds $(-q)+(+q)=0$ to that one $(u,c,G)$ coordinate's
sum and nothing to any other; the other three kinds of change a transaction bundles ---
unit-state, position-state, terms --- touch no balance. The sum on every $(u,c,G)$
coordinate is therefore invariant across every admitted transaction, and remains $0$.
\end{proof}

The per-(unit, coordinate) law is the image of the theorem under summation over agreements:
for each $(u,c)$,
\[
\sum_{w}\ \mathrm{bal}_c(w,u)\ =\ \sum_{w}\sum_{G}\ \mathrm{bal}_{c,G}(w,u)\ =\ \sum_{G}\Big(\sum_{w}\ \mathrm{bal}_{c,G}(w,u)\Big)\ =\ \sum_{G} 0\ =\ 0,
\]
the middle step exchanging two finite sums. Summation carries the finer conserved
quantities to the coarser one and so preserves conservation; but it is many-to-one, so the
coarser law is strictly weaker, and the finer law is the primitive.

\begin{lstlisting}
conserved :: Log -> Unit -> Coord -> Agreement -> Bool
conserved l u c g = sum [ balance c g w u (fold l) | w <- wallets l ] == 0
-- a law, not a guard: the fold cannot reach a prefix where this is False.
-- owned is the null-agreement coordinate; markers carry their agreement (Ch. 9).
\end{lstlisting}

The word \emph{conservation} is exact: the vanishing sum is the quantity a
symmetry of the log preserves. That symmetry is stronger than conservation alone. A
transaction's \emph{footprint} is everything its fold step reads or writes across the
four kinds of change it may bundle: the (wallet, unit, coordinate, agreement) balance
cells its moves touch, and the UnitStatus, PositionState, and ProductTerms targets its
unit-state, position-state, and terms changes touch. Two adjacent admitted transactions
whose footprints are disjoint do not merely leave every coordinate sum unchanged when
swapped --- they leave the whole state bit-identical.

\begin{proposition}[Independent commutation]\label{prop:independent-commute}
Let $t_1$ and $t_2$ be adjacent admitted transactions with disjoint footprints ---
disjoint balance cells, and disjoint UnitStatus, PositionState, and ProductTerms
targets. Then applying $t_1$ then $t_2$ and applying $t_2$ then $t_1$ yield
bit-identical resulting state. Transactions that share any footprint element may not be
presumed to commute: buy shares before a dividend's record date and the holder is
entitled; meet the record date before buying and the holder is not, so the two orders
reach different state (Chapter~\ref{ch:picture}).
\end{proposition}

The proof is the disjointness. The fold step is a function of exactly the state cells it
reads and writes --- its footprint --- and of nothing else. On disjoint footprints $t_1$
writes only cells $t_2$ neither reads nor writes and conversely, so each step sees the
same inputs in either order and writes the same outputs; the two compositions agree cell
by cell. This discharges C-2.7's obligation: the constitution refuses to reorder anything
unless harmlessness is proved, and disjoint footprints are that proof --- the only case in
which a swap is harmless. The shared-footprint case is not an exception to be handled but
a reorder that stays forbidden, and the record-date pair is its witness. Conservation is
the corollary read on the balance plane alone: independent moves share no balance cell, so
every coordinate sum survives their swap, each being a sum of paired-leg contributions
that commute. Permutation invariance and the vanishing coordinate sum are one fact seen
twice; the first is itself a checkable (metamorphic) property.

\medskip\noindent\textbf{Episode --- the future (T1): variation margin is
conservation, not a check.}
\emph{Trigger.} FUT-IDX (multiplier $10$, settled-to-market) prints official
settlements $1{,}000$, $990$, $1{,}005$ on its three days; W-ALPHA is long one
contract, bought at $995$ on day~0, against the central counterparty wallet
W-CCP. \emph{Contract.} Each print fires the future's contract, which proposes one
owned(USD) move between the two wallets (Chapter~\ref{ch:collateral}).
\emph{Transactions.} Three, one per print:
\begin{enumerate}[label=(\arabic*),nosep,leftmargin=2.2em]
\item day~1, $995\!\to\!1{,}000$: move owned(USD) \USD{50}, W-CCP $\to$ W-ALPHA;
\item day~2, $1{,}000\!\to\!990$: move owned(USD) \USD{100}, W-ALPHA $\to$ W-CCP;
\item day~3, $990\!\to\!1{,}005$ (final): move owned(USD) \USD{150}, W-CCP $\to$
  W-ALPHA; the position unwinds.
\end{enumerate}
\emph{The door.} Each move is paired-leg owned(USD): whatever W-ALPHA gains W-CCP
loses, so $\mathrm{bal}_{\text{owned}}(\text{W-ALPHA},\text{USD}) +
\mathrm{bal}_{\text{owned}}(\text{W-CCP},\text{USD})$ contributed by these moves is
zero at every print. The cumulative variation margin to W-ALPHA is
$50-100+150 = 100 = (1{,}005-995)\times 10$, exactly the position's profit and
loss (Chapter~\ref{ch:valuation}).
\emph{Design point.} \emph{Variation margin is zero-sum because it is
conservation per (unit, coordinate, agreement) --- here the owned coordinate, whose
agreement is null, so the finer grain and the per-coordinate grain coincide; no
reconciliation enforces the zero-sum ---
the paired-leg move makes any other outcome unrepresentable --- and the cumulative
$100$ W-ALPHA gains is exactly what W-CCP pays, and equals the profit and loss.}

\begin{secondtelling}
Fix $(u,c,G)$. The map sending a log to the wallet-indexed integer vector of
balances is a homomorphism from the free monoid of moves under concatenation to
the group of such vectors under addition: the empty log maps to the zero vector,
concatenation to addition. Each move maps to a vector carrying $+q$ and $-q$ in
two entries and zero elsewhere --- an element of the subgroup whose entries sum to
zero --- and a homomorphism into that subgroup keeps every image inside it.
Conservation is the statement that the image lies in the kernel of the
sum-over-wallets map. Nothing below depends on this telling.
\end{secondtelling}

\subsection{The log's structural invariants}

Four further properties are laws of the log itself, guaranteed by the
single-writer discipline of the Transaction Executor (Chapter~\ref{ch:machines}).
Each is stated once, in the form a test consumes.

\begin{lstlisting}
atomic       :: Tx  -> Ledger -> Bool   -- apply writes the whole bundle or nothing
appendOnly   :: Log -> Log    -> Bool   -- a later log extends an earlier one; no edit
verifyChain  :: Log -> Bool             -- each entry carries its predecessor's hash
idempotent   :: Tx  -> Ledger -> Bool   -- re-applying a seen txid is a no-op
\end{lstlisting}

\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
\item \textbf{Atomicity.} A transaction applies in full or not at all: the
  Transaction Executor computes the whole four-kind bundle against the current
  ledger and either appends all of it or writes nothing. A refusal is a returned
  value, never a half-applied state --- no log position exists at which some but
  not all of a transaction's changes are present. The split shows it: no instant
  exists at which $20{,}000$ shares stand beside pre-split terms
  (Chapter~\ref{ch:objects}).
\item \textbf{Append-only and hash-chained.} An admitted transaction is never
  edited or removed; a mistake is repaired by a compensating transaction that
  names what it supersedes. Each entry carries the hash of its predecessor, so
  any alteration of history breaks the chain and is detectable
  (Chapter~\ref{ch:objects}).
\item \textbf{Determinism of every arrow.} Each of the three arrows --- watch,
  contract, apply --- is a pure function of its recorded inputs
  (Chapter~\ref{ch:picture}): the same record and clock emit the same events, the
  same event and ledger propose the same transaction, the same transaction and
  ledger reach the same ledger. Determinism is the ground on which replay and
  economic recomputation (Chapter~15) stand.
\item \textbf{Idempotence.} Every proposed transaction carries the cause-derived
  identifier $\mathit{txid}=H(\mathit{causeEventId},\mathit{contractId},\mathit{unitId},\mathit{seq})$
  of Chapter~\ref{ch:machines}; the door keys on the $\mathit{txid}$, so a retried or
  late-arriving proposal recomputes the identical identifier and is committed once,
  while the distinct legs of one cascade carry distinct identifiers and all commit.
  Applying a transaction whose identifier is already on the log changes nothing.
\end{itemize}

\subsection{Consistency of reference}

A quantity and a price name value only when both refer to the same state of the
instrument. All recorded data carries the basis frame it was observed in
(Chapter~\ref{ch:marketdata}); the guard refuses a read that would combine a
quantity with a price drawn from a different frame.

\begin{invariant}[Consistency of reference]\label{inv:consistency}
A quantity $Q$ and a price $P$ may be combined only when $Q$ and $P$ are read in
the same corporate-action frame. Across an event carrying a quantity change (a
move) and a value-frame change (an operator), total value is preserved:
$Q_{\text{pre}}\times P_{\text{pre}} = Q_{\text{post}}\times P_{\text{post}}$.
This binding is the \emph{invariance witness}, and a product taken across
mismatched frames is refused as a phantom.
\end{invariant}

\medskip\noindent\textbf{Episode --- the split (T4): the phantom the witness
refuses.}
\emph{Trigger.} ACME's $2$-for-$1$ split is effective; W-ALPHA's $10{,}000$
shares become $20{,}000$ by a paired-leg move, and the recorded price frame reads
$100.00$ as $50.00$ under the split's operator (Chapter~\ref{ch:marketdata}).
\emph{The witness.} Before: $10{,}000\times 100.00 = \USD{1{,}000{,}000}$. After,
in the post-split frame: $20{,}000\times 50.00 = \USD{1{,}000{,}000}$. The phantom
is $20{,}000\times 100.00 = \USD{2{,}000{,}000}$ --- the new count against the
stale frame. \emph{The door.} The $20{,}000$ is stamped in the post-split basis
and the $100.00$ in the pre-split basis; the frames do not match, so the product
is refused. The count changes by a move, the frame by an operator, and the two
are bound so their product is invariant.
\emph{Design point.} \emph{$20{,}000\times$ stale $100.00$ is a phantom
\USD{2{,}000{,}000} against a true \USD{1{,}000{,}000}; the witness refuses it
because the quantity and the price are read in different frames --- the same
witness Chapter~\ref{ch:marketdata} leans on for every corporate action.}

The same witness governs a distribution's cum/ex boundary, where no quantity
changes at all. A \emph{cum} price includes the coming distribution; an \emph{ex}
price excludes it; the two are one data kind read in states separated by the
recorded ex-date event, whose declared operator subtracts the distribution's
value. A cum price consumed as ex, or an ex price consumed as cum, is a phantom of
the split's kind --- the right number in the wrong frame --- and the same witness
refuses it, because the frame is fixed by the recorded ex-date state, not the wall
clock. Chapter~\ref{ch:marketdata} cross-references this invariant for both
readings.

The witness has a third reading, for a unit whose terms reference another and that carries
no move at all. A derivative --- OT-1's barrier on OMEGA, an option's strike, a single-name
contract's multiplier --- changes no holder's quantity when its underlying acts: the holder
holds one contract before and after. Here the witness is over the \emph{declared terms}:
across a corporate action on a referenced underlying, the payoff read on the post-event term
frame equals the payoff read on the pre-event frame, and the binding is the term-adjustment
operator that the corporate-action out-edge appends as a new terms version
(Chapters~\ref{ch:objects} and~\ref{ch:marketdata}). Under OMEGA's two-for-one split OT-1's
barrier is carried $80.00 \rightarrow 40.00$ and its \USD{1{,}000{,}000} payout stands; the
phantom is a knock read against the stale $80.00$, on which the first post-split close
$\approx 47.50$ fires a payout on a stock that never fell, while against the adjusted $40.00$
no knock fires. Quantity-times-price for a holding, and payoff-on-terms for a derivative, are
the two readings of one witness: value is invariant across the event, and a stale frame ---
whether a stale price or a stale term --- is refused.

\subsection{Writer discipline}

\begin{invariant}[One legal writer]\label{inv:writer}
Every non-balance fact has exactly one legal writer --- the smart contract of the
instrument the fact belongs to --- and the Transaction Executor refuses a
transaction that writes a fact outside its writer's authority
(Chapter~\ref{ch:homes}, constitution \S 11).
\end{invariant}

One writer per fact, one home per fact (Chapter~\ref{ch:homes}): a fact cannot be
written into disagreement with itself, because no second writer and no second home
exist to hold a rival copy. This is what keeps the three homes ---
ProductTerms, UnitStatus, PositionState --- free of the divergence a second store
would breed. Each is a \emph{materialised projection}: a cached, one-writer,
rebuildable read of the log, provably equal to the fold it caches, and \emph{not}
a second store of any derived quantity. No figure a valuation or a report
recomputes is materialised anywhere --- NAV, profit and loss, availability,
encumbrance are computed on demand (Chapters~\ref{ch:valuation}
and~\ref{ch:reporting}) --- so there is nothing derived to reconcile against the
record. Writer discipline over facts, and the refusal to store views, are jointly
the condition under which replay reproduces the homes exactly.

\subsection{Time travel and replay}

Section \S 12's guarantees are theorems of the fold --- but the fold runs along two axes,
not one, and the gap between the two axes and the constitution's single timestamp is a
parked conflict, named at the close of this subsection.

\begin{theorem}[Time travel]\label{thm:timetravel}
Every historical state of the ledger is the fold of the log along two axes: a
transaction-time prefix $p$ --- how far down the immutable log the reconstruction reads,
the axis of what was known --- and a valid-time horizon $h$ --- which recorded observations
are in force, the axis of what those observations say (Chapter~\ref{ch:picture}). Fixing
the pair $(p,h)$, the ledger is $\mathrm{fold\ step}$ over the committed transactions of
$p$ with $h$'s observations in force, and every view --- a balance, a NAV, a report --- is
a further fold of that same pair. Two facts hold.
\emph{(a)~As-known-at permanence:} the view at $(p,h)$ is never rewritten by a correction
admitted after $p$, because such a correction is a transaction appended strictly beyond $p$
and the prefix is fixed.
\emph{(b)~As-of exactness:} let $p'$ be the prefix that includes a back-dated observation
recorded at valid-time index $k$, and $h'$ the horizon putting it in force; the view at
$(p',h')$ recomputes bit-exactly from $p'$, because the fold reads only committed
transactions, never re-executes contracts, and is deterministic. The historical state is
thus independent of the machinery that produced it and reconstructible at any time, forever.
\end{theorem}

\begin{proof}
The ledger is the fold of the log (Chapter~\ref{ch:picture}); restricting to a prefix folds
the transactions that prefix selects, and determinism of apply (the determinism of every
arrow, above) makes the result a function of those transactions alone. \emph{(a):} fix
$(p,h)$. A later correction is an admitted transaction at a log position strictly beyond $p$
(append-only, above), so it is not among the transactions $p$ selects; a function of a fixed
argument does not change when a value outside that argument is added, so the view at $(p,h)$
is unchanged. \emph{(b):} the back-dated observation enters $p'$ as a recorded observation
carrying its valid-time index $k$. Folding $p'$'s observations in valid-time order re-derives
the order-dependent accrual tail $A_k, A_{k+1}, \dots$ from $k$ forward; every input to that
fold --- the recorded closes, the accrued integers (Chapter~\ref{ch:homes}) --- is on the
record, so the result is a function of $p'$ alone and reproduces bit-for-bit.
\end{proof}

The as-of fold of clause~(b) is not the log-order fold of the next theorem, and the
difference is the crux of the two axes. Replay re-folds the committed transactions of a
prefix; the as-of view additionally orders the recorded observations by their valid-time
index, so a back-dated observation --- appended late in transaction time, hence last in
log order --- is folded at valid-time index $k$, re-sequencing the tail $A_k,\dots$ from $k$
forward. The two folds traverse the same observations in two different orders, and each is
a deterministic function of the prefix; a party recomputes either one and obtains the same
bits. Confusing them --- folding a back-dated close in log order, or asserting the log-order
replay already carries the as-of answer --- is exactly the single-axis error parked
below.

\begin{theorem}[Replay determinism]\label{thm:replay}
Discarding the ledger and the three homes and re-folding a log prefix reproduces
them bit for bit. The apply arrow is deterministic (the determinism of every
arrow, above), each home is a materialised projection with one legal writer
(Invariant~\ref{inv:writer}), and a retried or late proposal is
idempotent under its cause-derived identifier; so replay adds nothing and drops
nothing.
\end{theorem}

\begin{proof}
The ledger is defined as the fold of the log (Chapter~\ref{ch:picture});
restricting to a prefix folds fewer transactions, and determinism of apply makes
the result a function of the prefix alone. Each home is the fold of its own kind
of change over the same prefix (Chapter~\ref{ch:homes}), hence reproduced by the
same replay. Idempotence guarantees a duplicated cause contributes once.
\end{proof}

These two theorems are the executable ground of the constitution's Full
Reproducibility and Time-Travel commitments. Chapter~15 (Testability and the
Executable-Check Regime) exercises them --- regenerating a mid-life firing from
the recorded fixings and comparing, and inserting a back-dated observation to
witness that the as-known-at view holds while the as-of view recomputes --- and
this catalogue does not restate the test.

\emph{The parked conflict.} Constitution \S 12 reconstructs a historical state ``by
replaying the immutable transaction log up to the desired timestamp'': one timestamp, one
axis. Theorem~\ref{thm:timetravel} needs two --- a transaction-time prefix and a valid-time
horizon --- because a back-dated observation makes the as-known-at and as-of views two
distinct, both-correct answers. This is a genuine conflict between the single-axis guarantee
and a system that has required both axes since the market-data boundary
(Chapter~\ref{ch:marketdata}, failure mode W2). It is parked, with the exact proposed amendment
text, as entry PARK-2 of the parking index (Chapter~\ref{ch:scope}); no chapter amends the
constitution, which is the owner's alone.

\subsection{The record is the log}\label{sec:record-is-log}

Replay reconstructs more than balances and the three homes. Because every observation
whose reproduction the framework guarantees enters the log as a moveless
observation-recording transaction (Chapters~\ref{ch:objects} and~\ref{ch:marketdata}),
re-folding a log prefix reconstructs every recorded observation as well --- the record
\emph{is} the log, with no second store to reconcile against. The single-writer
discipline (Invariant~\ref{inv:writer}) therefore covers every observation, not only
every move, and the ``two views of one log'' of Chapter~\ref{ch:picture} is exact: the
Record the \texttt{watch} arrow reads and the Ledger the \texttt{contract} and
\texttt{apply} arrows consume are the same folded log. Chapter~\ref{ch:testability}
exercises this as \texttt{prop\_replayReconstructsRecord}, which regenerates the whole
record --- observations included --- from the log alone and compares. The observation
stream is therefore no second half of the record with a door of its own: there is one door,
and it is the Transaction Executor.

\subsection{The coverage invariant and the two-guard contrast}

Collateral introduces one guard and one obligation that look alike and are
structurally opposite. This is their canonical home; Chapter~\ref{ch:collateral}
uses both.

\begin{invariant}[Possession coverage]\label{inv:coverage}
No wallet delivers what it neither owns nor holds. On every admission, per wallet
$w$ and unit $u$, net over agreements $G$,
\[
\sum_{G}\ \mathrm{bal}_{\mathrm{posted},G}(w,u)\ \le\ \max\bigl(\mathrm{bal}_{\mathrm{owned}}(w,u),\,0\bigr).
\]
The sign convention on the posted coordinate is \emph{received-negative}:
collateral posted \emph{out} enters with positive sign, collateral received
\emph{in} enters with negative sign (the received ray, poster $+Q$, taker $-Q$;
Chapter~\ref{ch:collateral}). The analogue holds with lent in place of posted.
\end{invariant}

The sum over signed balances lets received mass offset re-posted mass. Worked once,
for citation from Chapter~\ref{ch:collateral}: take a wallet with owned $0$. It
receives $100$ under one agreement, so its posted coordinate there is $-100$; it
re-posts $60$ under another, coordinate $+60$; the net is $60-100=-40\le\max(0,0)=0$,
within the bound, so a rehypothecation chain passes and is bounded by possession.
Had the same wallet re-posted $200$ instead, the net would be $200-100=+100>\max(0,0)=0$
--- refused at the door: a wallet may re-post no more than it received plus what it
owns. The $\max$ binds only the posting direction, so a negative-owned
wallet with nothing posted --- an issuer wallet that conservation itself creates,
a plain short --- is admissible, while a poster with owned $-5{,}000$ attempting to
post $5{,}000$ is refused at the door. Coverage is never lawfully false, because a
transaction is atomic and possession is instantaneous within it: that is what
makes it an invariant.

Coverage binds \emph{instructed} ownership, and the qualifier is load-bearing. Owned is
booked at instruction --- trade-date (Chapter~\ref{ch:collateral}) --- so a wallet may
post or deliver a security it has bought but not yet settled: that state is reachable,
admitted at the door because instructed ownership covers it. The invariant forbids only
\emph{recording} a state whose posted mass exceeds instructed ownership. A later
involuntary reduction of owned --- a failed or cancelled purchase that would drop owned
below the mass already posted --- is therefore not a silent breach but a transaction
\emph{refused at the door}: it cannot be recorded while it would violate coverage. The
unmet return is carried instead as a recorded, deadline-bearing obligation
(Chapter~\ref{ch:collateral}), and on default it escalates to the supervised write-off
path there. The fails cascade thus lives inside the model --- order-forced at the door,
then obligation, then supervised write-off --- and the invariant is never false in any
committed state.

On fungible cash coverage is a per-move check: a cash posting is admitted against owned
cash at the instant it is recorded, and because the same owned USD backs every
obligation drawn on it, aggregate cash adequacy across all obligations is not coverage
at all but the sufficiency obligation --- a deadline, not an invariant.

\begin{center}
\begin{tabular}{@{}p{0.20\textwidth}p{0.30\textwidth}p{0.16\textwidth}p{0.24\textwidth}@{}}
\toprule
Guard & Statement & Lawfully false? & Kind \\
\midrule
Possession coverage & $\sum_G \mathrm{bal}_{\mathrm{posted},G} \le \max(\mathrm{bal}_{\mathrm{owned}},0)$ & never (atomic, instantaneous) & invariant, checked at the door \\
Value sufficiency & collateral value $\ge$ exposure & intraday, routinely & obligation: deadline, discharge predicate, close-out as compensation \\
\bottomrule
\end{tabular}
\end{center}

\noindent Value sufficiency is the guard coverage is never to be confused with. A
market move, a knock, or a call in flight leaves collateral value below exposure
with no party in breach; sufficiency is therefore not an invariant but an
obligation --- a deadline, a discharge predicate (the delivery amount met), and
close-out as the declared compensation (Chapter~\ref{ch:collateral}, \S 14.1). A
book below its sufficiency line is not a broken state; it is an open,
deadline-bearing item on the record.

\medskip\noindent\textbf{Episode --- the one-touch (T2): coverage on the
discharge; value extinguished, mass intact.}
\emph{Setup.} OT-1 (Cast, \S\ref{sec:cast}) pledged to B under agreement G as in
\S\ref{sec:knockpledged} (collateral value to B \USD{200{,}000} against exposure
\USD{150{,}000}). OMEGA prints its barrier close, $79.00 \le 80.00$.
\emph{The discharge transaction.} The payout obligation discharges under its
trapping predicate in one transaction, two moves: owned(USD) \USD{1{,}000{,}000}
W-BANK $\to$ W-ALPHA, and posted(USD) \USD{150{,}000} W-ALPHA $\to$ B under G;
\USD{850{,}000} stays free.
\emph{The door.} Coverage holds within the transaction: W-ALPHA's owned cash
$1{,}000{,}000 \ge$ posted $150{,}000$, so the posting is admitted --- a posting
of collateral W-ALPHA did not possess would be refused.
\emph{Value versus mass.} The knock extinguishes value, never mass: pricing the
triggered OT-1 at $0$ collapses its value, but its coordinate balance is unchanged
--- coordinates carry mass, prices carry value. Retirement is only from the zero
vector: the marker return clears the posted plane of OT-1 first --- coverage
forces the order, refusing to extinguish owned while posted mass remains --- then
owned(OT-1) moves W-ALPHA $\to$ W-BANK, taking every wallet's vector for OT-1 to
zero, and OT-1 retires.
\emph{Design point.} \emph{Coverage is checked at the door on the discharge
(owned $1{,}000{,}000 \ge$ posted $150{,}000$) and again as the order of close-out;
the lifecycle event extinguishes OT-1's value but not its mass; and between knock
and call the intraday shortfall is carried by the sufficiency obligation, not by a
broken invariant.}

\subsection{The post-and-return metamorphic property}

The last property makes ``only the owned coordinate carries economic value''
(Chapter~\ref{ch:objects}, constitution \S 4) machine-checkable.

\begin{proposition}[Post-and-return]\label{prop:postreturn}
A pledge posting writes the posted plane and never owned, so $V=\sum
\mathrm{owned}\cdot P$ is unchanged by any post-and-return round trip. Inserting a
post-and-return pair within any watch-free interval of the referenced agreement
leaves every later valuation identical.
\end{proposition}

The side condition is real and named: an insertion spanning the agreement's
valuation watch lawfully changes later firings and cash, so the property owns the
move algebra, not the watched case. The watched case is the temporal residual
carried by the safety and liveness model of Chapter~15 (Testability and the
Executable-Check Regime). As a test the proposition is metamorphic: insert the
pair into a generated history and assert that every subsequent NAV is unchanged to
the minor unit.

\subsection{What the catalogue buys}

Every property above is stated once and executable, none defended in prose alone.
Together with the product-graph consistency invariant of
Chapter~\ref{ch:objects}, they are the surface Chapter~15 tests over one
generator universe. Their joint effect is the constitution's claim in \S 11: the
classes of inconsistent state that internal reconciliation exists to detect ---
unbalanced books, a fact written twice into disagreement, a quantity valued in the
wrong frame, collateral posted in excess of the poster's instructed ownership --- are
not caught here; they are unreachable. What cannot be represented cannot break. The
residual economic exposures that remain --- a bought-not-yet-settled posting whose
purchase later fails, an intraday shortfall of value against exposure --- are reachable
and real, but the model carries them as recorded, deadline-bearing obligations
(Chapters~\ref{ch:collateral} and~\ref{ch:settlement}), never as silent breaches of an
invariant.
