\section{Where Unit State Lives: The Three-Home State Model}
\label{sec:states}

Three kinds of fact describe a unit, and each is kept in its own home. The first is what
the contract is: its fixed terms, written once at registration and thereafter only added
to, never rewritten. The second is where the contract stands in its life: its lifecycle
stage and last settlement price, a single shared value read the same way by everyone who
holds it. The third is what a given holder holds: that holder's own quantity and cost,
which differ from one holder to the next. These three facts are the three homes named
below.

Unit state lives in three homes, one map each: immutable \emph{ProductTerms}$[u]$
keyed by unit; shared \emph{UnitStatus}$[u]$ keyed by unit; and per-position
\emph{PositionState}$[w,u]$ keyed by (wallet, unit). There is no map keyed by wallet
alone: every per-wallet economic fact is keyed by a mandate or strategy unit, so it
lives in \emph{PositionState}$[w,u_{\mathrm{MA}}]$. Each economic fact has exactly one
home and one writer, so internal reconciliation failure cannot arise. Conservation and
deterministic replay then hold by construction: the single-writer discipline is sealed,
and replay is pure and total. This placement is the minimum basis of
the problem (\S\ref{sec:states-three}) and unique under any correctness gate
(\S\ref{sec:states-pareto}).

A \emph{unit} is anything held and conserved (\S\ref{sec:ledger}): an instrument, and
also a mandate or a strategy contract (\S\ref{sec:states-three}). The framework
primitives are those of \S\ref{sec:ledger}: wallet, unit, holding $h(w,u)$, move,
transaction, and conservation $\sum_{w}h(w,u)=0$.

\subsection{The placement rule and the 2\texorpdfstring{$\times$}{x}2}
\label{sec:states-rule}

One rule governs every placement: each economic fact has exactly one home and one
writer. A fact violates it in exactly two ways. \emph{Keyed wrong}: one home collapses
two facts, or one fact duplicates across homes that drift. \emph{Authored wrong}: two
authorities keep one fact, and their records drift apart. Two questions forestall the
two failures. Because there are only two failures, there is no third question.

\emph{Does the fact depend on the holder, or only on the unit?} The scope is one unit's
state, so a unit and a holder of it are the only terms in view. Every economic fact ---
one entering conservation, valuation, or profit and loss --- is then a (holder, unit)
fact or a unit fact. What stays keyed by the wallet alone (KYC, permissions, an audit
cursor) enters none of these three reckonings. It is identity, not economic state. It
lives in a \emph{WalletRegistry} that carries no quantity and joins no conservation law.

\emph{Is the fact's record owned by an outside authority or by the ledger?} The criterion
is who owns the record's history, not who sources the number. An externally authored fact
is one an outside authority --- the exchange, the contract, the reference-data provider
--- owns and restates; the ledger preserves it version by version. A ledger-authored
fact is one the ledger's own events produce and overwrite. One provider can supply both:
the benchmark level, which the ledger records and overwrites, is ledger-authored; the
benchmark identity, which the provider restates, is the provider's own versioned record
and externally authored. The two sort apart though both come from the same provider.

The two questions give a $2\times2$; three of its four cells are occupied, and each
occupied cell is one home.

\begin{center}
\begin{tabular}{@{}r p{0.36\textwidth} p{0.36\textwidth}@{}}
\toprule
 & \textbf{ledger-authored} & \textbf{externally authored} \\
\midrule
\textbf{per unit} & \textbf{Status}: lifecycle stage, last settlement price, current
weights, benchmark level & \textbf{Terms}: multiplier, currency, expiry, strike, ISIN,
fee schedule, mandate text, benchmark identity \\
\addlinespace
\textbf{per (holder, unit)} & \textbf{Position}: held quantity, accumulated cost,
high-water mark, entry NAV & \textit{--- empty ---} \\
\bottomrule
\end{tabular}
\end{center}

Terms and status share the unit key but cannot be one value, since their authorities of
record differ (\S\ref{sec:states-three}). They ride together on the unit key as a
co-present pair: two homes whose mutation disciplines differ, each kept in its own map
(\S\ref{sec:states-construction}).

\begin{figure}[tb]
\centering
\begin{tikzpicture}[
    >={Stealth[length=2.4mm]},
    home/.style={draw, rounded corners, align=center, font=\small,
                 minimum width=24mm, minimum height=11mm},
    empty/.style={draw, dashed, align=center, font=\footnotesize,
                  text=black, minimum width=24mm, minimum height=11mm},
    aux/.style={draw, align=center, font=\footnotesize,
                minimum width=20mm, minimum height=9mm},
    val/.style={draw, thick, align=center, font=\small,
                minimum width=52mm, minimum height=10mm},
    ext/.style={draw, dashed, align=center, font=\footnotesize,
                minimum width=40mm, minimum height=9mm},
    derive/.style={->, thick},
    reads/.style={->},
  ]
  % --- the 2x2 placement; three cells occupied, one empty ---
  \node[font=\footnotesize\itshape] at (0,3.7)   {ledger-authored};
  \node[font=\footnotesize\itshape] at (4.6,3.7) {externally authored};
  \node[font=\footnotesize\itshape, align=right] at (-2.9,1.8) {per\\unit};
  \node[font=\footnotesize\itshape, align=right] at (-2.9,0.0) {per\\(holder, unit)};

  \node[home]  (us) at (0,1.8)   {UnitStatus\\[1pt]\footnotesize shared projection};
  \node[home]  (pt) at (4.6,1.8) {ProductTerms\\[1pt]\footnotesize immutable};
  \node[home]  (ps) at (0,0.0)   {PositionState\\[1pt]\footnotesize per (wallet, unit)};
  \node[empty] (mt) at (4.6,0.0) {--- none ---};

  % --- the load-bearing arrow: UnitStatus is derived from the event log ---
  \node[aux] (log) at (-3.2,3.0) {Event log\\[1pt]\footnotesize immutable};
  \draw[derive] (log.east) -- node[above, font=\scriptsize, pos=0.5]{derive} (us.north);

  % --- valuation reads all three homes plus the external price vector ---
  \node[val] (v) at (2.2,-2.0) {Valuation:\ \ $V_t=\sum_{u\in\mathcal{U}} w_t(u)\,P_t(u)$};
  \draw[reads] (us.west) -- (-1.7,1.8) -- (-1.7,-2.0) -- (v.west);
  \draw[reads] (pt.east) -- (6.4,1.8) -- (6.4,-2.0) -- (v.east);
  \draw[reads] (ps.south) -- (v.north);
  \node[ext] (px) at (2.2,-3.5) {price vector $P_t$ \footnotesize(external input)};
  \draw[reads] (px.north) -- (v.south);
  \node[font=\scriptsize, align=center] at (-2.1,-1.0) {reads};
\end{tikzpicture}
\caption{The three state homes and their dataflow. Each economic fact has one home, fixed by
the placement above: \texttt{UnitStatus} (per unit, ledger-authored), \texttt{ProductTerms}
(per unit, externally authored), and \texttt{PositionState} (per (wallet, unit)); the fourth
cell---an externally authored per-position fact---is empty by construction (\S\ref{sec:states}).
\texttt{UnitStatus} is not an authoritative store but a projection derived from the immutable
event log, shown by the incoming arrow. Valuation reads all three homes together with an
externally supplied price vector $P_t$.}
\label{fig:threehome}
\end{figure}


\subsection{Why three homes, and why the fourth cell is empty}
\label{sec:states-three}

Each occupied cell exists for one reason, and the empty cell is empty for one. Exactly
three maps are forced, no fewer and no more.

\emph{Position is keyed by (holder, unit), because two holders of one contract hold
different state.} A buyer and a seller hold the same future. The buyer's quantity is
$+1000$, the seller's is $-1000$, and their accumulated costs differ. Substituting one
holder's position for the other's changes the value, so any unit-only keying identifies
the two and loses the difference. The holder is in the key.

\emph{Status is keyed by the unit alone, because one value is read identically by every
holder.} A future settles at one price on the day; an index stands at one level. Stored
per holder, the number is copied across thousands of rows free to drift --- a
reconciliation break by construction. Keyed by the unit, it is one value with one writer.

\emph{Terms are a home distinct from status, because the two have different authorities of
record.} Terms are externally authored; status is ledger-authored. Co-mingling the
authority's record with the ledger's own is the single-source-of-truth violation the
system exists to prevent. The mutation disciplines differ too. Terms are append-only and
versioned for reconstruction. The status cache is overwritten on every settlement, while
the settling event is appended immutably to the log (\S\ref{sec:states-unitstatus}).
Merging the two would put two mutation disciplines under one name.

\emph{The fourth cell is empty, because no authority issues a fact about one holder's
position.} A position is the ledger's own from origin: the held quantity is folded from
the move stream, and the high-water mark and entry NAV are written by ledger events. A
custodian statement or prime-broker position report is a reconciliation input the ledger
checks against, not a record it adopts as an authority's own. So the (holder, unit) key
carries only ledger-authored facts, and the externally authored cell is empty by
construction.

\paragraph{The $W$-sector collapses (C12).}
The managed account, whose state is the client's rather than the instrument's, appears
to demand a wallet-keyed economic-state map --- call it the \emph{$W$-sector},
$\mathrm{Map}[\mathrm{WalletId},\mathrm{WalletState}]$. No such map exists, because the
mandate is a unit. The manager issues a mandate unit $u_{\mathrm{MA}}$ to the
client (\S\ref{sec:managed}), and it conserves like any issued unit:
\[
h(w_{\mathrm{manager}},u_{\mathrm{MA}})=-1,\qquad
h(w_{\mathrm{client}},u_{\mathrm{MA}})=+1,\qquad
\sum_{w}h(w,u_{\mathrm{MA}})=0 .
\]
Every field that looked per-wallet is then a fact about the position
$(w_{\mathrm{client}},u_{\mathrm{MA}})$ and lives in \emph{PositionState}:
high-water-mark value, accrued management and performance fee, mandate breach flags,
benchmark NAV at this wallet's inception. Two mandates on one wallet are two rows, each with its own
high-water mark and fees; a wallet-keyed scalar would have merged them. This is why
\hyperref[cond:C12]{C12} keys on the mandate unit, not the wallet.

A relationship spanning
several instruments that does \emph{not} reduce to a single (holder, unit) row would be a
(holder, several-units) fact. Such a fact would require a fourth home under a new key
shape; the three-map count excludes it, and it lies outside this scope.

\subsection{The UnitStatus discipline, stated once}
\label{sec:states-unitstatus}

\begin{mentalmodel}
Picture a bound ledger-book written only in pen: each event is a line entered in ink,
and no line is ever erased. On the cover sits a running total in pencil, rubbed out and
rewritten after every new line. Rubbed out entirely, the pencil costs nothing: re-adding
the ink lines reproduces the exact total that stood at any chosen line. Unlike a
chequebook, where only the balance need be kept, here the pencilled total is the
disposable part and the ink lines never are --- the total answers at a glance but
certifies nothing on its own; the ink is the record.
\end{mentalmodel}

UnitStatus is a read cache. It holds one value per unit, shared: every holder reads it
identically, and that value changes over time. The source of truth is the immutable event
log, not UnitStatus (justified by the idempotent replay of \S\ref{sec:lifecycle}): every
change to the stored value is caused by a logged event and overwrites it in place, and
replaying the events up to any point rebuilds the exact value that held then. The cache
can therefore be discarded and rebuilt from the log without loss. There is no other way to
change the value, which exists from registration onward.

Equivalently, the stored value always equals what replaying the unit's events produces,
and no part of the system writes it except through a logged event. This is the
catamorphism property: the fold of the event stream always rebuilds the exact cached
value. In the reference (\S\ref{sec:states-reference}) it holds by construction. Every
\emph{UnitStatus} field is written only by \texttt{applyStatus}, folded over the unit's
status writes inside the sole transition \texttt{applyTx}; \texttt{register} is
\texttt{applyTx} of a move-less introduction. The \texttt{Ledger} constructor is
unexported, so no out-of-band \texttt{setStatus} exists. Externally sourced inputs --- a
settlement price, the current benchmark level --- enter only as a logged observation
event, so the fold stays pure at the boundary. Every later reference to cached unit state
in this specification is read in this sense.

\subsection{The construction}
\label{sec:states-construction}

The homes are built from the smallest pieces, each forced by the one before. The
conditions of the register (\S\ref{sec:states-conditions}) are introduced where the
construction forces them: C1--C12 here, C13 and C14 at the data plane's basis edge
(\S\ref{sec:state-basis}). Quantities are exact integer minor units, never floating
point, so determinism and conservation are arithmetic facts.

\paragraph{Scalars and keys.}
The smallest pieces are the number types and the keys that name them. A balance is a
quantity with a zero, associative addition, and a negation --- an abelian group. The
negation forces the two legs of a transfer to cancel. The zero makes the sum over no
wallets conserved with no special case, which discharges the vacuous base of conservation
(\hyperref[cond:C9]{C9}). A price is a number but not a quantity: prices are never
summed, so a price is a separate type with neither identity nor inverse. A balance is
named by two keys, wallet and unit, the holder in the key.

\begin{lstlisting}[language=Haskell]
{-# LANGUAGE GADTs, DataKinds, KindSignatures, StandaloneDeriving #-}
import           Control.Monad      (foldM)
import           Data.List          (foldl')
import           Data.List.NonEmpty (NonEmpty (..))
import qualified Data.List.NonEmpty as NE
import           Data.Map.Strict    (Map)
import qualified Data.Map.Strict    as Map
import           Data.Set           (Set)
import qualified Data.Set           as Set

newtype Qty = Qty Integer deriving (Eq, Ord)        -- exact minor units, never Float
instance Show Qty      where show (Qty n) = show n
instance Semigroup Qty where Qty a <> Qty b = Qty (a + b)
instance Monoid    Qty where mempty = Qty 0          -- the zero discharges the vacuous base (C9)

qneg :: Qty -> Qty                                   -- group inverse: a transfer's legs cancel
qneg (Qty n) = Qty (negate n)
qmax :: Qty -> Qty -> Qty
qmax (Qty a) (Qty b) = Qty (max a b)

newtype WalletId = WalletId String deriving (Eq, Ord, Show)
newtype UnitId   = UnitId   String deriving (Eq, Ord, Show)

newtype Price = Price Integer deriving (Eq, Ord, Show)   -- minor units per unit; NO Monoid

-- The corporate-action basis coordinate (sec:state-basis). BoundaryId is exported
-- ABSTRACT: the address of a Boundary -- a corporate-action boundary -- is assigned
-- by content at the attested notice gateway, never minted by a caller. A BasisId is
-- a content address, never an ordinal, so a retro-effective boundary renumbers
-- no stored value.
newtype BoundaryId = BoundaryId String deriving (Eq, Ord, Show)
data    BasisId    = Origin UnitId | Boundary BoundaryId deriving (Eq, Ord, Show)
\end{lstlisting}

\paragraph{ProductTerms --- the authority's record, grown only by appending (C6, C7).}
The picture is a stack of versions that only grows: the first is laid down at
registration, later ones are laid on top, and nothing beneath is ever touched.
``One or more versions, append-only'' is a non-empty list, so a registered unit always has
a current version and ``registered but versionless'' is not representable. The type is
abstract: its constructor is unexported, so no importer can lay down a fresh one-version
value and discard history. The only growth operations are \texttt{register}, which writes
version one, and \texttt{appendVersion}, which appends. No setter exists, so the
append-only discipline holds by absence, not by convention (\hyperref[cond:C6]{C6}). The
first version is written at registration, so terms are total on the registered units
(\hyperref[cond:C7]{C7}).

\begin{lstlisting}[language=Haskell]
data TermsVersion = TermsVersion
  { tvLabel :: String, tvFields :: Map String String }  -- opaque payload: multiplier, ISIN, ...
  deriving (Eq, Show)

newtype ProductTerms = ProductTerms (NonEmpty TermsVersion) deriving (Show)
  -- exported ABSTRACT: no data constructor, no setter -> append-only holds by absence (C6)

currentTerms :: ProductTerms -> TermsVersion           -- total: NonEmpty always has a last
currentTerms (ProductTerms vs) = NE.last vs
allVersions  :: ProductTerms -> NonEmpty TermsVersion
allVersions  (ProductTerms vs) = vs
appendVersion :: TermsVersion -> ProductTerms -> ProductTerms   -- C6: append, never rewrite
appendVersion tv (ProductTerms vs) = ProductTerms (vs <> (tv :| []))

productTerms :: Ledger -> UnitId -> Maybe ProductTerms  -- Nothing = unregistered
productTerms l u = Map.lookup u (ledgerPT l)
\end{lstlisting}

\paragraph{UnitStatus --- the shared read cache (C5).}
The picture is one shared record per unit, overwritten in place, with the log keeping the
event behind every overwrite. Status records the unit's lifecycle stage, its
corporate-action basis (state basis, \S\ref{sec:state-basis}), its last settle mark, and
any successor, under the discipline of \S\ref{sec:states-unitstatus}. The settle mark is
dimensionally a \texttt{Price}, paired with the basis it was struck at. The lifecycle is
a closed sum,
\texttt{Registered}$\to$\texttt{Active}$\to$\texttt{Expired}$\to$\texttt{Closed}.
\texttt{Registered} is the at-registration stage: a unit recorded in and known to the
ledger before any settlement. An OTC option, a bespoke mandate, or an internal strategy
reaches it exactly as an exchange instrument does. Every \emph{UnitStatus} field changes
only through a \texttt{StatusWrite}, the closed writer set whose four constructors each
name one field. Each write is applied by the single total writer \texttt{applyStatus},
the projection step of \S\ref{sec:states-unitstatus}. No other code overwrites a status
field, so each write is last-write-wins and idempotent (P6). Product-declared defaults
are written at Unit Store registration (\S\ref{sec:unit-store}), so status is total on
the registered units (\hyperref[cond:C5]{C5}).

\begin{lstlisting}[language=Haskell]
data Lifecycle = Registered | Active | Expired | Closed deriving (Eq, Show)

data UnitStatus = UnitStatus
  { usLifecycle    :: Lifecycle
  , usBasis        :: BasisId                 -- corporate-action basis; tip-welded (sec:state-basis)
  , usLastSettle   :: Maybe (Price, BasisId)  -- a settle mark names its basis; None at default (C5)
  , usSupersededBy :: Maybe UnitId            -- set only by a Breaking amendment (C8)
  } deriving (Eq, Show)

defaultStatus :: UnitId -> UnitStatus -- C5: the per-unit default written at registration;
defaultStatus u = UnitStatus Registered (Origin u) Nothing Nothing
                                      --   the basis origin is forced by the unit itself

data StatusWrite                      -- the closed status-writer set; each names ONE field
  = SetLifecycle    Lifecycle         -- a lifecycle event
  | SetBasis        BasisId           -- a corporate-action boundary crossed (sec:state-basis)
  | SetLastSettle   (Price, BasisId)  -- a settle event (a logged observation)
  | SetSupersededBy UnitId            -- a Breaking amendment (C8)
  deriving (Eq, Show)

applyStatus :: StatusWrite -> UnitStatus -> UnitStatus   -- the ONLY writer of a status field;
applyStatus (SetLifecycle l)    s = s { usLifecycle    = l }      -- total; every equation an
applyStatus (SetBasis b)        s = s { usBasis        = b }      -- ABSOLUTE replacement write,
applyStatus (SetLastSettle m)   s = s { usLastSettle   = Just m } -- never an increment:
applyStatus (SetSupersededBy u) s = s { usSupersededBy = Just u } -- last-write-wins, idempotent (P6)

unitStatus :: Ledger -> UnitId -> Maybe UnitStatus       -- Nothing = unregistered
unitStatus l u = Map.lookup u (ledgerUS l)
\end{lstlisting}

\paragraph{PositionState --- the per-position home (C1).}
The picture is one row per (holder, unit): the holder's own quantity, and beside it the
holder-specific facts that ride with it. A position carries more than a balance. Its
primary field is the held quantity, which conserves. Beside it ride non-conserved
per-(holder, unit) facts --- a high-water mark, an entry NAV --- which no zero-sum law
binds. A first touch starts a position at
$0_P$, the all-zero state. Two orthogonal disciplines govern the home
(\hyperref[cond:C1]{C1}). \emph{(a)} The accessor returns $\mathrm{Option}$, so
$\mathrm{None}$ (never held) and $\mathrm{Some}(0_P)$ (held, now flat) stay distinct;
variation-margin settlement, wash-sale lookback, and record-date entitlement each act on
the difference. \emph{(b)} The carrier is monotone: a row, once created, is never
removed, and close-out leaves it flat. The two are independent: (a) constrains the return
type, (b) the domain. The monotone carrier keeps the key set stable across any checkpoint
cut, which is what makes replay a literal fold (P8). The high-water-mark and entry-NAV
writers are valuation events specified in the valuation satellite
(\S\ref{sec:valuation}); they are out of scope here and leave those fields at their
defaults throughout.

\begin{lstlisting}[language=Haskell]
data PositionState = PositionState
  { psAc       :: !Qty           -- conserved, additive        <- Settle / Trade
  , psBalance  :: !Qty           -- conserved, additive        <- Transfer
  , psHwm      :: !Qty           -- NOT conserved, monotone    <- FeeCrystallise (qmax)
  , psEntryNav :: !(Maybe Qty)   -- NOT conserved, write-once  <- Subscribe
  } deriving (Eq, Show)

zeroP :: PositionState           -- flat row on first touch; also the held-and-flat baseline
zeroP = PositionState mempty mempty mempty Nothing

position :: Ledger -> WalletId -> UnitId -> Maybe PositionState   -- C1 Option accessor:
position l w u = Map.lookup (w, u) (ledgerPS l)
  -- Nothing = never held;  Just zeroP = held once, now flat.  The two stay distinct.
  -- Monotone carrier (C1, storage half): applyTx only inserts/updates a row, and no
  -- row deleter is exported anywhere -- close-out leaves a flat row, never removes it.
\end{lstlisting}

\paragraph{The three homes, three maps --- the sealed Ledger.}
The picture is the three maps travelling as one sealed value, built and read only through
the operations that follow. Terms and status share the unit key but keep separate maps
--- \texttt{ledgerPT} append-only, \texttt{ledgerUS} overwrite-in-place --- because their
mutation disciplines differ. Their co-presence, ``$u$ registered in terms $\iff$ $u$
registered in status,'' is preserved by construction, not policed: \texttt{register} is
the sole introducer and writes both, and every later writer touches them only for an
already-registered unit. The position map is the balance map enriched. The
\texttt{Ledger} constructor and its field selectors are unexported: a \texttt{Ledger} is
built only by the writers below and read only through the accessors. Were a selector
exported, record update through it would install a non-conserving map without naming the
constructor, bypassing the single-writer discipline. Coherence is carried by the single
introducer; the seal keeps conservation true by construction. A fourth field,
\texttt{ledgerBounds}, is not a home. It caches the committed boundary projection read by
the basis welds --- the basis admission conditions \texttt{applyTx} checks
(\S\ref{sec:state-basis}). Like \texttt{ledgerUS}, it is rebuilt by replay and is never a
source of truth.

\begin{lstlisting}[language=Haskell]
data Ledger = Ledger
  { ledgerPT :: Map UnitId             ProductTerms   -- append-only terms, by unit
  , ledgerUS :: Map UnitId             UnitStatus     -- overwrite status,  by unit
  , ledgerPS :: Map (WalletId, UnitId) PositionState  -- positions,   by (holder, unit)
  , ledgerBounds :: Map UnitId (Set (Timestamp, Integer, BoundaryId))
      -- committed-boundary projection: each unit's (t_eff, prec, bid) triples, whose
      -- derived Ord IS the effective order. A DERIVABLE CACHE of the log, exactly
      -- like ledgerUS -- rebuilt by replay, never a source of truth.
  } deriving (Show)
  -- exported ABSTRACT: constructor and selectors unexported. A Ledger is built only by the
  -- writers below and read only through the accessors -> single-writer discipline; with no
  -- row deleter, the carrier is monotone (C1).

emptyLedger :: Ledger
emptyLedger = Ledger Map.empty Map.empty Map.empty Map.empty

data LedgerError = ReRegistration UnitId | UnitAlreadyExists UnitId | UnknownUnit UnitId
                 | DanglingSupersede UnitId    -- SetSupersededBy target is not a registered unit
                 | BasisNotTip UnitId          -- the three basis welds at the same single
                 | InvarianceViolation UnitId  --   door, specified in sec:state-basis and
                 | ForeignOrigin UnitId        --   implemented in reference/Ledger.hs
                 | ScheduleIncomplete UnitId   -- C14: the written version's schedule fails
                                               --   totality over (CAClass x field)
                 | CauseNotCommitted UnitId    -- txCause names no strictly-earlier
                                               --   committed boundary id
  deriving (Eq, Show)
\end{lstlisting}

\paragraph{The Transaction --- the one atomic event (C2, C3).}
The picture is one value carrying everything one event does to one unit; it lands whole
or not at all. One event records, for a single unit, the whole change across the three
maps, in two complementary parts. First, the conserved economic flow in \emph{edge} form:
a list of moves, each debiting one wallet and crediting another by the same magnitude.
The signed per-unit sum of any move list is therefore the group identity \emph{by
construction}: $\mathrm{qneg}\,q \mathrel{\diamond} q = \mathit{mempty}$ for each leg,
summed by the monoid laws. The empty list (a move-less registration or amendment) is
included; it is the vacuous base (\hyperref[cond:C9]{C9}). Second, the non-balance state
delta that rides alongside the moves: per-wallet bookkeeping (accumulated cost,
high-water mark, entry NAV, via the C11 \texttt{FieldWrite} table), the shared
\emph{UnitStatus} writes, and a \emph{ProductTerms} introduction or append.

Conservation is then a property of the type, not a checked side-condition. There is no
unconserved \texttt{Transaction} to reject, hence no \texttt{validate} gate
(\hyperref[cond:C2]{C2}). A transaction applies in full or not at all; a partially applied
transaction is not a representable state (\hyperref[cond:C3]{C3}). One structural field,
\texttt{txIntroduce}, is the sole asymmetry between registration and every other event:
\texttt{Just} introduces the unit, writing terms and status together; \texttt{Nothing}
operates on an existing one. Registration is thus the move-less instance of this one type,
not a separate operation.

An event touching several units --- a rebalance trading three futures, or a future
bought against cash --- is a transaction per unit applied as a single fold. Per-unit
conservation composes to event-level conservation. Induction over the stream lifts the
per-event identity to the global invariant P1.

\begin{lstlisting}[language=Haskell]
-- One value, two notations: txMoves is the per-edge view of the conserved flow; the
-- remaining fields are the non-balance bookkeeping that rides with it.
-- (Move and moveDelta are the sec:ledger primitives in their reference form: the
-- magnitude mQty is a PosQty -- positive by type, sole door mkPosQty, the parse
-- boundary of sec:futures -- read back through unPosQty. sec:ledger presents the
-- same algebra in illustrative Qty form.)
data Transaction = Transaction
  { txUnit      :: UnitId                           -- the unit this event concerns
  , txMoves     :: [Move]                           -- conserved flow as edges (by construction)
  , txRows      :: Map WalletId [SomeWrite]         -- non-balance per-wallet bookkeeping
  , txStatus    :: [StatusWrite]                    -- shared UnitStatus writes
  , txIntroduce :: Maybe (TermsVersion, UnitStatus) -- Just => introduce txUnit (registration)
  , txAppend    :: Maybe TermsVersion               -- Just => append a version (C6, Preserving)
  , txBoundary  :: Maybe BoundaryEvent              -- Just => declare a corporate-action boundary
                                                    --   on txUnit: id, t_eff, prec, declaration
                                                    --   (Definition and welds in sec:state-basis)
  , txCause     :: Maybe BoundaryId                 -- PROVENANCE ONLY: the parent notice's
                                                    --   boundary id on a follow-on or cascade;
                                                    --   Nothing on a unit's own boundary tx (its
                                                    --   identity lives in txBoundary -- one fact,
                                                    --   one home). Admission requires the referent
                                                    --   be committed strictly earlier (refusal
                                                    --   CauseNotCommitted), so self-reference is
                                                    --   unrepresentable; read only by the
                                                    --   recomputation witnesses (P27/P28) and by
                                                    --   audit -- NO code path branches on it.
  } deriving (Show)

-- Conservation witnesses. Each move contributes qneg q <> q = mempty, so the foldMap lands
-- at mempty BY THE MONOID LAWS -- not by a sum computed and compared at run time (C2, C9).
unitDelta :: UnitId -> Transaction -> Qty
unitDelta u tx =
  foldMap (\m -> if mUnit m == u then let n = unPosQty (mQty m) in qneg n <> n else mempty) (txMoves tx)
-- law:  unitDelta u tau == mempty   for every unit u and transaction tau

netDelta :: Transaction -> Qty
netDelta tx = foldMap (\m -> let n = unPosQty (mQty m) in qneg n <> n) (txMoves tx)
-- law:  netDelta tau == mempty   for every tau (the empty/registration case too)

-- The canonical move-less instances. Trade/settlement/transfer use the record constructor
-- directly (txIntroduce = Nothing, moves + rows present).
registerTx :: UnitId -> TermsVersion -> UnitStatus -> Transaction   -- C5+C7+C10: introduction
registerTx u tv us = Transaction
  { txUnit = u, txMoves = [], txRows = Map.empty
  , txStatus = [], txIntroduce = Just (tv, us), txAppend = Nothing
  , txBoundary = Nothing, txCause = Nothing }

appendTx :: UnitId -> TermsVersion -> Transaction                   -- C6 Preserving amendment
appendTx u tv = Transaction
  { txUnit = u, txMoves = [], txRows = Map.empty
  , txStatus = [], txIntroduce = Nothing, txAppend = Just tv
  , txBoundary = Nothing, txCause = Nothing }

supersedeTx :: UnitId -> UnitId -> Transaction                      -- C8 Breaking: stamp old unit
supersedeTx u uFresh = Transaction
  { txUnit = u, txMoves = [], txRows = Map.empty
  , txStatus = [SetSupersededBy uFresh], txIntroduce = Nothing, txAppend = Nothing
  , txBoundary = Nothing, txCause = Nothing }

-- withCause stamps a follow-on or cascade with the parent notice's boundary id:
--   withCause :: BoundaryId -> Transaction -> Transaction
\end{lstlisting}

\paragraph{Per-field canonical writer (C11).}
The picture is a table naming, for each position field, the only writers allowed to touch
it; the type system holds every writer to that table. Each \emph{PositionState} field is
tagged with the closed set of field-writers permitted to mutate it: accumulated cost by
settle and trade; high-water mark by fee-crystallisation; entry NAV by subscription. The
balance field is not in this table. Its sole canonical writer is the \texttt{Move} edge
(handler \texttt{Transfer}), applied in \texttt{applyTx}, so a balance change without a
conserving move is unrepresentable. A write by a field-writer outside the permitted set
is a type error at the writer's authorship site. The tag is erased once writes share one
delta row, so the guarantee binds at authorship, not at the stored row
(\hyperref[cond:C11]{C11}).

These field-writers are a different axis from the event classes of
\hyperref[cond:C2]{C2}: one event class drives one or more field-writers, and the two
name-sets are not meant to coincide. The discipline is carried by a GADT (a generalised
algebraic data type, in which each constructor fixes its own type index) whose
constructors \emph{are} the field-to-writer table.

\begin{lstlisting}[language=Haskell]
data Handler = Settle | Trade | Transfer | FeeCrystallise | Subscribe deriving (Eq, Show)

-- The GADT constructors ARE the C11 field->writer table; no other pairing is representable.
-- The balance field is absent: its sole writer is the Move edge (handler 'Transfer), applied
-- in applyTx -- so "balance written off-ledger" is unrepresentable.
data FieldWrite (h :: Handler) where
  WAc       :: Qty -> FieldWrite 'Settle           -- ac, by settle
  WAcTrade  :: Qty -> FieldWrite 'Trade            -- ac, by trade
  WHwm      :: Qty -> FieldWrite 'FeeCrystallise   -- hwm, by fee crystallisation
  WEntryNav :: Qty -> FieldWrite 'Subscribe        -- entry_nav, by subscribe
deriving instance Show (FieldWrite h)

data SomeWrite where SomeWrite :: FieldWrite h -> SomeWrite  -- index erased once writes share a row
instance Show SomeWrite where show (SomeWrite w) = show w

applyWrite :: FieldWrite h -> PositionState -> PositionState  -- total over all constructors
applyWrite (WAc q)       p = p { psAc      = psAc p <> q }
applyWrite (WAcTrade q)  p = p { psAc      = psAc p <> q }
applyWrite (WHwm q)      p = p { psHwm      = qmax (psHwm p) q }                    -- monotone
applyWrite (WEntryNav q) p = p { psEntryNav = maybe (Just q) Just (psEntryNav p) } -- write-once

-- A settle handler authors ONLY 'Settle writes -- the result type IS the C11 checkpoint.
settleHandler :: [(WalletId, Qty)] -> Map WalletId [FieldWrite 'Settle]
settleHandler legs = Map.fromList [ (w, [WAc q]) | (w, q) <- legs ]

erase :: Map WalletId [FieldWrite h] -> Map WalletId [SomeWrite]  -- authorship -> erasure (S3)
erase = fmap (map SomeWrite)

_c11_ok_settle :: Qty -> FieldWrite 'Settle
_c11_ok_settle = WAc
_c11_ok_fee    :: Qty -> FieldWrite 'FeeCrystallise
_c11_ok_fee    = WHwm
-- _c11_bad :: Qty -> FieldWrite 'Settle
-- _c11_bad = WHwm    -- TYPE ERROR: WHwm :: FieldWrite 'FeeCrystallise, not 'Settle
\end{lstlisting}

\paragraph{Registration, settlement, moves, and amendment --- one door, \texttt{applyTx}.}
The picture is a single door: every write to the three maps passes through
\texttt{applyTx}, and nothing else writes. The ledger is sealed, and there is no second
write path. Registration is the move-less transaction whose \texttt{txIntroduce} is
\texttt{Just}. It writes its first terms version and default status as one entry, and it
refuses a unit already present: re-registration is a hard error, never a silent reset
(\hyperref[cond:C10]{C10}). Because terms and status are written together, ``registered
in terms $\iff$ registered in status'' holds by construction. \texttt{register} is
therefore \texttt{applyTx} of \texttt{registerTx}, not a separate operation.

A move's effect is its per-wallet net delta, written into the balance field and gated on
registration. Its two legs are exact inverses, so it sums to zero over wallets. A
zero-quantity or self-move nets to zero on every wallet and writes no net change.
\emph{Held} means named in a move that nets nonzero.

Amendment has two tracks, chosen by a product-declared total predicate
(\hyperref[cond:C8]{C8}). A fungibility-preserving amendment is one move-less transaction
appending a terms version, leaving positions untouched. A fungibility-breaking amendment
spans two units, so it is the composition of two transactions, run atomically through
\texttt{applyTx} by \texttt{amend}: introduce the successor $u_{\mathrm{new}}$, then
stamp \texttt{superseded\_by}$(u_{\mathrm{old}}\!\to u_{\mathrm{new}})$ on the
predecessor's status. Moving the holders is a separate paired-issuance event, never part
of the amendment. The two transactions are ordered so the successor is introduced before
the predecessor is stamped; the supersede pointer therefore always names a registered
unit. A \texttt{SetSupersededBy} whose target is absent is rejected as a
\texttt{DanglingSupersede}, so the pointer can never dangle. Reads are capability-scoped:
a cross-overlay read of another holder's position is forbidden, and strategy exports flow
only through \emph{UnitStatus} (\hyperref[cond:C4]{C4}).

\begin{lstlisting}[language=Haskell]
-- C5+C7+C10: register is applyTx of the move-less introduction. It is not a primitive
-- standing outside the fold; registration, trade, settlement and amendment all flow through
-- the ONE apply.
register :: UnitId -> TermsVersion -> UnitStatus -> Ledger -> Either LedgerError Ledger
register u tv us = applyTx (registerTx u tv us)

-- The single door. C3 (all-or-nothing) is structural: the WHOLE new Ledger or a Left, never
-- a partial state. txIntroduce=Just must introduce an ABSENT unit (else ReRegistration, C10),
-- writing PT and US together (C5/C7); txUnit and every unit a move names must be registered
-- (else UnknownUnit), so PT<->US co-presence is preserved; rows insert/update, never delete
-- (C1 monotone). Conservation is already a property of txMoves (edges) -- nothing to validate.
applyTx :: Transaction -> Ledger -> Either LedgerError Ledger
applyTx tx l0 = do
    l1 <- introduce l0
    ()  <- requireKnown l1
    ()  <- requireSupersedeTargets l1
    ()  <- requireBasisTip l1      -- tip weld     \ bodies elided; the origin weld guards
    ()  <- requireInvariance l1    -- invariance weld / introduce. Code in the reference.
    Right (commit l1)
  where
    u = txUnit tx

    introduce l = case txIntroduce tx of
      Nothing -> Right l
      Just (tv, us)
        | u `Map.member` ledgerPT l -> Left (ReRegistration u)              -- C10
        | usBasis us /= Origin u    -> Left (ForeignOrigin u)  -- origin weld: C5 stays a construction
        | otherwise -> Right l
            { ledgerPT = Map.insert u (ProductTerms (tv :| [])) (ledgerPT l) -- C7
            , ledgerUS = Map.insert u us                        (ledgerUS l) -- C5
            }

    requireKnown l =
      case filter (\x -> not (x `Map.member` ledgerPT l)) (u : map mUnit (txMoves tx)) of
        (bad : _) -> Left (UnknownUnit bad)
        []        -> Right ()

    -- Referential integrity for the supersede pointer: a SetSupersededBy target must name a
    -- registered unit, else the stamp would dangle. A Breaking amendment introduces the
    -- successor first, so its id is registered by the time supersedeTx is applied (C8).
    requireSupersedeTargets l =
      case filter (\x -> not (x `Map.member` ledgerPT l))
                  [ t | SetSupersededBy t <- txStatus tx ] of
        (bad : _) -> Left (DanglingSupersede bad)
        []        -> Right ()

    commit l = l
      { ledgerBounds = maybe (ledgerBounds l)
                             (\be -> Map.insertWith Set.union u
                                       (Set.singleton (bevTEff be, bevPrec be, bevId be))
                                       (ledgerBounds l))
                             (txBoundary tx)
      , ledgerPT = maybe (ledgerPT l)
                         (\tv -> Map.adjust (appendVersion tv) u (ledgerPT l))   -- C6 append
                         (txAppend tx)
      , ledgerUS = if null (txStatus tx)
                     then ledgerUS l
                     else Map.adjust (\s -> foldl' (flip applyStatus) s (txStatus tx)) u (ledgerUS l)
      , ledgerPS = foldl' (flip applyMoveRow)
                          (Map.foldrWithKey applyRow (ledgerPS l) (txRows tx))
                          (txMoves tx)
      }

    applyRow w writes ps =                                      -- non-balance bookkeeping
      let key = (w, u)
          cur = Map.findWithDefault zeroP key ps               -- first touch -> flat row
          new = foldl' (\acc (SomeWrite fw) -> applyWrite fw acc) cur writes
      in  Map.insert key new ps                                 -- insert/update only

-- A Move writes ONLY psBalance, via the conserving pair from moveDelta (from -= q, to += q),
-- keyed by the move's own unit -- the sole writer of the balance field.
applyMoveRow :: Move -> Map (WalletId, UnitId) PositionState
             -> Map (WalletId, UnitId) PositionState
applyMoveRow m ps0 = foldl' step ps0 (moveDelta m)
  where
    step ps (w, dq) =
      let key = (w, mUnit m)
          cur = Map.findWithDefault zeroP key ps
      in  Map.insert key cur { psBalance = psBalance cur <> dq } ps

-- C8: a product-declared total predicate chooses the track. Either track reaches the maps
-- ONLY through applyTx -- there is no second write path.
data Fungibility = Preserving | Breaking deriving (Eq, Show)
type FungibilityPredicate = ProductTerms -> TermsVersion -> Fungibility
data AmendResult = Appended UnitId | Superseded UnitId UnitId deriving (Eq, Show)

amend :: FungibilityPredicate -> UnitId -> TermsVersion -> UnitId -> Ledger
      -> Either LedgerError (AmendResult, Ledger)
amend isFungible u tvNew uFresh l =
  case Map.lookup u (ledgerPT l) of
    Nothing -> Left (UnknownUnit u)
    Just pt -> case isFungible pt tvNew of
      Preserving ->                                 -- ONE move-less Transaction; positions untouched
        (\l' -> (Appended u, l')) <$> applyTx (appendTx u tvNew) l
      Breaking                                      -- spans two units: a composite of TWO Transactions
        | uFresh `Map.member` ledgerPT l -> Left (UnitAlreadyExists uFresh)
        | otherwise -> do
            l1 <- applyTx (registerTx uFresh tvNew (defaultStatus uFresh)) l  -- introduce the successor
            l2 <- applyTx (supersedeTx u uFresh)                          l1  -- stamp the predecessor
            Right (Superseded u uFresh, l2)         -- old terms never rewritten (C7/P7)
\end{lstlisting}

\paragraph{Deterministic replay.}
The picture is the ledger rebuilt from nothing: the events are folded, in order, over the
empty ledger. Replay is the monadic left-fold of the event stream over the empty ledger.
It is deterministic because application is a pure, total function of the event and the
prior ledger: the same events give the same ledger. On a well-formed stream it returns
the rebuilt ledger. On an ill-formed one --- a repeated registration, or a move or
settlement on an unregistered unit --- it returns an error, the fold halting at the first
refusal. Checkpointing is sound by the monadic left-fold law: the fold over a
concatenation splits at any cut. Registration, settlement, and moves are all events, so
replay from the empty ledger rebuilds every unit's terms, status, and positions. Every
view --- balances, profit and loss, the cached \emph{UnitStatus} of
\S\ref{sec:states-unitstatus} --- is a projection of the stream. Row retention serves
audit, not determinism (P8).

\begin{lstlisting}[language=Haskell]
-- A Transaction IS the event; applyTx is the pure, total transition. Replay is the
-- error-stopping left fold of the stream. Determinism (P8) is the fold-homomorphism law,
--     replay (xs <> ys)  =  replay xs >=> replay ys,
-- so the result is independent of where a checkpoint is cut. On a well-formed stream (every
-- event on a registered unit) no Left arises, so replay is total there. (Balances and PnL are
-- further projections of the same stream -- netBal, in the balance-sheet section.)
replay :: [Transaction] -> Ledger -> Either LedgerError Ledger
replay ds l0 = foldM (\l d -> applyTx d l) l0 ds
\end{lstlisting}

\begin{remark}[Projection generalisation (optional)]
\texttt{balances = foldMap contribution} is a monoid homomorphism ---
\texttt{balances (xs <> ys) = balances xs <> balances ys} --- into the Balances monoid;
\emph{UnitStatus} is a left fold of its status-write stream and value is a fold of the
materialised position state, so every named view is a catamorphism of the log (directly for
balances and \emph{UnitStatus}, by composition with replay for value).
\end{remark}

\subsection{The conditions C1--C14}
\label{sec:states-conditions}

Each condition renders one illegal state unrepresentable. The labels are stable tags, not
a sequence. Each is stated once and forced at the site named. C1--C12 are forced
at the construction steps and instruments worked in \S\ref{sec:states-construction} and
\S\ref{sec:states-three}: a future whose state depends on past trades, a managed
account, a strategy that trades futures, and an instrument registered but never traded. C13 and C14 are stated and forced at the data plane's basis edge, in
\S\ref{sec:state-basis}.

\begin{longtable}{@{}p{0.8cm} p{11.0cm} p{2.5cm}@{}}
\toprule
 & \textbf{Condition} & \textbf{Forced at} \\
\midrule
\endhead
\hypertarget{cond:C1}{C1}\label{cond:C1} & Position accessor returns $\mathrm{Option}$
(never-held vs.\ held-and-flat distinct) \emph{and} the carrier is monotone (no row
deleted); both. & PositionState \\
\addlinespace
\hypertarget{cond:C2}{C2}\label{cond:C2} & Every \texttt{Transaction} carries its conserved
flow as edge moves, whose signed per-unit sum is \texttt{mempty} by construction; vacuous
base case (empty move list) included. & Transaction \\
\addlinespace
\hypertarget{cond:C3}{C3}\label{cond:C3} & A \texttt{Transaction} spans all three maps for
one unit and applies in full or not at all; partial application unrepresentable. &
Transaction \\
\addlinespace
\hypertarget{cond:C4}{C4}\label{cond:C4} & Reads are capability-scoped; cross-overlay reads
forbidden; strategy exports flow only through \emph{UnitStatus}. & strategy \\
\addlinespace
\hypertarget{cond:C5}{C5}\label{cond:C5} & \emph{UnitStatus} is registration-total:
product-declared defaults written at registration, including the basis origin. &
untraded unit \\
\addlinespace
\hypertarget{cond:C6}{C6}\label{cond:C6} & \emph{ProductTerms} is a non-empty version list,
append-only; no version mutated in place. & ProductTerms \\
\addlinespace
\hyperref[cond:C7]{C7} & \emph{ProductTerms} is registration-total: first
version written at registration. & ProductTerms \\
\addlinespace
\hypertarget{cond:C8}{C8}\label{cond:C8} & Two-track amendment by a total fungibility
predicate: \texttt{Preserving} appends a version; \texttt{Breaking} allocates a fresh $u$
and records \texttt{superseded\_by}. & amendment \\
\addlinespace
\hypertarget{cond:C9}{C9}\label{cond:C9} & A move-less event has an empty move list, which
folds to \texttt{mempty}, so it is conserved vacuously; conservation sums edge legs and never
divides by a holder count. & untraded unit \\
\addlinespace
\hyperref[cond:C10]{C10} & Re-registering an existing unit is a hard
error, never a silent reset. & registration \\
\addlinespace
\hypertarget{cond:C11}{C11}\label{cond:C11} & Each \emph{PositionState} field has a
canonical writer set; a writer outside it is a type error at authorship, erased at the
delta row. & PositionState \\
\addlinespace
\hypertarget{cond:C12}{C12}\label{cond:C12} & All per-(wallet, mandate/strategy) economic
state lives at \emph{PositionState}$[w,u_{\mathrm{MA}}]$; the $W$-sector is empty by
schema. & managed account \\
\addlinespace
\hyperref[cond:C13]{C13} & Every consumed datum names its joint basis; consumption
requires pointwise agreement with the prevailing total assignment or a ledger-authored
arrow (\S\ref{sec:state-basis}). & basis edge \\
\addlinespace
\hyperref[cond:C14]{C14} & Adjustment-schedule totality: registration and every terms
append admitted only if the schedule yields exactly one well-formed entry per
(corporate-action class, field); an append introducing a field extends the schedule in
the same transaction (\S\ref{sec:basis-schedule}). & basis edge \\
\bottomrule
\end{longtable}

\subsection{Why exactly three maps}
\label{sec:states-pareto-three}

Three independent constraints each force one map, and removing any one breaks its
constraint. \emph{(i)} Distinct per-holder state under the same contract forces a
$(w,u)$-keyed map: \emph{PositionState}. \emph{(ii)} Shared observables read identically
by every holder force a $u$-keyed map distinct from per-holder state: \emph{UnitStatus}.
\emph{(iii)} The two mutation disciplines --- append-only versioned terms versus an
overwrite-in-place status cache --- force a third, append-only map distinct from status:
\emph{ProductTerms}. A fourth map, the $W$-sector, adds a sector whose economic content
is empty (\hyperref[cond:C12]{C12}). Three maps are the minimum basis of the problem, not
a compromise.

\subsection{The unique design}
\label{sec:states-pareto}

The adopted design --- three maps under the conditions of \S\ref{sec:states-conditions},
design~B --- is the unique Pareto-optimum
on testability, correctness, and simplicity under any correctness gate. Each rejected
alternative fails on one forcing point.

\begin{itemize}[leftmargin=1.4em]
\item \textbf{A} (per-unit state, plus per-$(w,u)$ for futures only) has no sector for
shared observables distinct from immutable terms, so settlement prices and terms share
one overwrite cell; the two mutation disciplines of \S\ref{sec:states-pareto-three}(iii)
collapse.
\item \textbf{C} (a sentinel unit carrying wallet state) has a sentinel with no issuer and
no conservation partner, so $\sum_{w}h(\cdot)=0$ fails by fiat.
\item \textbf{D} (an explicit $W$-sector) adds a sector whose economic content is empty:
every candidate datum is $(w,u_{\mathrm{MA}})$-keyed (\hyperref[cond:C12]{C12}).
\item \textbf{E} (state indexed only over the held set) equals B on correctness but has no
implementation in available tooling and is far less simple; shippability decides, not
elegance.
\item \textbf{F} (a universe wallet $w_\star$ folding \emph{UnitStatus} into
\emph{PositionState}$[w_\star,u]$) re-introduces the per-unit/per-$(w,u)$ split as a
runtime predicate on the wallet axis, and conservation must then carve out $w_\star$ by
hand. \emph{UnitStatus} is the type-level expression of what F hides as a wallet-id
convention.
\end{itemize}

\subsection{Invariants made unrepresentable}
\label{sec:states-unrep}

Under design~B with C1--C12, seven illegal states are placed beyond reach rather than
merely tested; the further states C13's welds place beyond reach at the same door are
enumerated in \S\ref{sec:state-basis}. Each line names the hub invariant
(\S\ref{sec:invariants}) it discharges; where the guarantee is a contributing
mechanism rather than the invariant itself, the line names the hub invariant it
underpins. The P-numbers
are the canonical hub numbering of \S\ref{sec:invariants}. The mechanism differs by
invariant and is named in each line. ``Unrepresentable'' is precise. An abstract type
whose only constructor is a checked smart constructor keeps the illegal value from
reaching the operation that would act on it. A non-empty list makes the versionless state
untypable. A per-field tag is a type error at authorship. Conservation is carried by the
edge form of a move, whose two legs cancel by construction. The claim is not that every
guarantee is a type-level fact: read-scoping, in particular, is enforced at the
capability layer wrapping the accessors, not in the data reference.

\begin{longtable}{@{}p{2.0cm} p{6.7cm} p{5.8cm}@{}}
\toprule
\textbf{Invariant} & \textbf{Mechanism} & \textbf{Discharged by} \\
\midrule
\endhead
P1 (conservation of quantity) & Conserved flow is carried as \texttt{Move} edges, each
naming both legs (\texttt{qneg}\,$q \mathrel{\diamond} q = \mathit{mempty}$), so the signed
per-unit sum of any transaction is \texttt{mempty} by construction; there is no unconserved
\texttt{Transaction} to reject, hence no \texttt{validate} gate. & C2 (edge-conserved flow),
C3 (atomic transaction) \\
\addlinespace
P8 (snapshot/replay determinism) & Replay is the \texttt{foldM} of the stream, a fold
homomorphism: replaying a concatenated log equals replaying each part and composing; the
monotone carrier keeps the key set stable across any cut. & C1(b) (monotone carrier) \\
\addlinespace
P6 (lifecycle idempotency) & Lifecycle stage is $u$-keyed in \emph{UnitStatus},
written by replacement, so re-applying a stage write is the identity; non-conserved fields
carry the same algebra (high-water mark by $\max$, entry NAV write-once). & C5
(registration-total status) \\
\addlinespace
Terms immutability (underpins P4) & Append-only non-empty version list; no in-place
mutation, no exported setter. & C6 (append-only), C7 (registration-total) \\
\addlinespace
Identity immutability (underpins P4) & Re-registration rejected; a breaking amendment
allocates $u_{\mathrm{new}}$ and never rewrites $u_{\mathrm{old}}$. & C8 (two-track),
C10 (no re-registration) \\
\addlinespace
P7 (virtual/real isolation; read-scoping) & Cross-overlay reads forbidden at the capability
layer wrapping the accessors, not in the data reference. & C4 (capability-scoped reads) \\
\addlinespace
Handler-field canon (underpins P9) & Canonical writer set per field; a writer outside it is
a type error at authorship, erased once writes share a delta row. & C11 (per-field writer) \\
\bottomrule
\end{longtable}

No other candidate design places more than three of the core invariants beyond reach.

\subsection{Reference implementation}
\label{sec:states-reference}

The reference is \texttt{reference/Ledger.hs}, woven through \S\ref{sec:states-construction} and
collected here. It is total and deterministic, and makes the illegal states unrepresentable
rather than merely checked:
\begin{itemize}[nosep]
\item \texttt{Ledger} is sealed, with no row deleter (C1 monotone half) and an
\texttt{Option} accessor (C1 Option half);
\item a \texttt{Transaction} carries its conserved flow as \texttt{Move} edges, so the
signed per-unit sum is \texttt{mempty} by construction and no \texttt{validate} gate is
needed (C2, the empty move list giving C9);
\item \texttt{ProductTerms} is abstract over a non-empty list grown only by registration
and \texttt{appendVersion} (C6, C7);
\item the \texttt{FieldWrite} GADT is the C11 field-to-writer table;
\item \texttt{applyTx} is the single door, returning the whole ledger or an error with no
partial state (C3);
\item registration is the move-less \texttt{txIntroduce}=\texttt{Just} instance, rejecting
re-registration (C10) and writing terms and status together;
\item \texttt{amend} is the two-track C8.
\end{itemize}
Every \emph{UnitStatus} field is written only by \texttt{applyStatus}, folded inside
\texttt{applyTx} --- the projection-of-log discipline of \S\ref{sec:states-unitstatus} held
by construction.

The pieces woven through \S\ref{sec:states-construction} are excerpted from
\texttt{reference/Ledger.hs} and constitute its canonical core: the scalar layer, the
three homes, the \texttt{Transaction}, and the single door \texttt{applyTx}. Two elisions
are made, both marked in the listings: the basis-weld bodies, and the
\texttt{BoundaryEvent} and \texttt{Declaration} declarations, specified in
\S\ref{sec:state-basis} and carried in the reference. The remaining parts of the
reference serve their own sections and appendices. The \texttt{Move} the core builds on
is the reference form, magnitude \texttt{PosQty}; \S\ref{sec:ledger} presents the same
algebra in illustrative \texttt{Qty} form. Run on the reference source, the following
evaluation makes the behaviour concrete; the comments are the printed results.

\begin{lstlisting}[language=Haskell]
expect :: Show e => Either e a -> IO a
expect = either (error . show) pure

main :: IO ()
main = do
  let uES = UnitId "CME-ES"; wB = WalletId "BUYER"; wS = WalletId "SELLER"
  l0 <- expect $ register uES (TermsVersion "ES-v1" (Map.fromList [("mult","50")]))
                              (defaultStatus uES) emptyLedger
  print (position   l0 wB uES)        -- Nothing : never held (C1 Option half)
  print (unitStatus l0    uES)
  -- Just (UnitStatus {usLifecycle = Registered, usBasis = Origin (UnitId "CME-ES"),
  --   usLastSettle = Nothing, usSupersededBy = Nothing})   : the C5 default, basis at origin
  -- a conserving trade: +1000 to BUYER as a Move EDGE (balance), with accumulated-cost rows
  -- authored at 'Settle then erased (C11). (move builds the PosQty magnitude via mkPosQty.)
  Just mv <- pure (move wS wB uES (Qty 1000) (Timestamp 0) (SourceId "demo"))
  let tradeTx = Transaction uES [mv]
                  (erase (settleHandler [(wB, Qty 1000), (wS, Qty (-1000))]))
                  [] Nothing Nothing Nothing
  print (netDelta tradeTx)            -- 0 : the move legs cancel BY CONSTRUCTION (C2)
  l1 <- expect $ applyTx tradeTx l0
  print (psBalance <$> position l1 wB uES) -- Just 1000 : balance, written by the Move edge
  print (psAc      <$> position l1 wB uES) -- Just 1000 : accumulated cost, via the row
  -- There is no unconserved Transaction to reject: a one-legged flow is not a Move, so it is
  -- not representable in txMoves -- conservation is a property of the type, not a validate gate.
  -- close-out: a Move back keeps a flat row (monotone carrier): held-and-flat /= never-held
  Just mvC <- pure (move wB wS uES (Qty 1000) (Timestamp 1) (SourceId "demo"))
  l2 <- expect $ applyTx (Transaction uES [mvC] Map.empty [] Nothing Nothing Nothing) l1
  print (psBalance <$> position l2 wB uES) -- Just 0 : retained, not Nothing
  print (register uES (TermsVersion "dup" Map.empty) (defaultStatus uES) l2)
                                      -- Left (ReRegistration (UnitId "CME-ES")) (C10)
\end{lstlisting}

\begin{remark}[Discharged by construction]
The listings above, save the evaluation driver, are excerpted verbatim from
\texttt{reference/\allowbreak Ledger.hs}, the reference core, under the two marked
elisions. Here and wherever an excerpt is claimed verbatim, the claim is code-level:
the code tokens match the reference exactly; comments and line layout are re-set for
the page.
Quantities are exact \texttt{Integer} minor units, a correctness choice that makes
determinism and conservation arithmetic facts. Three structural points carry the
discharge: the per-unit homes are \emph{three coupled maps} (\texttt{ledgerPT}, \texttt{ledgerUS},
\texttt{ledgerPS}) with co-presence preserved by construction, as
\S\ref{sec:states-pareto-three} requires; the lifecycle is the closed sum
\texttt{Registered}/\texttt{Active}/\texttt{Expired}/\texttt{Closed} with the last settle
mark --- a \texttt{Price} paired with its basis --- a separate \texttt{Maybe} field written
only through a logged \texttt{StatusWrite}, rather than carried on an \texttt{Active}
constructor; and one atomic \texttt{Transaction}
carries both the conserved move edges and the non-balance state delta, so conservation holds
by construction and there is no separate validate step.
\end{remark}

The testing commitments for this construction --- handler mutation score 85--90\%,
lifecycle-guard 70--80\%, core overall $\geq80\%$, and a state-machine model bounded by a
finite generator universe --- are carried by the testing agenda
(\S\ref{sec:conclusion}, Appendix~\ref{app:properties}), not restated here.

\subsection*{Key Invariants and Consequences}

Unit state lives in three homes under one placement rule. The conditions of the register
(\S\ref{sec:states-conditions}) each render an illegal state unrepresentable rather than
merely tested; C13's identical claim is stated in \S\ref{sec:state-basis}.
Each guarantee carries its canonical number in \S\ref{sec:invariants}.

\begin{itemize}[nosep]
\item \textbf{One home, one writer.} Each economic fact has exactly one home and one
writer, so internal reconciliation failure cannot arise. The homes are immutable
\emph{ProductTerms}$[u]$, shared \emph{UnitStatus}$[u]$, and per-position
\emph{PositionState}$[w,u]$; the externally-authored per-(holder, unit) cell is empty by
construction (C12). (\S\ref{sec:states}.)
\item \textbf{UnitStatus is a projection, not a source of truth.} It holds one value per
unit, overwritten in place only as a read cache. The immutable event log is
authoritative, and replaying the events to any point rebuilds the exact value that held
then. This discipline discharges lifecycle idempotency P6---write-by-replacement at one
key---via C5. (\S\ref{sec:states-unitstatus}.)
\item \textbf{Conservation by construction (P1).} A \texttt{Transaction} carries its flow
as \texttt{Move} edges whose signed per-unit sum is the group identity; C2 (edge-conserved
flow) and C3 (atomic transaction across all three maps) leave no unconserved transaction
to reject. (\S\ref{sec:states}.)
\item \textbf{Snapshot/replay determinism (P8).} Replay is the fold of the stream; the
monotone carrier C1 keeps the key set stable across any checkpoint cut, so the result is
independent of where the cut falls. (\S\ref{sec:states}.)
\item \textbf{Terms and identity immutability (underpin P4).} \emph{ProductTerms} is an
append-only non-empty version list (C6, C7); re-registration is a hard error and a breaking
amendment allocates a fresh unit, never rewriting the old one (C8, C10).
(\S\ref{sec:states}.)
\item \textbf{Read scoping (P7).} Cross-overlay reads are forbidden at the capability layer
wrapping the accessors; strategy exports flow only through \emph{UnitStatus} (C4).
(\S\ref{sec:states}.)
\end{itemize}


