\section{Managed Accounts, Virtual Portfolios, and TRS}
\label{sec:managed}

A managed account is not a primitive. It is the composition of four primitives that already
exist (\S\ref{sec:ledger}, \S\ref{sec:states}). The first is a wallet partition that holds
positions and produces PnL. The second is a mandate unit $u_{\mathrm{MA}}$, issued by the
manager and held by the client, conserved by the issuance law. The third is a deterministic
reset contract that observes performance, crystallises one net cash flow, and resets its
baseline. The fourth is the three-home state model of \S\ref{sec:states}, which homes the
mandate's terms, the shared observables, and the per-client economic state. A total return
swap is the same composition with two changes: the wallet partition lives in a virtual
ledger, and the reset contract's settlement crosses into the real one. Both products reduce
to a single conservative cash move. Its magnitude is a function of observed performance and
a recorded baseline. The worked example (\S\ref{sec:managed-example}) carries one account
through subscription, trade, NAV, crystallisation against a high-water mark, and redemption,
with $\sum_w w(u)=0$ at every step.

\subsection{The mandate is a unit}
\label{sec:managed-mandate}

A managed account is opened by issuing the mandate unit $u_{\mathrm{MA}}$ from the manager
$w_M$ to the client $w_C$. This gives $w_M(u_{\mathrm{MA}})=-1$, $w_C(u_{\mathrm{MA}})=+1$,
$\sum_w w(u_{\mathrm{MA}})=0$. The issuance is a real two-named-wallet transfer with a real
issuer and holder, not a self-reflexive sentinel. Conservation holds by the issuance law (P1).

A \texttt{Move} is one conserved transfer of a unit between two named wallets:
$\text{from}\mathbin{-}\!=q$, $\text{to}\mathbin{+}\!=q$. Conservation (P1) holds by
construction, not by check: the per-wallet deltas are group negatives, so they sum to the
identity. The mandate issuance is the canonical instance. In code, the conservation witness
is explicit: a move's two deltas cancel.

\begin{lstlisting}[language=Haskell]
-- Move in abridged four-field form, used by this section's listings (illustrative);
-- the reference Move carries the magnitude as PosQty and adds time and source.
data Move = Move
  { mFrom :: WalletId, mTo :: WalletId, mUnit :: UnitId, mQty :: Qty }  -- q > 0

-- The per-wallet deltas a move induces; their sum is the conservation witness.
moveDelta :: Move -> [(WalletId, Qty)]
moveDelta (Move s d _ q) = [(s, qneg q), (d, q)]   -- qneg q <> q = mempty : conserved

-- Mandate issuance: a real two-named-wallet transfer of u_MA, manager -> client,
-- q = 1. A real issuer and holder, not a self-reflexive sentinel.
issueMandate :: WalletId -> WalletId -> UnitId -> Move
issueMandate manager client uMA = Move manager client uMA (Qty 1)
\end{lstlisting}

One mandate binds exactly one manager to exactly one client. The invariant states this as a
support condition on the unit's balances.

\begin{invariant}[Mandate is a singleton bilateral contract]
\label{inv:singleton}
While the mandate is live,
\[
\operatorname{support}\big(w\mapsto w(u_{\mathrm{MA}})\big)=\{(w_M,-1),(w_C,+1)\}.
\]
\end{invariant}

Conservation alone does not establish Invariant~\ref{inv:singleton}: P1 admits $(-5,+5)$ or
a three-way split. Four existing primitives together enforce the invariant across the
mandate's life, each covering one phase. Before issuance, the \texttt{Option} accessor reads
\texttt{None}, distinct from the \texttt{Some(zero)} of a terminated but retained mandate. The
\textit{UnitStatus} lifecycle guard admits issuance only from an unissued status, so a
terminated $u_{\mathrm{MA}}$ cannot be resurrected. Through the active phase, the singleton
support restriction (Invariant~\ref{inv:singleton}) holds. C8 routes a re-mandate to a fresh
$u'_{\mathrm{MA}}$ stamped \texttt{SetSupersededBy}.

The mandate entitles the client to an account whose contents are already valued. Pricing the
mandate itself would value that account a second time.

\begin{principle}[Mandate is non-valued]
\label{prin:nonvalued}
$u_{\mathrm{MA}}$ is excluded from the domain of valuation: $P_t(u_{\mathrm{MA}})$ is undefined,
not zero. A price on $u_{\mathrm{MA}}$ would add $+1\cdot P_t(u_{\mathrm{MA}})$ to the client's
per-wallet NAV and double-count the underlying exposure; the term cancels only in the global sum.
\end{principle}

\paragraph{What is recorded where.}

\begin{longtable}{p{8.0cm}p{6.4cm}}
\toprule
\textbf{Datum} & \textbf{Home} \\
\midrule
Mandate text, fee schedule, benchmark identity, position limits, HWM/hurdle methodology,
crystallisation frequency & \textit{ProductTerms}$[u_{\mathrm{MA}}]$ (immutable, versioned,
append-only) \\
Lifecycle stage, \texttt{superseded\_by} & \textit{UnitStatus}$[u_{\mathrm{MA}}]$ (shared) \\
HWM value, entry NAV, accrued fees, mandate-breach flags, subscription/redemption cursor,
benchmark NAV at this wallet's inception & \textit{PositionState}$[w_C,u_{\mathrm{MA}}]$ \\
Current benchmark level (externally sourced, captured as a logged observation event) &
\textit{UnitStatus}$[u_{\mathrm{bench}}]$ \\
\bottomrule
\end{longtable}

The benchmark level is a shared cached observable, governed by the same discipline as every
\textit{UnitStatus} value (\S\ref{sec:states-unitstatus}), and consumed in NAV under
single-basis consumption (Invariant~\ref{inv:basis}): the benchmark level and the account
balances it is struck against stand in one basis.

The per-client economic scalars are homed in \textit{PositionState}. Each is nevertheless a
fold of the event log. Whether they are stored or re-derived is escalation~ME1
(\S\ref{sec:conclusion}).

\subsection{Subscription}
\label{sec:managed-subscription}

Capital enters as a cash move from the client's external funding wallet $w_{\text{ext}}$
into $w_C$. $w_{\text{ext}}$ is a virtual wallet at the ledger boundary. The move is tagged
\texttt{SUBSCRIPTION} and indexed by the subscription/redemption cursor in
\textit{PositionState}$[w_C,u_{\mathrm{MA}}]$. The tag is load-bearing: it lets the net
external flow over any interval be reconstructed as a projection over the stream
(\S\ref{sec:managed-fees}). At subscription the entry NAV and the initial high-water mark
are set to the subscribed capital base. Before subscription the high-water mark is
\texttt{None} (unrepresentable), never $0$. A $0$ mark would charge a performance fee on the
entire contributed capital at the first crystallisation. The benchmark NAV at inception is
recorded from \textit{UnitStatus}$[u_{\mathrm{bench}}]$ for performance reporting.

\subsection{Trading under mandate}
\label{sec:managed-trading}

A trade is an ordinary two-legged transaction against a virtual market wallet $w_{\text{mkt}}$,
admitted only if it satisfies the mandate guards. A guard sits in front of trade admission:
either the trade passes and executes unchanged, or it is rejected and nothing happens.

\begin{definition}[Mandate guard]
A quantitative mandate constraint is a precondition $g$ on the move-generation function. The
contract evaluates $g$ against the current state and the recorded price snapshot; if $g$ fails, no
moves are emitted and the state is unchanged.
\end{definition}

Guard determinism is price-relative. ``Same wallet state $+$ same trade $\Rightarrow$ same
decision'' is false for value-based limits such as leverage, concentration-by-value, and
currency caps: holding positions fixed and doubling a price flips the same trade from admit
to reject. Determinism is recovered only when the guard is a pure function of
$(\textit{ProductTerms},\,\text{state},\,\text{trade},\,P_t)$ with $P_t$ an explicit,
snapshot-pinned input. Two totality obligations follow. First, a missing price
($P_t(u)=\texttt{None}$) makes the guard fail closed: a typed reject, never a silent admit.
Second, a passive breach --- a limit crossed by price appreciation with no admitted
trade --- lies outside the domain of a move-time guard and is detected only by a periodic
valuation sweep. The pinned-snapshot requirement is escalation~ME2
(\S\ref{sec:conclusion}).

\subsection{Valuation and NAV}
\label{sec:managed-nav}

The account NAV is the wallet valuation with the mandate excluded
(Principle~\ref{prin:nonvalued}):
\[
\mathrm{NAV}_t \;=\; \sum_{u\neq u_{\mathrm{MA}}} w_{C,t}(u)\,P_t(u).
\]
By state-sufficiency, $\mathrm{NAV}_t$ depends only on current balances, current unit state,
and current prices, not on history. Performance against the benchmark is computable from the
inception benchmark NAV and the current level in \textit{UnitStatus}$[u_{\mathrm{bench}}]$.
It feeds reporting. It feeds the fee only when the mandate's methodology names a benchmark
hurdle.

\subsection{Fee logic}
\label{sec:managed-fees}

\paragraph{Performance is net of capital flows.} Over a reset interval $[t_{k-1},t_k]$ the
performance is the gross NAV change less the net external flow:
\[
\mathrm{Perf}_k \;=\; \big(\mathrm{NAV}_{t_k}-\mathrm{NAV}_{t_{k-1}}\big)\;-\;
\mathrm{NetFlow}_{[t_{k-1},t_k]},\qquad
\mathrm{NetFlow}=\!\!\sum_{\substack{m\ \text{tagged sub/redemption}\\ t_{k-1}<m.t\le t_k}}\!\!\pm\,m.q.
\]
The gross change conflates performance with capital flows. By the PnL decomposition, a
mid-period subscription is a flow; crystallising it as performance would charge a fee on the
client's own contributed capital. $\mathrm{NetFlow}$ is a projection over the already-tagged
cursor moves (\S\ref{sec:managed-subscription}), so the correction adds no new coordinate.

\paragraph{Fee is not PnL settlement.} A client managed account pays the manager only a fee
fraction; the gain stays in the client book. This differs from the internal-desk case, where
a single move of the whole $\mathrm{Perf}$ to a beneficiary sweeps a book to Treasury. Two fees
are computed at each reset: the management fee accrues on assets in every period, and the
performance fee applies only to gains above the high-water mark, which never decreases. The
fee handler is total over $\operatorname{sign}(\mathrm{NAV}-\mathrm{HWM})$:
{\small\begin{align*}
\text{management fee } f^{m}_k &= r_{\text{m}}\cdot\Delta t_k\cdot \mathrm{NAV}^{\text{base}}_k
&&\text{\footnotesize(accrues on AUM regardless of sign),}\\
\text{performance fee } f^{p}_k &= r_{\text{p}}\cdot\max\!\big(0,\ \mathrm{NAV}^{\text{net}}_k-\mathrm{HWM}_{k-1}\big)
&&\text{\footnotesize(floored at zero; no clawback),}\\
\mathrm{HWM}_k &= \max\!\big(\mathrm{HWM}_{k-1},\ \mathrm{NAV}^{\text{net}}_k-f^{p}_k\big)
&&\text{\footnotesize(monotone ratchet; C11 writer \texttt{fee\_crystallise}),}
\end{align*}}
where $\mathrm{NAV}^{\text{net}}_k=\mathrm{NAV}_{t_k}-f^{m}_k$. On a loss period the
performance fee is $0$ and the high-water mark is unchanged. The management fee still
accrues.

\paragraph{Crystallisation is double-entry.} A fee is realised as a real cash move
$w_C\to w_M$, so it is conserved and carries its own contra-entry. A non-conserved stored
scalar would be single-entry and sit outside P1.

A signed amount $x$ crystallises into \emph{at most one} conserved cash move: magnitude
$|x|$, direction by $\operatorname{sign}(x)$, none at $x=0$. The \texttt{Maybe} makes this
total: $x=0$ maps to \texttt{Nothing}, never to an illegal $q=0$ move.

The fee handler emits two such moves $w_C\to w_M$ atomically (C3) and ratchets the
high-water mark through one canonical writer, \texttt{WHwm :: FieldWrite 'FeeCrystallise},
reused verbatim from the three-home model (\S\ref{sec:states}). The phantom type index
carries the handler's name in the write's type, so a settle or trade handler bumping the HWM
is a compile error: the mark has exactly one author. \texttt{applyWrite (WHwm \_)} applies
the \texttt{qmax} ratchet, so the monotone law above holds by construction, not by a runtime
branch.

\begin{lstlisting}[language=Haskell]
-- Crystallise(magnitude, from, to): |x|, direction by sign(x), no move at x = 0.
crystallise :: Qty -> WalletId -> WalletId -> Maybe Move
crystallise (Qty x) from to
  | x > 0     = Just (Move from to usd (Qty x))            -- from -> to
  | x < 0     = Just (Move to   from usd (Qty (negate x))) -- direction flips
  | otherwise = Nothing                                    -- no move at zero

data FeeEvent = FeeEvent
  { fcMgmt :: Qty, fcPerf :: Qty, fcNavNet :: Qty }   -- f_m, f_p, NAV net of f_m

-- Two cash moves w_C -> w_M (mgmt, perf), atomic (C3), + the HWM ratchet write.
-- f_m, f_p >= 0; on a loss period f_p = 0, so crystallise returns Nothing and the
-- performance move is absent (no zero-quantity move).
crystalliseFees :: WalletId -> WalletId -> FeeEvent
                -> ([Move], FieldWrite 'FeeCrystallise)
crystalliseFees wC wM fc =
  ( [ m | Just m <- [ crystallise (fcMgmt fc) wC wM
                    , crystallise (fcPerf fc) wC wM ] ]
  , WHwm (fcNavNet fc <> qneg (fcPerf fc)) )   -- HWM_k = max(HWM_{k-1}, NAV^net - f_p)
\end{lstlisting}

\noindent At the worked example's Step~E,
\texttt{crystalliseFees wC wM (FeeEvent (Qty 5750) (Qty 28850) (Qty 1144250))} yields the
two legs \texttt{Move wC wM USD (Qty 5750)} and \texttt{Move wC wM USD (Qty 28850)} together
with \texttt{WHwm (Qty 1115400)}. Applied to a row with $\mathrm{HWM}_{k-1}=1{,}000{,}000$,
the \texttt{qmax} ratchets the mark to $1{,}115{,}400$. On a loss period $f^p_k=0$ leaves
one leg and the mark unchanged.

The fee is a transfer, not an accrual: whatever leaves the client wallet arrives in the
manager wallet in the same transaction.

\begin{theorem}[Fee zero-sum per crystallisation]
\label{thm:fee-zerosum}
$\sum_w \Delta_{\text{fee}}(\text{USD})=0$ at each crystallisation: the client pays exactly what
the manager receives, creating and destroying no value.
\end{theorem}
\noindent\textit{Proof.} Each leg is $w_C\mathbin{-}\!=q;\ w_M\mathbin{+}\!=q$; summing over
wallets gives $-q+q=0$ per leg, hence $0$ over the transaction. \qed

\paragraph{Quantity, baseline, residual.} Every settlement leg here is a \texttt{crystallise}
call (above). The reset baseline is the post-settlement capital base
$B_k=\mathrm{NAV}_{t_k}-(f^{m}_k+f^{p}_k)+\mathrm{NetFlow}$. Reading the pre-settlement NAV
instead would claw the distributed performance back next period. The rounding residual
$\mathrm{Perf}_k-\operatorname{round}(\mathrm{Perf}_k)$ is retained in $w_C$ as a conserved
un-crystallised remainder, never dropped. It is the reconciling item between economic
performance and moved cash. Whether $B_k$, $\mathrm{HWM}_k$, and the accrued fee are stored
or re-derived is escalation~ME1 (\S\ref{sec:conclusion}). Crystallising unrealised
mark-to-market can drive the cash wallet negative. Conservation alone cannot distinguish a
funded obligation from an insolvent overdraft; this liveness gap is recorded as
escalation~ME5 (\S\ref{sec:conclusion}).

\subsection{Segregation}
\label{sec:managed-segregation}

Conservation checks that a move balances; it never checks whose wallets the move names.
Keeping clients apart is therefore a question of authorisation, not of arithmetic.

\begin{theorem}[Segregation]
\label{thm:seg}
Client partitions are mutually isolated if and only if conservation is combined with locality and
capability scoping. A cross-client move is an illegal state ruled out by C4 under locality, not by
conservation.
\end{theorem}

Conservation alone is not segregation. A move $m=(w_A,w_B,\text{USD},q)$ between two
different clients' partitions is perfectly conservative ($\sum_w\Delta=0$), so P1 does not
forbid it. Two independent facts are at work: conservation (CONS, $\sum_w\Delta=0$) and
locality (LOC, $\operatorname{supp}(\Delta)\subseteq\{w_s,w_d\}$). Locality gives ``no
effect beyond the named wallets''. Isolation of distinct clients additionally requires an
authorisation predicate: C4 capability scoping, which forbids a handler from naming or
reading across a $(w,u_{\mathrm{MA}})$ boundary it has no capability for. Segregation is
therefore a theorem under CONS $\wedge$ LOC $\wedge$ C4. It supplies the logical-segregation
component of CASS~6 / MiFID~II~Art.~16(8). Legal segregation, meaning distinct custodial
accounts and trust or nominee structure, is outside scope: the ledger evidences it, it does
not establish it. That C4 is asserted in prose rather than typed on the accessor signature
is escalation~ME4 (\S\ref{sec:conclusion}).

\subsection{CSA margin and collateral}
\label{sec:managed-csa}

The CSA margin contract is a wallet-level smart contract attached to a per-counterparty
collateral wallet. It reads, in the real ledger, the aggregate mark-to-market of the trades
the CSA governs. It computes the required collateral under the threshold,
minimum-transfer-amount, eligible-collateral, and haircut rules. It emits margin-call moves
to or from the collateral wallet. Margin eligibility and haircut are contractual data in
\textit{ProductTerms}; the exposure is the observed valuation.

For a synthetic managed account the netting set is the single real unit $u_{\mathrm{TRS}}$,
never the simulated constituents in the virtual ledger. Reading those constituents would
breach the isolation invariant (\S\ref{sec:managed-trs}) and over-margin unmargined
positions. The realm tag and C4 realm-scoped reads enforce this. The margin exposure
$N_k\cdot\mathrm{TR}_k$ is itself an observation of the virtual book, lifted across the
realm boundary as a number. That this observation is unattested and unversioned is
escalation~ME2 (\S\ref{sec:conclusion}).

\subsection{TRS as synthetic managed account}
\label{sec:managed-trs}

A total return swap connects a virtual ledger $\mathcal{L}_v$ to real counterparty wallets
in the real ledger $\mathcal{L}_r$ through periodic cash settlement. Ledger isolation is
invariant~P7: no move has one endpoint in $\mathcal{L}_v$ and the other in $\mathcal{L}_r$.
A minimal virtual ledger is not a second engine. It is a
$\mathrm{realm}\in\{\text{real},\text{virtual}\}$ tag enforcing P7.

At reset $t_k$ the TRS contract in $\mathcal{L}_r$ observes the virtual book's performance
and emits one real move, $w_{\text{payer}}\to w_{\text{receiver}}$ in USD, tagged
\texttt{TRS\_NET\_SETTLEMENT}: magnitude $\big|\mathrm{Payment}_k\big|$, direction by sign,
where $\mathrm{Payment}_k=N_k\cdot\mathrm{TR}_k-N_k\cdot r_k\cdot\Delta t_k$. The
total return $\mathrm{TR}_k=(V^v_{t_k}-V^v_{t_{k-1}})/V^v_{t_{k-1}}$ is partial: it requires
$V^v_{t_{k-1}}>0$, and failure of that precondition is a typed failure, not a division by
zero.

Viewed from the ledger, a funded managed account and a TRS do the same thing at each reset:
observe a performance, compare it with a recorded baseline, and pay one net cash amount.

\begin{theorem}[Equivalence to the funded managed account]
\label{thm:trs-equiv}
The TRS settlement and the funded managed-account fee/PnL settlement are the same ledger
operation: one net conservative cash move whose magnitude is a deterministic function of an
observed performance and a recorded baseline. The funded book plays the role of $\mathcal{L}_v$;
the periodic settlement is the TRS payment.
\end{theorem}
\noindent\textit{Proof sketch.} Both reduce to
$\mathrm{Crystallise}(\text{magnitude},\text{from},\text{to})$ with magnitude $=$ observed
performance over the interval and direction by sign; both reset a baseline to the post-settlement
value; neither moves a position across a boundary. The settlement primitive is one; the products
differ, and that difference is a CDM (reporting) distinction, not a ledger one: a CDM
\texttt{TotalReturnSwap} is two-legged ($N\,\mathrm{TR}$ and $N\,r\,\Delta t$), so the report
derives from the retained CDM \texttt{BusinessEvent}, never from the netted move
(\S\ref{sec:managed-conformance}). \qed

\paragraph{Price consistency.} $\mathcal{L}_v$ valuation and $\mathcal{L}_r$ settlement use
one price vector $P_t$. Because P7 forbids a cross-ledger move, a price divergence produces
unexplained PnL with no internal reconciliation path. This is the one place the
no-internal-break guarantee can be silently voided. Both realms read the price from the same
logged event, so a fresh replay reconstructs one price for both. What remains is attestation
of that logged price: escalation~ME2, the attestation-envelope requirement
(\S\ref{sec:conclusion}).

\subsection{Redemption and wind-down}
\label{sec:managed-redemption}

Wind-down strikes a final NAV, settles the final fee, returns capital, and retires the
mandate. Positions are liquidated against the market wallet. The cash is paid to the
client's external wallet. The mandate unit is returned to the manager. All closed-out rows
are retained at zero under the monotone carrier (C1). This preserves the audit trail, the
final high-water mark, and the quantities needed for tax reconstruction.
\begin{lstlisting}
Move(from: w_C,     to: w_mkt, unit: <asset>, quantity: <held>)      -- liquidate
Move(from: w_mkt,   to: w_C,   unit: USD,     quantity: <proceeds>)
Move(from: w_C,     to: w_ext, unit: USD,     quantity: <final NAV>) -- capital-out
Move(from: w_C,     to: w_M,   unit: u_MA,    quantity: 1)           -- retire mandate
\end{lstlisting}

\subsection{Account-level balance-sheet substantiation}
\label{sec:managed-substantiation}

Each account's balance at any date is a deterministic projection of the move stream filtered
to its wallets. Client statements, desk PnL, and regulatory position reports differ only in
the filter. Within the ledger there is no separate account-level quantity record to
reconcile against. The move stream substantiates quantities and provenance. Valuations
depend on the external price vector; disclosures and legal status require external evidence.
Net settlement is operational netting. Presenting a balance sheet net requires, in addition,
a legally enforceable right of set-off and the intent to settle net (IAS~32.42 /
ASC~210-20), a determination outside the ledger. The accounts'
assets~$=$~liabilities~$+$~equity reconciliation holds at the quantity level by P1: the
trial balance balances by construction. The valued and classified presentation does not
follow from P1 alone.

\subsection{Worked example}
\label{sec:managed-example}

One managed account $w_C$ under mandate $u_{\mathrm{MA}}$ issued by manager $w_M$; virtual
market wallet $w_{\text{mkt}}$; client external wallet $w_{\text{ext}}$. USD reference.
Terms: management fee $2\%/\text{yr}$, performance fee $20\%$ over a high-water mark (no
hurdle), quarterly ($\Delta t=0.25$). Conventions: the management fee accrues on
end-of-period NAV; the performance fee is charged on NAV net of the management fee, above
the mark; the mandate is non-valued. $\sum_w w(u)=0$ holds for every unit after every move.

\begin{longtable}{p{4.3cm}p{7.2cm}r}
\toprule
\textbf{Step} & \textbf{Moves} ($\text{src}\to\text{dst}:\text{qty unit}$) & \textbf{NAV} \\
\midrule
A. Mandate issuance & $w_M\to w_C:1\ u_{\mathrm{MA}}$ & --- \\
B. Subscription & $w_{\text{ext}}\to w_C:1{,}000{,}000\ \text{USD}$ & $1{,}000{,}000$ \\
C. Trade (buy $5{,}000$ AAPL @100) & $w_C\to w_{\text{mkt}}:500{,}000\ \text{USD}$;\ \ $w_{\text{mkt}}\to w_C:5{,}000\ \text{AAPL}$ & $1{,}000{,}000$ \\
D. Q1 NAV (AAPL $\to 130$) & (price move; no ledger move) & $\mathbf{1{,}150{,}000}$ \\
E. Fee crystallisation & $w_C\to w_M:5{,}750\ \text{USD}$ (mgmt);\ \ $w_C\to w_M:28{,}850\ \text{USD}$ (perf) & $\mathbf{1{,}115{,}400}$ \\
F. Wind-down & $w_C\to w_{\text{mkt}}:5{,}000\ \text{AAPL}$;\ \ $w_{\text{mkt}}\to w_C:650{,}000\ \text{USD}$;\ \ $w_C\to w_{\text{ext}}:1{,}115{,}400\ \text{USD}$;\ \ $w_C\to w_M:1\ u_{\mathrm{MA}}$ & $0$ \\
\bottomrule
\end{longtable}

\paragraph{Fee arithmetic (Step~E).} No capital flow occurred inside the period, so
$\mathrm{Perf}=(1{,}150{,}000-1{,}000{,}000)-0=150{,}000$. Management fee
$=1{,}150{,}000\times0.02\times0.25=5{,}750$. NAV net of management $=1{,}144{,}250$. Performance
fee $=0.20\times(1{,}144{,}250-1{,}000{,}000)=28{,}850$. NAV after both fees $=1{,}115{,}400$. New
high-water mark $=\max(1{,}000{,}000,\,1{,}115{,}400)=1{,}115{,}400$; reset baseline
$B_1=1{,}115{,}400$. Fee zero-sum (Theorem~\ref{thm:fee-zerosum}):
$\sum_w\Delta_{\text{fee}}(\text{USD})=-34{,}600\,(w_C)+34{,}600\,(w_M)=0$.

\paragraph{Closing identity (all wallets).} Final non-zero balances: $w_M=+34{,}600$ (fee
revenue), $w_{\text{mkt}}=-150{,}000$ (the gross gain paid out by the market),
$w_{\text{ext}}=+115{,}400$ (client net). Their sum is
$34{,}600-150{,}000+115{,}400=0$. The client received principal $1{,}000{,}000$ plus net
gain $115{,}400$. The gross gain $150{,}000$ equals net $115{,}400$ plus fees $34{,}600$.
Per-unit conservation holds across the chain. $u_{\mathrm{MA}}$ is issued $(+1,-1)$ then
returned $(0,0)$. USD nets to $0$ at $w_C$ and globally after every move. AAPL is acquired
and liquidated with $w_{\text{mkt}}$ as the mirror. Every position row is retained at zero
under the monotone carrier.

\subsection{Conformance and flags}
\label{sec:managed-conformance}

No conformance is asserted; each item is demonstrated against a cited source or flagged for
the responsible owner. The CDM strata named below (\texttt{LegalAgreement}, \texttt{Trade},
\texttt{TotalReturnSwap}, \texttt{TradeState}, \texttt{BusinessEvent}) are identified at the
type level only. Pinning them to a named CDM release is owed to the CDM owner and is
flagged, not asserted.

\begin{itemize}[leftmargin=*]
\item \textbf{CDM mandate representation.} $u_{\mathrm{MA}}$ maps to the CDM
\texttt{LegalAgreement} stratum, the ISDA Create-digitised investment-management agreement:
parties, fee schedule, benchmark identity, methodology, limits. It does not map to the
\texttt{Trade}/product stratum. A discretionary mandate is a legal agreement, not a tradable
contract.
\item \textbf{CDM TRS representation.} A CDM \texttt{TotalReturnSwap} is two-legged. The
report is the retained \texttt{BusinessEvent} (performance leg and financing leg
$N\,r\,\Delta t$), never the netted settlement move. The ledger ``same mechanism'' identity
(Theorem~\ref{thm:trs-equiv}) is a settlement-primitive identity; it is flagged as not a
CDM/reporting identity.
\item \textbf{SFTR/EMIR reporting surface.} $u_{\mathrm{MA}}$ issuance is not reportable: a
discretionary mandate is the MiFID~II Annex~I~\S~A(4) portfolio-management service, not a Section~C
instrument, not an SFT, not a derivative.
The reportable artefact is never $u_{\mathrm{MA}}$; it is the underlying book trade or
$u_{\mathrm{TRS}}$, each carrying its own UTI/UPI/LEI. Reportability is a per-relationship,
regime-cited attribute on the wrapped unit's \textit{ProductTerms}; it is not inferable from
the coarse unit-type key. Governance of that attribute is escalation~ME3
(\S\ref{sec:conclusion}). So is the dual-sided EMIR/CFTC/SFTR reporting, which needs
validated ISO-17442 counterparty LEIs and classification. Neither is a quantity fact;
neither has a home in the move stream.
\item \textbf{IFRS classification.} Economic NAV and economic PnL are derived here.
Accounting classification of fair-value changes (FVTPL/FVOCI/amortised cost) and net
presentation (IAS~32.42) are outside scope and require external evidence. Accrued fees are
modelled as conserved moves rather than stored scalars, so they carry their own contra-entry
and are auditable as double-entry.
\item \textbf{CDM \texttt{TradeState} alignment.} That the CDM
\texttt{TradeState}-per-\texttt{Trade} view and \textit{PositionState}$[w,u]$ stay
replay-consistent (both projections of one minted event) is asserted, not yet verified. It
is the outstanding F6 deliverable: a Rosetta mapping rerun against the three-map schema.
\end{itemize}

Escalations ME1--ME5 and the attestation-envelope requirement are cited at their points of
origin above and collected in the flagged register (\S\ref{sec:conclusion}). The workflow is
sound as quantity algebra: every step is atomic $q>0$ moves with $\sum_w w(u)=0$ verified,
and the worked example closes to the penny. The escalations mark the points at which the
performance and reporting layer outruns the framework's treatment of recorded observations
and external identity.
