\documentclass[11pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[margin=0.88in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{enumitem}
\usepackage{microtype}
\usepackage{array}
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{7pt}{3pt}
\titlespacing*{\subsection}{0pt}{5pt}{2pt}
\titleformat*{\section}{\large\bfseries}
\titleformat*{\subsection}{\normalsize\bfseries}
\setlist{nosep,leftmargin=1.4em}
\setlength{\parskip}{1.5pt}
\setlength{\parindent}{0pt}
\newtheorem{invariant}{Invariant}
\title{\vspace{-3.5em}Deferred Settlement, v16-Native\\[1pt]
\large The open window between trade-date recognition and custody movement ---\\
\large deferred-settlement design cell, converged Round 10, seven approvals \quad
{\normalsize 2026-07-16}}
\author{}
\date{}
\begin{document}
\maketitle
\vspace{-4.5em}

\section{The Problem}

Equity markets settle on a delay. On Monday at 14:32:11 UTC a portfolio wallet $w$
buys 100 XYZ at \$50.00 --- \$5{,}000.00 against 100 shares, T+2, settling
Wednesday. The trade is economically complete on Monday: exposure held, cash owed,
dividends due from that instant. The custody fact happens on Wednesday, when the
depository moves the shares and the cash agent debits the nostro. Between the two
lies the open settlement window $[T, t_d]$; its length is a parameter --- T+1, T+2,
T+0, T+5 on a cross-border fail --- never a doctrine.

The window makes a number real before anything moves. XYZ closes at \$52.00 on
Tuesday. The book must show $+\$200.00$ of unrealised profit at Tuesday's close ---
recognised, capital-bearing, reported --- while the custodian still reports
\$1{,}000{,}000.00 and zero XYZ, exactly as on Monday morning. Both records are
right: different facts, different calendars, different authorities. Collapse them
and the system fails one way or the other: debit the nostro at $T$ and the
statement breaks by \$5{,}000 all window with nothing left to absorb Wednesday's
real debit; recognise nothing until $t_d$ and Tuesday's \$200 does not exist ---
profit and loss, risk, and capital misstated. The design must keep both facts,
separately observable and separately attested, bound by an identity computable
inside the system. A bound living in prose is checked in a spreadsheet, and the
spreadsheet is a second store of truth --- the defect this ledger abolishes.

v16.0 already carries the machinery: trade-date booking as a declared term, the
settlement-obligation unit with its \texttt{instructed}/\texttt{settled}/\texttt{failed}
node, the partial edge, market claims, the event-kind registry, the bitemporal
doctrine. What it lacked was the bound (gap NS-01). This document is that bound.

\section{The Design}

Everything below is recorded events and folds of them. No state exists outside the
log: no mirror wallet, no stored settlement status, no stored break register, and the
execution substrate owns no fact.

\subsection{Events: the witness kinds}

External assertions cross the one door as registered, typed, moveless observations
--- provenance carried, quarantined when malformed, superseded only at the tip of a
per-(account, as-of) chain, never edited. Three asserter classes, one trust
assumption (TA-CUSTODY, \S5):

\begin{itemize}
\item \textbf{Balance statements} (camt.053, the depot statement): per account and
unit, a balance at an as-of time; fields account, coordinate, balance in integer
minor units, as-of, statement-id, supersedes-ref. An unresolvable account
quarantines and re-admits when the reference resolves. Each account declares a
cadence on its venue calendar; a missed cadence mints an overdue item --- a dark
custodian never leaves a stale CLEAN standing. Statements are authenticated under
an identified key; keys live in a valid-timed registry, and a revocation carries a
compromise-effective time $t_c$ scoping re-quarantine deterministically.
\item \textbf{Settlement and cash confirmations} (sese.025, camt.054), including
partial clips, always recorded --- a clip below the venue's declared minimum fill
size accumulates on the one live residual instead of splitting it.
\item \textbf{Penalty advices and emission receipts} (semt.044; dispatch-intent,
dispatch receipt or CSD acknowledgment, match confirmation with its matching
timestamp, cancellation confirmation). The dispatch-intent is recorded
\emph{before} the send --- the egress dual of witness-before-signal --- and keys
the repair discriminator. The dispatch receipt is an internal-egress witness (our
gateway attesting our own send), its detection indirect. The matched date and
timestamp are CSD-asserted, single-source: bounded and confronted, never verified.
\end{itemize}

\subsection{The fold}

Each event kind folds into a fixed consequence. Instruction: the paired
\texttt{owned} re-book and the moveless minting of the settlement-obligation unit at
\texttt{instructed} (legs, due date, positive inherited minimum fill size, root
settlement-instruction reference, per-unit settlement channel) --- on a declared
atomic-finality venue (a TA-TERMS term) the unit is born \texttt{settled} and no
timer is armed, so the T+0 race is unrepresentable. Confirmation: the moveless walk
to \texttt{settled}. Partial clip at or above the floor: the certified split ---
settled leg plus one live residual, root reference inherited. Fail notice, or the
due-date watch reading cumulative confirmed quantity short of instructed: the walk
to \texttt{failed} and the fails cascade. Repair event: in-place repair when no
dispatch-intent is recorded; otherwise cancel-and-re-instruct on the recorded
cancellation confirmation, one atomic transaction. Recall: a paired lent-coordinate
move plus a possession-plane return leg; a terminally failed return routes to
close-out, the defaulter's wallet receiving the securities it cannot return, paired
with the lender's write-off, a valued claim minted against collateral. Custody
statement: no balance change --- the reconciliation projection reads it.
Reconciled penalties book, per (participant, currency, period), an \texttt{owned}
charge plus a penalty settlement-obligation unit discharged by the collection
advice. A buy-in delivery discharges the failed residual it compensates; it never
books a second long.

CSDR penalties fold over dated confirmation clips keyed by root reference, on
sequential windows: late-matching over $[\mathrm{ISD}, M)$ or $[\mathrm{ISD}, M]$
as the matching timestamp falls before or after the venue's declared settlement
cut-off on day $M$; settlement-fail over $[\max(\mathrm{ISD}, M),\,
\min(\text{settlement},\text{buy-in},\text{cancellation}))$ --- ISD inclusive, all
right endpoints exclusive, no day charged under both. Each advice line passes
three decidable checks: its own arithmetic, exact (quantity $\times$ its reference
price $\times$ rate $=$ its amount); its reference price against that business
day's recorded close from the SECR-designated source, within a governed
de-minimis tolerance; its failing quantities against our clip fold. CLEAN or
BREAK --- no fuzzy middle class. A lateness charge against us is confronted with
the matched instruction's own send-completion witness; if that witness is absent
from the log, the advice line quarantines as an open item, never a silent CLEAN.

\subsection{The invariants}

\begin{invariant}[The reconciliation identity]\label{inv:identity}
For every custody account $a$, unit $u$, and recorded statement with as-of time
$t$, the residual
\[
r(a,u,t) \;=\; \mathrm{stmt}(a,u,t) \;-\; \sum_{w \in W(a)}
\bigl[\mathrm{owned} - \mathrm{lent\_out} + \mathrm{borrowed}
+ \mathrm{inflight\_out} - \mathrm{inflight\_in}\bigr](w,u,t)
\]
is zero within the tolerance declared for $(a, u)$. $W$ maps each wallet's mass to
exactly one account per coordinate class (total, disjoint, covering); the in-flight
sums run over the directed legs of $w$'s unsettled settlement-obligation units.
The three-way reading is a report projection, not part of the invariant:
CLEAN iff $r \approx 0$ and the in-flight explanation $F = 0$; EXPECTED-LEAD-LAG
iff $r \approx 0$ and $F \neq 0$, each line carrying the obligation-liveness status
of its naming units; BREAK iff $r \neq 0$, identified by (account, coordinate,
residual sign), recomputed from the log, never stored, repaired forward only.
\end{invariant}

\begin{invariant}[The Spine: the as-of read-set]\label{inv:spine}
Everything Invariant~\ref{inv:identity} reads --- balances, in-flight legs, the
account map $W$, the tolerance schedule, the signing key --- folds in force at the
statement's as-of time. Registry facts, tolerances, and keys are recorded declared
data with valid time, amended forward only. A past classification is never
rewritten by a later amendment; a revoked key re-quarantines exactly the statements
admitted in $[t_c, \text{discovery}]$ under it.
\end{invariant}

\begin{invariant}[The Binding: one-parameter construction]\label{inv:binding}
Every transaction that creates or changes a settlement-obligation unit's legs
derives the legs and the balance moves from one restated parameter set, so a unit
whose legs disagree with its booking cannot be constructed and
$\Delta r = 0$ is a theorem of the transaction shape. Successors net recorded
settled mass per coordinate, at the amounts actually moved. The forward-or-return
decision is taken independently per coordinate: forward leg of
$\mathrm{restated}-\mathrm{settled}$ where restated $\geq$ settled; return leg of
$\mathrm{settled}-\mathrm{restated}$, opposite direction, where it is not. All legs
are positive. Legs partition into units by external instruction --- each unit's
legs belong to exactly one root reference (a function, not a bijection: one
reference carries several units only through the certified partial split, whose
inheritance the penalty fold depends on). Whether a mixed correction settles
netted or gross is a declared term of the agreement, and both conserve
identically. In numbers (price corrected to \$10.00 over 30 settled at \$50.00):
netted, receive 70 free of payment plus a \$500.00 cash return; gross,
delivery-versus-payment 70 against \$700.00 plus a \$1{,}200.00 compensation ---
either way predicted nostro is $999{,}000 + \mathrm{out} - \mathrm{in} =
998{,}500.00$ mid-window and total cash paid ends at the restated
\$1{,}000.00.
\end{invariant}

\begin{invariant}[Terminality: settled mass is terminal]\label{inv:terminality}
Recorded settled mass is never retired, never negated, never raced. Successors net
it; corrections below it unwind it forward through return obligations; a
settlement confirmation racing a pending cancellation resolves in settlement's
favour, and the pending repair re-routes to a post-settlement compensating
transaction --- never a hung amendment.
\end{invariant}

\subsection{Temporal mapping}

The workflow is the settlement-obligation unit's own unit-workflow, started by
signal-with-start on the recorded instruction; the record is primary, and a wiped
substrate re-derives every workflow, timer, and in-flight quantity from the log.
Activities are the four side-effecting steps: capture a witness through the door
(before any signal exists), evaluate the unit's contract, propose to the door,
emit the instruction. The timer firing at $t_d$ is the due-date watch's liveness
half and carries no data: it wakes the unit; the recorded quantities and the
contract decide. A witness signal carries only the log position of an
already-recorded fact --- a lost signal costs latency, never the record. Egress
runs intent, send, receipt capture; emission idempotence rides (unit, node), never
a substrate identifier. A split completes the parent workflow and starts one per
successor; an atomic-finality unit arms no timer at all.

\section{Worked Example}

Buy 100 XYZ at \$50.00; $T$ Monday, $t_d$ Wednesday (T+2); closes \$50.00 /
\$52.00 / \$51.50. Pre-trade: $\mathrm{owned}(\mathrm{USD}) = 1{,}000{,}000.00$,
$\mathrm{owned}(\mathrm{XYZ}) = 0$; statements: nostro $1{,}000{,}000.00$, depot
$0$; both accounts CLEAN ($r=0$, $F=0$); $V = 1{,}000{,}000.00$.

\textbf{Monday, $T$.} One transaction: move owned(XYZ) 100, seller $\to$ $w$; move
owned(USD) $5{,}000.00$, $w \to$ seller; the unit SO-1 minted at
\texttt{instructed} (deliver 100 in, pay $5{,}000.00$ out, due Wednesday).
Conservation: $\Sigma_{\mathrm{XYZ}} = +100 - 100 = 0$;
$\Sigma_{\mathrm{USD}} = -5{,}000 + 5{,}000 = 0$. Balances: owned XYZ $100$, USD
$995{,}000.00$. $V_T = 995{,}000 + 100 \times 50.00 = 1{,}000{,}000.00$: a fair
exchange, zero profit and loss at the trade instant. Identity --- nostro:
$995{,}000 + 5{,}000 - 0 = 1{,}000{,}000 = $ statement, $r=0$, $F \neq 0$:
EXPECTED-LEAD-LAG, named by SO-1. Depot: $100 + 0 - 100 = 0 = $ statement:
EXPECTED-LEAD-LAG.

\textbf{Tuesday close, \$52.00.} No moves; conservation vacuous.
$V_{T+1} = 995{,}000 + 100 \times 52.00 = 1{,}000{,}200.00$: \emph{profit and loss
$+\$200.00$, unrealised, with zero cash movement and zero custody movement.} The
custodian still reports $1{,}000{,}000.00$ and $0$; the identity still balances
through SO-1's in-flight legs. The \$200 is real, the custody silence is real, and
the record carries both without contradiction.

\textbf{Wednesday, finality.} sese.025 and camt.054 cross the door as moveless
observations; SO-1 walks \texttt{instructed} $\to$ \texttt{settled} --- zero moves,
conservation vacuous, owned untouched. Statements: depot $100$, nostro
$995{,}000.00$. Identity: $995{,}000 + 0 - 0 = 995{,}000$ and $100 + 0 - 0 = 100$:
CLEAN, CLEAN. $V_{T+2} = 995{,}000 + 100 \times 51.50 = 1{,}000{,}150.00$;
settlement contributed zero to $V$ at every step.

\textbf{Fail and recover (buy-in).} Monday and Tuesday as above; the $+\$200.00$
stands. Wednesday: no confirmation; the due-date watch fires; cumulative confirmed
$0 < 100$; SO-1 walks to \texttt{failed} on the full 100. Owned untouched,
statements unchanged, legs still in-flight under the unsettled unit: both accounts
read $r = 0$, $F \neq 0$ --- EXPECTED-LEAD-LAG, its naming unit printed as ``SO-1
failed, overdue 1~bd'', never a false BREAK. Thursday a buy-in executes at the
\$51.50 mark. Its delivery confirmation discharges SO-1 (settled-by-buy-in): no
second long; owned XYZ stays 100. One compensating cash transaction, three paired
moves: $5{,}000.00$ seller $\to w$ (the unearned trade-date payment returns);
$150.00$ seller $\to w$ (the differential, borne by the failing party);
$5{,}150.00$ $w \to$ buy-in counterparty. $\Sigma_{\mathrm{USD}} = +5{,}000 + 150
- 5{,}150 = 0$ on $w$: owned(USD) remains $995{,}000.00$ --- the buyer paid
exactly \$5{,}000.00 for its 100 shares; the \$150.00 landed on the failing
seller. Once the buy-in settles: depot 100, nostro $995{,}000.00$, CLEAN, CLEAN;
the buy-in date closes the penalty window.

\section{What Was Considered and Rejected}

\begin{itemize}
\item \emph{Debit the nostro at $T$.} The statement breaks by \$5{,}000 all
window; Wednesday's real debit has nothing to discharge.
\item \emph{Recognise nothing until $t_d$.} Tuesday's \$200 does not exist;
trade-date accounting, risk, and capital misstated.
\item \emph{Mirror wallets} (the v11 contras and depot/nostro mirrors): our own
copy of the custodian's book --- a second store that itself needs reconciling.
The unit's legs and the recorded statement replace both.
\item \emph{Stored obligation rows, break registers, materialised statuses.} All
computable from the record; storing them re-admits divergence.
\item \emph{Multi-source quorum on statements.} A statement is not a price; a
quorum over one custodian's own account is theatre.
\item \emph{An in-flight coordinate on PositionState.} A projection stored twice.
\item \emph{A recursion depth cap on partial splits} (adopted Round~1, deleted
Round~2: exactly one residual is ever live, and a cap drops witnessed
confirmations). Replaced by the venue minimum fill size plus aggregation.
\item \emph{An ``expected difference'' band on penalty reconciliation} (adopted
Round~4, retired Round~5): an ungoverned band masks a real overcharge. Replaced by
three decidable checks.
\item \emph{An identity-excluded payable coordinate:} an identity with an
exclusion list is an identity with a memorised hole.
\item \emph{A sixth trust assumption per message family:} a new named assumption
is earned by a different proposition, never a different publisher.
\end{itemize}

\section{Declaration}

\textbf{Rounds used: 10.} Seven named reviewers (correctness-architect, minsky,
jane-street-cto, nazarov, sbl-specialist, regulatory-reporter, rosetta-cdm); every
break answered with a mechanism or a counterexample; four interim constructions
falsified by review and corrected in print (the depth cap, the parallel penalty
windows, the single-sided write-off, the bundled mixed-case successor). Round 10:
seven approvals, each with an attack record.

\textbf{DS1--DS19 disposition} (the v11 acceptance checklist; none silent):

{\footnotesize\renewcommand{\arraystretch}{0.88}
\setlength{\LTpre}{4pt}\setlength{\LTpost}{4pt}
\begin{longtable}{@{}llp{8.6cm}@{}}
\toprule
DS & Invariant & Disposition \\
\midrule
DS1 & Economic exposure at $T$ & Satisfied: owned re-books at instruction; no settlement event moves it. \\
DS2 & Conservation & Satisfied: per (unit, coordinate) at the one door; no carve-out. \\
DS3 & Reconciliation identity & Satisfied: Invariant~\ref{inv:identity}, closed by this design (gap NS-01). \\
DS4 & No discharge without witness & Satisfied: node walks only on recorded triggers; no clip is ever dropped. \\
DS5 & Replay determinism & Satisfied: Spine + tip-only supersession; same log, same classification. \\
DS6 & Idempotent finality messages & Satisfied: cause-derived identity at the door. \\
DS7 & Failure non-reversal & Satisfied: failure is a recorded event; unwind is compensating, forward. \\
DS8 & Status monotonicity & Satisfied: property of the unit's certified product graph. \\
DS9 & Buy-in compensation closure & Satisfied: trade plane by the binding; possession plane by close-out routing. \\
DS10 & Cross-currency Herstatt & Subsumed: routed to the NS-02 currency workstream (interface below). \\
DS11a & Partial per-step conservation & Satisfied: certified split; $q_s + q_r = Q$; parent retires from zero. \\
DS11b & Partial monotonicity, bound & Satisfied: positive inherited fill-size floor + aggregation (closes NS-04). \\
DS12 & Variant degeneration T+0..T+5 & Satisfied: the cycle is a declared term; atomic finality births the unit settled. \\
DS13 & --- & Not defined in the source checklist (v11 numbering gap); nothing to disposition. \\
DS14 & Penalty schema determinism & Satisfied: both CSDR penalties, partitioned windows, decidable checks, booked cash leg. \\
DS15 & Default close-out & Satisfied: certified close-out; the fails cascade routes into it. \\
DS16 & Bitemporal restatement & Satisfied: longer log, never an edit; supersession tip-only. \\
DS17 & Capability-scoped status writes & Satisfied: one writer, the unit's own contract. \\
DS18 & DvP atomicity & Satisfied at ledger level, stated honestly: CSD atomicity reconciled, never guaranteed. \\
DS19 & Witness-identity determinism & Satisfied: cause-derived identifiers; origin authenticity on statements. \\
\bottomrule
\end{longtable}}

\textbf{Interfaces to queued problems.} DS10's cross-currency legs and Herstatt
projection go to the NS-02 currency workstream (an FX leg is two
settlement-obligation units whose joint open state projects as exposure); the
close-out claim's accrued rebate/fee/manufactured-payment income leg goes to the
billing/income workstream (the claim is never the complete close-out amount); the
executable-property set --- identity, Binding constructibility, window partition,
race interleavings, each with a required firing witness --- goes to the test
regime.

\textbf{Decision log.} DL-01 (panel, 3--0): the custodian statement crosses the
one door as a registered witness event kind, and TA-CUSTODY is named ---
proposition: a recorded custody assertion faithfully renders its asserter's book;
owner: data governance; violation: misrender, forgery, replay, silence; detection:
Invariant~\ref{inv:identity} classifying BREAK, signature and sequence
verification, the cadence watch; residual: the coincidental misrender, the
send-to-CSD-entry gap, the matched-date assertion inside a genuinely late window.
DL-02 (unanimous, 7--0): penalty advices and emission receipts ride under
TA-CUSTODY as enumerated asserter classes --- a new named assumption is earned by
a different proposition, never a different publisher.

\textbf{Deferred to the owner (spec-pass proposals, not constitutional
amendments).} Naming TA-CUSTODY as the fifth trust assumption edits one sentence
of the v16.0 specification, Chapter~16: ``The boundary rests on four named trust
assumptions and no other'' becomes ``\dots five \dots: TA-KIND, TA-TERMS,
TA-EXDATE, TA-ARRIVAL, TA-CUSTODY,'' with the four-part clause above appended.
This is precisely a specification amendment: the constitution (C-4.12) names the
trust-assumption pattern and fixes no count; only the specification sentence
enumerates four. The venue cut-off's faithfulness rides the existing TA-TERMS.
Both edits are parked with exact text for the owner's spec pass.

\textbf{Confidence.} HIGH on the design: ten adversarial rounds, seven reviewers,
every mechanism individually attacked, all residual trust honestly named. The one
open obligation is execution: the property set is specified with required firing
witnesses but has not yet run, and a property that has never fired witnesses
nothing. Confidence becomes final when the witnesses fire.

\end{document}
