% =====================================================================
%  THE LEDGER FRAMEWORK — Manifesto, Version 1.1
%  R. Delloye — July 2026
%  v1.1: Amendments 1–3 (collateral ruling), ratified by the owner
%  2026-07-11 per the Phase 1 Design Ruling memo.
%  Build: pdflatex (twice). Requires a standard TeX Live; lmodern and
%  microtype are used when available and skipped otherwise.
% =====================================================================
\documentclass[11pt,a4paper]{article}

\IfFileExists{lmodern.sty}{%
  \usepackage[T1]{fontenc}\usepackage{lmodern}%
  \IfFileExists{microtype.sty}{\usepackage{microtype}}{}%
}{}
\usepackage[margin=2.5cm]{geometry}
\usepackage{parskip}
\usepackage[hidelinks]{hyperref}

\setcounter{secnumdepth}{0}          % section numbers are carried in the titles
\makeatletter\def\verbatim@font{\normalfont\small\ttfamily}\makeatother
\providecommand{\tightlist}{\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}}

\begin{document}

\begin{center}
  {\LARGE\bfseries THE LEDGER FRAMEWORK}\\[0.9em]
  {\large\bfseries Manifesto and Architectural Constitution}\\[0.6em]
  {\normalsize A Unified, Closed-System, Event-Driven Post-Trade Architecture}\\[1.4em]
  R.~Delloye \quad---\quad July 2026 \quad---\quad Version 1.1
\end{center}

\vspace{0.6em}
\begin{center}\rule{0.35\linewidth}{0.4pt}\end{center}
\vspace{0.6em}

\begin{center}\textbf{Abstract}\end{center}
\begin{quote}
Reconciliation exists because many systems store the same facts, and
independently maintained copies eventually disagree. The Ledger removes
the cause: every fact is recorded exactly once, in an immutable log of
atomic transactions, and everything else --- balances, valuations,
profit and loss, reports --- is a deterministic projection of that
single record, reproducible by any party and reconstructible at any past
date; illegal states are not checked for but made unrepresentable. This
manifesto is the project's founding document: assuming no prior
knowledge, it states the objective, the six commitments that govern
every design decision, the objects the system manipulates, the arrows
between them, and the machines that evaluate those arrows --- the
architecture, not the engineering, with implementation questions named
and left to the specification. It closes with the minimum requirements
for the Event Monitor, the Events Executor, and the pricing layers, and
with the rule that governs everything after: all future specifications
and implementations are rated against this document, never the other way
round.
\end{quote}

\vspace{0.6em}
\begin{center}\rule{0.35\linewidth}{0.4pt}\end{center}
\vspace{0.6em}

\hypertarget{the-objective-and-the-problem}{%
\section{1. The Objective and the
Problem}\label{the-objective-and-the-problem}}

The objective of the Ledger project is to build a clean post-trade and
risk-management platform: one system of record from which positions,
valuations, profit and loss, collateral, and reports are all derived ---
exactly, deterministically, and reproducibly.

The obstacle is the architecture of today's infrastructure. Post-trade
processing is not one system but many --- trading systems, risk engines,
general ledgers, custodians, regulatory engines --- and each holds its
own copy of positions, cash balances, corporate-action adjustments, and
lifecycle states, updated on its own schedule by its own logic. The
copies diverge, and reconciliation --- detecting and repairing the
divergences --- is one of the largest operational costs and risks in the
industry. The root cause is architectural: \textbf{the same fact is
stored more than once.} Two stores of one fact maintained by independent
logic will eventually disagree, and no amount of process removes a
failure mode the architecture itself provides.

The Ledger removes it at the root. There is exactly one canonical record
--- an ordered, append-only stream of atomic transactions --- and every
other number is computed on demand as a projection of that stream:
balances, valuations, profit-and-loss statements, balance sheets,
regulatory reports. Two projections of one record cannot disagree about
any fact the record contains. That sentence is the design's entire
ambition; the rest of this document is what it forces.

\hypertarget{the-six-commitments}{%
\section{2. The Six Commitments}\label{the-six-commitments}}

Six commitments are non-negotiable, and every design decision in this
project is tested against them.

\textbf{Full auditability.} Every number is traceable, mechanically and
completely, to the recorded events that produced it. History is never
rewritten: a mistake is repaired by a recorded correction that names
what it supersedes, so the error and its repair both survive.

\textbf{Full reproducibility.} Given the same record, any party --- the
firm, a counterparty, an auditor, a regulator --- recomputes the same
state, the same views, the same contract outputs, to the last minor
unit.

\textbf{Time travel embedded in the design.} The exact state of the
system at any past moment is reconstructible from the record alone, at
any time, forever --- independently of whether the software that
produced the events still exists.

\textbf{Correctness.} Illegal states are made unrepresentable wherever
possible: a check can be forgotten or bypassed; a type cannot. Where
wrongness cannot be prevented --- a contract can always compute the
wrong economics --- it is made mechanically detectable and
deterministically repairable. Defaults fail closed: the system stops
rather than improvises.

\textbf{Testability.} The ledger is correct because it is built to be
tested. Every step is a pure, versioned function with explicit inputs,
explicit outputs, and strong types --- typing alone eliminates whole
classes of defects before a single test runs --- and every core
invariant is expressed as an executable check over generated products,
events, and histories, never defended only in prose. Correctness then
composes: because the steps are pure and their composition is fixed ---
the map, then the fold --- verifying each part verifies the whole. A
property that cannot be tested mechanically is redesigned until it can
be. This is not an afterthought: the architecture is deliberately shaped
so that its core guarantees are executable --- property-based testing is
a construction principle, not a phase at the end.

\textbf{Clarity.} One name per component and one component per name;
behaviour carried as declared data rather than hidden in code; every
claim stated so that it can be checked by a reader with no prior
exposure.

The document builds in a fixed order: the whole architecture in one
picture (Section 3); the objects it manipulates (Section 4); the
machines that run it (Section 5); the smart contracts that animate it
(Section 6); and then the consequences, domain by domain, ending with
minimum requirements and scope.

\hypertarget{the-whole-architecture-in-one-picture}{%
\section{3. The Whole Architecture in One
Picture}\label{the-whole-architecture-in-one-picture}}

Everything this framework does can be stated as a map followed by a fold
--- two words borrowed from programming, worth a line each. In plain
words: transform each event, then accumulate the results. A \emph{map}
processes a list element by element: given a way to turn an a into a b,
it carries {[}a{]} to {[}b{]}, one element at a time. A \emph{fold}
accumulates a list into a single result: each element of {[}a{]} is
processed in turn and absorbed into a running value, {[}a{]} to b. Four
concepts complete the picture, each defined precisely in Section 4: a
\textbf{unit} is anything that can be held or owed --- a stock, a bond,
an option, a currency; a \textbf{wallet} is where units are held, in
signed quantities called balances; a \textbf{transaction} is the atomic
bundle of changes the ledger records; and a \textbf{projection} is any
view computed from the record. Everything else this section names, it
explains as it goes.

The world presents the system with a list of events, E = {[}e{]}. Some
arrive from outside: trades executed in the market, corporate-action
notices, fixings and other observations, captured and recorded at the
boundary. Others arise from the instruments themselves --- a coupon date
arriving, an expiry falling due, a close printing below a declared
barrier --- and these are not received but noticed: every unit declares,
from its terms, the conditions its life requires watching, and the Event
Monitor evaluates those watches against the clock and the recorded
observations, emitting an event onto the record the moment a condition
is met.

Each event is mapped into a transaction --- itself a bundle of moves,
state changes, and terms changes, defined precisely in the next section
--- and the responsibility for the map divides in two. The smart
contracts are the mapping functions: each is a stateless piece of code
that, given one event and the current state of the ledger, computes the
transaction the event requires. The Events Executor applies the map over
the list, E = {[}e{]} → T = {[}tx{]}: for each event it fires the
responsible contract at the right time, hands it the event together with
the current projection of the ledger, and collects the proposed
transaction. Many contracts, one Events Executor; the contracts carry
all the product knowledge, the Events Executor carries none.

The resulting list of transactions is then folded --- reduced, one
transaction at a time --- into the current state of the system:
balances, unit states, and product terms. The Transaction Executor
performs the fold, T = {[}tx{]} → Ledger: it serialises, validates, and
applies each transaction, and the immutable log is precisely the list
being folded. Every view --- a balance, a valuation, a report --- is a
further fold of the same list.

Three kinds of thing appear in this picture, and they must never be
confused: data, functions, and machines. The data are Event,
Transaction, and Ledger. The functions are the arrows between the data.
The machines are what evaluate the arrows --- one machine per arrow.
Typed:

\begin{verbatim}
watch    : (Record, Clock)       -> [Event]      -- the Event Monitor
contract : (Event, Ledger)       -> Transaction  -- the Events Executor
apply    : (Transaction, Ledger) -> Ledger       -- the Transaction Executor

step (ledger, e) = apply (contract (e, ledger), ledger)  -- map, then fold
final ledger     = fold step over E = [e]
\end{verbatim}

One instrument can walk the whole pipeline. Take a one-touch option that
pays 1 USD if the spot of its underlying ever trades at or below a fixed
barrier. At registration, its terms declare the watch: the underlying's
recorded prices against the barrier. Then, the day a low print is
recorded:

\begin{itemize}
\tightlist
\item
  \textbf{watch} --- the condition is met: the Event Monitor emits the
  touch event onto the record.
\item
  \textbf{contract} --- the Events Executor fires the option's contract,
  handing it the event and the current projection of the ledger; the
  contract reads the terms and who holds the option, and returns one
  transaction: a single move of 1 USD from the issuer's wallet to the
  holder's wallet, plus one state change on the unit ---
  \emph{triggered}.
\item
  \textbf{apply} --- the Transaction Executor admits it, and the fold
  advances: the ledger now shows the cash with the holder and the option
  triggered.
\end{itemize}

Should the Monitor ever emit the touch again, the next firing finds the
unit already triggered and proposes nothing --- the recorded state is
exactly what makes a duplicate harmless.

Time deserves separate attention. The clock is the single input that is
not on the record, and the Event Monitor is the single place it is
consulted. The event the Monitor emits is recorded, so everything
downstream of it is a function of the record alone --- which is exactly
why a late firing produces the identical transaction. Because watches
are declared and the Monitor is deterministic, even the emission of
events is verifiable: given the record and the clock, anyone can
recompute which events should have been emitted, and when.

Both contract and apply can refuse --- a contract may find nothing due;
apply rejects a malformed transaction and writes nothing --- and a
refusal is a returned value, never a half-applied state.

Two remarks complete the picture. First, the map is not blind to state:
to turn a dividend event into cash moves, the contract must read who
owns what at that moment --- the fold of everything admitted so far. Map
and fold therefore advance together, one event at a time, in a single
total order (Section 5). Second, the map is proposed while the fold is
authoritative: a transaction joins the list only when the Transaction
Executor admits it, and economic verification (Section 13) is simply
re-running the map on the recorded event and comparing --- which is why
both the events and the transactions are on the record.

A note for readers who know Event Sourcing, because the resemblance is
real and the difference matters. The Ledger is event-driven because
finance is: trades execute, prices print, coupons mature, and barriers
are touched whether or not the software cares. Events are recorded
because reality presents them --- not because event sourcing was chosen
as a pattern. The architectural difference is where authority sits. In
event sourcing, the event stream is the canonical history, state is
whatever the current code derives from it, and reinterpreting history
through new code is a feature. Here, events are inputs to computation,
never the accounting record: contracts transform them into proposed
transactions, the Transaction Executor admits or refuses --- an
admission door event sourcing does not have --- and the admitted
transaction log alone is the canonical history. Replay re-reads
transactions and never re-executes contracts, so history cannot be
silently reinterpreted; a wrong interpretation is repaired by a
compensating transaction, in the open. The events stay on the record for
one purpose: so the map can be re-run and verified. Verification lives
with the events; authority lives with the transactions.

\hypertarget{the-objects}{%
\section{4. The Objects}\label{the-objects}}

At the highest level, a ledger is three things and a record: units ---
what can be held; wallets --- where it is held; balances --- how much of
each unit each wallet holds; and the immutable log that records every
change to them. This section defines each object once; nothing later in
the document introduces another.

A \textbf{unit} is anything that can be held or owed inside the ledger:
a currency (USD in cents), an equity security (identified by its ISIN),
a bond, a listed or OTC derivative, a securities loan, a managed-account
mandate, a margin obligation under a collateral agreement (a CSA), or
any other contractual instrument. Units are first-class and extensible:
adding a new instrument type adds a new element to the universe of
units, never a modification of the machinery. Because every unit ---
asset, liability, or contractual obligation --- is represented
uniformly, the same move machinery and the same conservation discipline
apply without special cases.

Units divide into valued and non-valued. A unit is \textbf{valued} if
and only if it represents an economic claim whose value exists
independently of the governance or continued operation of the instrument
that created it. A stock, a bond, and a client's redemption claim on a
managed account are valued. A mandate's rulebook, a portfolio-swap reset
mechanism, or a quantitative investment strategy's governance shell is
non-valued: its price is undefined --- not zero --- because pricing the
governance structure beside the claim it governs would count the same
economic reality twice.

Every unit carries three things: its terms (the contractual text,
versioned and append-only), its state (where it stands in its
lifecycle), and its smart contracts --- the executable form of those
terms. From its terms follow its \textbf{watches}: the dates,
observations, and conditions the Event Monitor must observe on its
behalf. Declaring them is the unit's duty --- at registration, and again
whenever a later transaction creates a new obligation --- and the
declaration is exhaustive: the Monitor does exactly what the units have
asked of it, nothing more and nothing less.

A \textbf{wallet} is a named logical partition of position space --- a
container, and deliberately nothing more. Its state at any moment is a
signed quantity, its balance, for each unit it holds. A wallet names two
parties: a \textbf{manager}, who is authorised to perform transactions
on it, and a \textbf{beneficial owner}, who is legally accountable for
the book. Wallets carry no smart contract and no logic of their own;
where a wallet appears to need behaviour --- the rules of a managed
account, for instance --- that behaviour is a unit (the mandate).

An \textbf{atomic move} is the sole operation that changes a balance: it
transfers a positive quantity of exactly one unit from a source wallet
to a destination wallet in a single indivisible step. Quantities are
exact integers in minor units; where an event divides unevenly, the
rounding convention is part of the declared terms and the remainder is
booked explicitly as its own move --- nothing is ever silently dropped.
Because a move only transfers, it can never create or destroy quantity.

A \textbf{transaction} is an atomic, all-or-nothing bundle of four kinds
of change: moves, unit-state changes, position-state changes, and terms
changes. The four align one-for-one with the four components of the
ledger's state: balances, and the three homes of Section 7. A
transaction applies in full or not at all. Some transactions carry no
moves (for example, the registration of a new unit or the recording of a
barrier breach).

The \textbf{immutable log} is the ordered, append-only, hash-chained
record of every admitted transaction --- hash-chained so that any
alteration of history is detectable. It is the sole canonical source of
truth. Three words are used precisely throughout this document: the
\textbf{log} is the recorded sequence itself; the \textbf{ledger} is the
state obtained by folding the log; and the \textbf{record} is everything
that has been recorded --- the log and all it carries, transactions,
events, and observations alike. ``On the record'' means present there,
and a projection is any view computed from it --- which is also why, in
the typed picture of Section 3, the watch reads the Record while the
contract and apply consume the Ledger.

Every counterparty outside the system is represented by a virtual wallet
inside it, so the system is closed. Conservation holds by construction,
not by check.

When a balance must carry more information than a single number, it
generalises to a signed vector of coordinates --- owned, lent, posted ---
whose named rays are: lent out and borrowed, the two signs of lent;
posted as collateral and received as collateral, the two signs of
posted. A quantity earns a
coordinate only when a distinct real-world action can change that
coordinate independently of ownership. Only the owned coordinate carries
economic value and drives profit and loss. The remaining coordinates
carry mass, never value: they record possession and encumbrance under a
named agreement, no quantity reaches them without reference to the
agreement unit that governs it, and a lifecycle event extinguishes
value, never mass --- a unit leaves the ledger only from the zero
vector. Anything computable from the
coordinates is a projection, computed when needed and never stored.

\hypertarget{the-machines}{%
\section{5. The Machines}\label{the-machines}}

The framework has three machines --- one for each arrow of the picture
in Section 3. All three are generic machinery carrying no product
knowledge, and only the last ever writes the ledger: three roles, one
writer. In one picture --- machines in brackets, data on the arrows, the
dotted line the trust boundary:

\begin{verbatim}
 ................... outside the trust boundary ...................
 :                                                                :
 :      clock                          external events            :
 :        |                                  |                    :
 :        v                                  v                    :
 :   [Event Monitor] -- emits Event --> [Events Executor]         :
 :                                           |  fires             :
 :                                           v                    :
 :                                    [Smart Contract]            :
 :                                           |  proposes          :
 :                                           v                    :
 :                                      Transaction               :
 :...........................................|....................:
                                             v
               [Transaction Executor] -- admits --> Immutable Log
                    (the one writer)                     |
                                                         v
                                                    projections
\end{verbatim}

The \textbf{Event Monitor} owns time and attention. It holds every
declared watch and evaluates them against the clock and the recorded
observations, emitting an event onto the record the moment a condition
is met. It holds no private list --- the watches it evaluates are on the
record --- and it emits events, never transactions. The relationship
between the units and the Monitor is a recorded handshake. Each declared
watch is acknowledged by the Monitor, and the acknowledgement --- an
event like any other, flowing through the same pipeline --- is written
into the unit's state; a watch is thus visibly \emph{declared}, then
\emph{armed}, then \emph{fired} or \emph{expired}.

A declared watch that is never armed is therefore a visible gap, never a
silent one. The full chain is on the record and can be walked in either
direction: register a unit → the conditions it declares → the
acknowledgements → the events → the transactions → the state updates.
The Monitor can affect timeliness and liveness, but it cannot affect the
correctness of ledger state, because it only emits events --- it never
proposes transactions.

The \textbf{Events Executor} owns delivery. It captures and records the
events arriving from outside, and, for each event --- arrived or emitted
--- it fires the responsible smart contract, handing it the event
together with the current projection of the ledger, and collects the
proposed transaction. It runs the contracts; it never writes the ledger;
its output is only proposals.

The \textbf{Transaction Executor} owns the record. It is the gatekeeper
of the ledger door: the single component through which every transaction
must pass, and the only component permitted to change the ledger's
state. It validates structure --- conservation is automatic;
authorisation, idempotence, consistency of reference, and writer
discipline are enforced --- and, if the transaction is admissible,
appends it to the immutable log. Every proposed transaction carries an
identifier derived from its cause, so a retried or duplicated proposal
is recognised and committed only once. It creates nothing and runs
nothing; it admits or refuses. There is no other way to change the
ledger's state.

The one-touch option of Section 3 met all three machines in order: the
Monitor noticed the low print and emitted the touch; the Events Executor
fired the option's contract with the event and the projection; the
Transaction Executor admitted the single move and the \emph{triggered}
state. One event, three machines, one write.

The single-writer discipline is a correctness requirement. The
Transaction Executor therefore presents to the log a strictly monotonic
sequence of committed transactions; where two simultaneous events on the
same instrument do not commute and declare no precedence, it refuses
rather than guesses an order.

The roles sit at different levels of trust. The Transaction Executor is
the trusted component: the ledger's integrity rests on it and on nothing
else. The contracts, the Event Monitor, and the Events Executor are
outside the trust boundary, untrusted in different ways: a contract can
be economically wrong, so its output is verified by recomputation
(Section 13); the Monitor and the Events Executor cannot corrupt the
ledger at all --- a late emission or firing produces the identical
transaction, a duplicate is committed once, a spurious one finds nothing
due and is refused --- but the timeliness of the system depends on them.
Integrity owes them nothing; liveness is conditional on them. Section 14
states their minimum requirements.

\hypertarget{smart-contracts-event-driven-pure-and-faithful-to-the-legal-contract}{%
\section{6. Smart Contracts: Event-Driven, Pure, and Faithful to the
Legal
Contract}\label{smart-contracts-event-driven-pure-and-faithful-to-the-legal-contract}}

Financial instruments are not passive data records; they are active
contracts that, when triggered by an event, emit the transactions
reflecting the economic consequences of that event. The Ledger performs
none of this product-specific logic: it admits well-formed transactions
and records them. The logic that decides what a dividend, an expiry, a
barrier knock, or a corporate action means for a particular instrument
resides in the instrument's smart contract, outside the ledger's trust
boundary. Since the ledger is only a transaction processor, anything
coming from outside must be converted into a transaction, and the smart
contract is the converter.

Examples illustrate the pattern:

\begin{itemize}
\item
  A dividend is announced by the issuer, and the announcement is
  recorded when received. The announcement fires the stock's smart
  contract once --- not to pay, but to register the watches the dividend
  creates: its record date and its payment date. When the record date
  arrives, the Event Monitor emits the event and the Events Executor
  fires the contract again; it reads the positions as the ledger then
  stands and records each holder's entitlement --- a moveless
  transaction: no cash changes hands, but the position state of every
  holder is updated with the entitlement it is owed. When the payment
  date arrives, the Monitor emits again and the contract fires a third
  time, emitting the cash moves against the recorded entitlements.
  \textbf{Three firings, three transactions, one stateless contract.}
\item
  A cash-settled future declares, from its terms, the watches its life
  requires: each day's settlement-price publication and its expiry date.
  On each recorded settlement print the contract fires and emits the
  day's variation-margin moves, updating in the same transaction the
  position state that records the margin applied and the settlement
  level used; on expiry the Monitor emits the final event, and the
  contract unwinds the position and pays the last cash settlement ---
  one atomic transaction.
\item
  A listed option's smart contract is triggered at expiry by the
  official settlement price; it evaluates whether the option is
  in-the-money and emits the transaction that extinguishes the position
  and delivers the underlying or the cash amount, in one atomic step.
\item
  A note's barrier is a declared watch on the underlying's official
  close. The close is observed and recorded; the Event Monitor emits the
  knock event, and the note's contract fires and records the breach ---
  a lifecycle state change that may involve no moves at all.
\end{itemize}

Two principles govern all of these. First, everything a smart contract
consumes must itself be on the record. Second, a contract never waits
and never remembers: the watch waits in the record, the Event Monitor
notices when its condition is met, the Events Executor delivers the
firing --- and whatever a contract must carry from one firing to the
next, it records.

Purity and statelessness follow from the same demand. Every smart
contract is a pure, deterministic function of its inputs: the declared
terms of the instrument, the recorded observations, and the visible unit
and position state at the moment of invocation. Each contract is
therefore a stateless, machine-readable encoding of the legal terms and
conditions of the instrument it governs.

The alignment with the legal contract runs deeper: every right and
obligation the legal contract creates is represented in the ledger as a
unit. A subscription is booked as an exchange --- cash against the
client's redemption claim --- never as a transfer for nothing. Treating
the client's redemption claim as a valued unit has direct consequences
for fee accrual and NAV attribution in managed accounts; these
consequences are resolved in the detailed specification consistently
with the principles stated here.

\hypertarget{state-the-three-homes}{%
\section{7. State: The Three Homes}\label{state-the-three-homes}}

Every unit progresses through a well-defined sequence of lifecycle
states. Some elements of state are properties of the unit alone; others
are joint properties of the unit and a specific wallet.

The three-home state model encodes the distinction: (i) ProductTerms ---
the immutable, append-only versions of an instrument's contractual
terms; (ii) UnitStatus --- a materialised, replayable projection of
events that affect the unit as a whole; and (iii) PositionState ---
per-wallet, per-unit attributes. Because all three are derived from the
immutable log rather than being independent mutable stores, they can be
discarded and rebuilt by replay at any time without loss of information.
Each non-balance fact has exactly one legal writer.

The three homes are not bookkeeping conveniences: together with the
balances, they must carry \textbf{all the information the smart
contracts require in order to execute} --- the past fixings a variance
swap has accrued, the last margin call a future has applied, the coupons
a bond has already paid. Each such fact is written into its home by an
earlier firing, so that the next firing finds its entire context on the
record; if a contract would ever need to remember something, that
something has a home.

The daily life of a cash-settled future shows all three homes at work.
The official settlement price, once published and recorded, is the same
fact for every holder: it lives in UnitStatus. The variation margin a
particular holder paid or received yesterday depends on that holder's
position: it lives in PositionState. And when the underlying stock
splits, the contract's strike and multiplier are adjusted by appending a
new terms version: that is a change to ProductTerms. One event --- a
settlement print, a margin payment, a split --- lands in exactly one
home, and the home is determined by what the fact is a property of: the
unit alone, the unit-and-holder pair, or the contract's terms.

\hypertarget{valuation-and-profit-and-loss}{%
\section{8. Valuation and Profit and
Loss}\label{valuation-and-profit-and-loss}}

Net Asset Value has an operational meaning before it has a formula. The
initial NAV of a wallet is the cash it took to build the portfolio at
inception; the current NAV is the cash you would hold if you unwound
every position at prevailing prices; and the profit and loss (PnL) is
the difference between the two.

At any point in time the NAV of a wallet is obtained by (i) replaying
the transaction log to recover the wallet's current composition and (ii)
multiplying that composition by the current price vector supplied by the
pricing data layer, the sum running over the valued units only.

Because unit state affects value, the pricing function is also
parameterised by the current lifecycle state of the instrument ---
obtained from the ledger, never from a store of its own.

A central and non-negotiable property follows: the profit and loss of
any wallet is exactly the variation in its NAV. PnL between two dates is
path-independent, and no second copy of ``realised PnL'' or ``book
cost'' is ever maintained alongside the position record.

For the variation in NAV to be exactly profit, deposits and
contributions must not create profit. Every inflow of value is one of
three things: an exchange paired with an equal-valued outflow ---
settled-to-market margin is this case, a settlement extinguishing the
day's exposure with no return obligation; a contribution or financing
received against an equal obligation created in the same transaction ---
collateral received under title transfer, cash included, is this case,
owned by the receiver against an equivalent-return obligation that is
itself a unit; or value held in custody without being owned, recorded on
the collateral-received coordinate --- collateral received under a
security interest without right of use, segregated cash included, is
this case. Where an inflow moves under a margin or collateral agreement,
which case governs it is declared once, on that agreement unit. A
deposit therefore cannot move profit
and loss --- not because a formula subtracts it, but because it cannot
change net owned value.

\hypertarget{corporate-actions-and-the-market-data-operator}{%
\section{9. Corporate Actions and the Market Data
Operator}\label{corporate-actions-and-the-market-data-operator}}

A corporate action is, first of all, a transaction like any other: a
list of moves, together with unit-state changes and a new version of the
instrument's terms, all applied atomically through the same Transaction
Executor.

A corporate action also changes the meaning of all the market data
recorded before it. To carry pre-event market data into the post-event
frame, a transformation must be applied. The framework gives it a name:
the \textbf{market data operator}. Every kind of market data is equipped
with its own operators, one per kind of event --- declared once, when
the kind of data is introduced and the corporate action is recorded,
never improvised at the moment of reading. A split shows the variety:
the spot scales; cash-per-share dividend figures scale while
proportional (percentage) figures stand; index compositions are updated
--- in every case until fresh post-event data are recorded.

Three principles govern the operators. First, the operator is intrinsic
to the pairing of the kind of data and the kind of event. Second, the
ledger is the authority; the pricing data layer applies. Third,
originals are never overwritten. The observed price stays on the record
exactly as observed; the adjusted value is computed at each read.
Derived quantities are recomputed from operator-adjusted inputs.

\hypertarget{strategies-virtual-ledgers-and-total-return-swaps}{%
\section{10. Strategies, Virtual Ledgers, and Total Return
Swaps}\label{strategies-virtual-ledgers-and-total-return-swaps}}

Nothing in this architecture is specific to real holdings. A
\textbf{virtual ledger} is a complete instance of the same machine ---
units, wallets, balances, an immutable log, the three machines --- whose
holdings are notional: it settles nothing, and none of its wallets ever
reaches a settlement instruction or the balance sheet. Virtual ledgers
are subject to the same six commitments as true ledgers, especially full
reproducibility and time travel. The word virtual keeps one meaning: on
the far side of the settlement boundary.

The motivating case is a quantitative investment strategy. A
deterministic strategy is a smart contract whose rulebook is the terms
of a (non-valued) strategy unit. The hypothetical portfolio it manages
is a wallet in a virtual ledger. Each rebalancing is a recorded
transaction triggered by stamped market observations. The strategy's
level is that wallet's NAV --- a projection computed by the same fold as
any other; because the strategy ledger is replayable, any party holding
the recorded rules and observations recomputes the level exactly, and an
index restatement is a compensating entry, never a rewrite.

Two disciplines make this safe. First, the bridge between ledgers is an
observation, never a move. Second, the wall between models and the
ledger holds level by level: the true ledger never runs the strategy; it
consumes the level as an observation.

When the strategy's trades are real, the strategy's output toward the
world is an order (an instruction, not a fact). The executions that come
back are external events, mapped by contracts and folded into a real
wallet of the true ledger. The virtual ledger becomes the benchmark: it
holds what the rules say the portfolio should be, the real wallet holds
what execution achieved, and the difference between the two folds is the
slippage --- exact, never estimated.

A total return swap is then one definition: the reference leg of a TRS
is the NAV of a designated wallet (or virtual ledger). In every case the
TRS is an ordinary unit in the true ledger whose smart contract, at each
reset, reads the stamped NAV observation and emits the reset and
financing moves.

\hypertarget{structural-invariants-and-the-impossibility-of-broken-states}{%
\section{11. Structural Invariants and the Impossibility of Broken
States}\label{structural-invariants-and-the-impossibility-of-broken-states}}

The Ledger is deliberately engineered so that certain classes of
inconsistent or illegal state cannot be represented at all:

\begin{itemize}
\tightlist
\item
  \textbf{Atomicity.} A transaction is applied in full or not at all.
\item
  \textbf{Consistency of reference.} A quantity and a price may be
  combined only when they refer to the same state of the instrument.
  After a two-for-one split, 1,000 shares at 100 become 2,000 at 50;
  multiplying the new count of 2,000 by the stale price of 100 would
  book a phantom valuation.
\item
  \textbf{Writer discipline.} Lifecycle facts are written only by the
  single legal writer for that fact.
\end{itemize}

These three join the invariants established earlier --- conservation by
construction (Section 4), the append-only log (Section 4), the single
writer (Section 5), the determinism of every arrow (Section 3), and
idempotence under the cause-derived identifier (Sections 5 and 12) ---
to form the catalogue that the property-based tests of the Testability
commitment exercise. Together they render the classes of internal
reconciliation failure that dominate conventional architectures
unreachable within the Ledger's scope: what cannot be represented cannot
break.

\hypertarget{time-travel-replayability-and-idempotence}{%
\section{12. Time Travel, Replayability, and
Idempotence}\label{time-travel-replayability-and-idempotence}}

Every historical state of the Ledger is exactly reconstructible by
replaying the immutable transaction log up to the desired timestamp.
Because it reads only committed transactions and never re-executes smart
contracts, the historical record is independent of the machinery that
produced it.

Smart contracts must be idempotent. The mechanism is the cause-derived
identifier of Section 5: a retried or late-arriving proposal for an
already-processed event produces no duplicate effect.

\hypertarget{two-layers-of-correctness}{%
\section{13. Two Layers of
Correctness}\label{two-layers-of-correctness}}

The correctness of the Ledger rests on two distinct layers:

\begin{itemize}
\tightlist
\item
  \textbf{Structural correctness} is guaranteed by the Transaction
  Executor. These properties hold regardless of whether the smart
  contract that produced the transaction was correct, careless, or
  malicious.
\item
  \textbf{Economic correctness} is not guaranteed at admission; it is
  made checkable by recomputation. A smart contract may emit a perfectly
  well-formed transaction that nevertheless books the wrong economics.
  The discrepancy is detected after the fact by recomputing the
  transaction from the recorded inputs. The difference constitutes a
  defect of the contract, not of the ledger, and it is repaired by a
  subsequent compensating transaction through the same door.
\end{itemize}

This separation is fundamental. The Ledger relies on structural
correctness because it is guaranteed by construction; it expects
economic correctness of its contracts and renders it mechanically
verifiable. Detection is after the fact. The framework prefers a
detectable, repairable error over a pretence of prevention it cannot
deliver.

\hypertarget{minimum-requirements}{%
\section{14. Minimum Requirements}\label{minimum-requirements}}

\hypertarget{the-event-monitor-and-the-events-executor}{%
\subsection{14.1 The Event Monitor and the Events
Executor}\label{the-event-monitor-and-the-events-executor}}

The Event Monitor and the Events Executor are infrastructure outside the
ledger's trust boundary. Any implementation must satisfy, at minimum:

\textbf{Watches and timers} - Durable watches and timers. A watch
registered for a future date or condition fires when it is met,
regardless of restarts or failures; timer state is persisted. -
At-least-once emission. No event is ever lost; a duplicate is harmless
because of idempotence checking. - Acknowledged watches. Every declared
watch is acknowledged, and the acknowledgement is recorded in the unit's
state; a watch that remains unacknowledged beyond its arming window is a
visible, overdue item, never silence.

\textbf{Delivery} - Triggers carry timing, not data. A trigger says
when; everything the event means is read from the record. -
Deterministic, replayable orchestration. Each machine's own state is
reconstructible by replaying its append-only history. - No privilege.
Neither machine holds write access to the ledger. - Obligation liveness.
Every contractual obligation is registered with a deadline, a test for
its discharge, and a fallback action, and by its deadline it is
discharged, compensated, or declared defaulted; a duty that fails to
fire is a visible, overdue item, never silence.

\hypertarget{the-pricing-stack-and-the-pricing-data-layer}{%
\subsection{14.2 The Pricing Stack and the Pricing Data
Layer}\label{the-pricing-stack-and-the-pricing-data-layer}}

\begin{itemize}
\tightlist
\item
  The ledger records the outputs of pricing models and never runs one.
\item
  Valuation is state-dependent and obtains that state from the ledger.
\item
  Valuations are reproducible from recorded inputs.
\item
  Prices and quantities are never mixed across an event.
\item
  Originals are preserved; adjusted values are computed; derived values
  are recomputed.
\item
  Estimates and entitlements are kept apart.
\item
  Model outputs re-enter only as new observations, carrying sufficient
  provenance. Reproducing a valuation from recorded inputs and
  identified prices is unconditional; reproducing a model's number
  requires the model itself, whose retention is a governance matter
  outside the ledger's scope.
\end{itemize}

\hypertarget{scope-of-the-ledger-project}{%
\section{Scope of the Ledger
Project}\label{scope-of-the-ledger-project}}

The Ledger project encompasses the design, formal specification,
reference implementation, and verification of the following components:

\begin{itemize}
\tightlist
\item
  The closed double-entry move model, the conservation law, and the
  single-door admission function.
\item
  The immutable, hash-chained transaction log and the projection
  machinery that derives all views from it.
\item
  The three-home state architecture (ProductTerms, UnitStatus,
  PositionState) and the conditions that render illegal states
  unrepresentable.
\item
  The smart-contract framework, its event-triggering discipline, and the
  purity and idempotence requirements.
\item
  The market data operators and the integration layer between lifecycle
  state, pricing models, and the corporate-action adjustment machinery.
\item
  The time-travel and historical-replay machinery, with its guarantees
  of determinism and reproducibility.
\item
  The formal invariant catalogue and the property-based testing regime.
\item
  The settlement-projection layer.
\item
  The virtual-ledger mechanism: subordinate, non-settling instances of
  the same machine whose projections enter the true ledger only as
  stamped observations.
\end{itemize}

Explicitly out of scope are: price discovery and executable pricing;
legal enforceability or validity of contracts; settlement finality,
payment-system access, and CSD connectivity; regulatory-submission
gateways and report formatting; model governance and validation; and
accounting-policy decisions. These concerns lie outside the Ledger
boundary and are reconciled against it only at its perimeter.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

By making reconciliation failures structurally unreachable within its
scope, by rendering broken states impossible through type discipline and
admission enforcement, and by cleanly separating the admission of
structurally valid transactions from the verification of their economic
correctness, the Ledger framework offers a principled foundation for a
post-trade system of record: simpler, because there is only one record,
and more robust, because the classes of internal break that dominate
conventional architectures cannot arise. Every historical state remains
exactly reconstructible, and every economic outcome is traceable to a
deterministic computation over recorded data.

The principal engineering challenge --- and the principal opportunity
--- lies in the disciplined construction of the smart-contract suite
that faithfully captures the terms of every instrument the system must
support, and in the orchestration of the event streams that trigger
those contracts. When that challenge is met, the dominant operational
risk of post-trade processing --- reconciliation --- will have been
substantially eliminated by construction rather than mitigated by
process.

\hypertarget{authority-and-amendment}{%
\section{Authority and Amendment}\label{authority-and-amendment}}

This manifesto is the authoritative statement of intent and scope for
the Ledger project --- its architectural constitution --- and the
direction of authority is one-way: the manifesto is rated against
nothing; everything else is rated against it. All specifications,
implementations, tests, and integration layers --- including every
existing version of the ledger specification --- shall be assessed for
conformance against it and brought into line where they deviate. Where a
genuine improvement requires departing from this document, the manifesto
is amended explicitly first; a design choice that cannot be traced to
its principles shall be rejected. Its vocabulary --- unit, wallet,
balance, move, transaction, watch, the immutable log, projection, the
Event Monitor, the Events Executor, the Transaction Executor, smart
contract, the market data operator, the three homes, virtual wallet,
virtual ledger --- is fixed: one name per component, and no
specification shall introduce a synonym for any of them.

\subsection*{Amendment record}

\textbf{Version 1.1 (11 July 2026).} Three amendments, proposed by the
v15 Phase~1 Design Ruling on the collateral generalisation and ratified
by the owner: Section~4's coordinate sentence restated as a signed
vector --- owned, lent, posted --- with the five original names as its
rays; Section~4 extended by one sentence fixing that non-owned
coordinates carry mass never value, always under a named agreement, and
that a lifecycle event extinguishes value never mass; Section~8's
three-case inflow sentence sharpened to key the case on the declared
legal regime of the governing agreement --- settled-to-market margin as
settlement, title-transfer collateral (cash included) as financing
against an obligation unit, and security-interest collateral without
right of use as custody.

--- End of Manifesto ---

\end{document}