\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage[hidelinks]{hyperref}

\titlespacing*{\section}{0pt}{3pt}{1pt}
\titleformat*{\section}{\large\bfseries}
\setlength{\parskip}{0pt}
\setlength{\abovedisplayskip}{3pt}\setlength{\belowdisplayskip}{3pt}
\setlength{\emergencystretch}{1em}

\theoremstyle{definition}\newtheorem{definition}{Definition}
\theoremstyle{plain}\newtheorem{proposition}{Proposition}
\theoremstyle{remark}\newtheorem{remark}{Remark}

% ---- Frozen notation: single source of truth. Never type a symbol literally. ----
\newcommand{\Log}{L}
\newcommand{\ev}{e}                  % event
\newcommand{\st}{x}                  % true state
\newcommand{\obs}{y}                 % observed value
\newcommand{\acc}{p}                 % accuracy = precision = 1/variance
\newcommand{\Amap}{A}                % observation map
\newcommand{\Rcov}{R}                % noise covariance
\newcommand{\mean}{\mu}
\newcommand{\cov}{\Sigma}
\newcommand{\Vsub}{V}                % constraint subspace
\newcommand{\Ent}{H}                 % differential entropy (E1)
\newcommand{\KL}{D}                  % relative entropy / information gain (E3)
\newcommand{\Ncal}{\mathcal{N}}
\DeclareMathOperator{\fold}{fold}
\DeclareMathOperator{\tr}{tr}

% verdict marks
\newcommand{\yes}{\ensuremath{\checkmark}}
\newcommand{\no}{\ensuremath{\times}}
\newcommand{\hf}{\ensuremath{\circ}}

\title{\vspace{-2em}Ledger Entropy: Does an Entropy of the Log Exist?}
\author{Ledger Entropy Seminar}
\date{\today}

\begin{document}
\maketitle
\vspace{-2.2em}

\begin{center}\fbox{\parbox{0.95\linewidth}{\small\textbf{Non-normative.}
This note is an exploratory seminar work product. It states no requirement, binds no
design, and amends no part of the specification or the Constitution. A negative result is
a successful outcome.}}\end{center}
\vspace{-0.4em}

\section{The Question}
The events of a ledger are not always fully verified. The loose statement ``each event
carries a probability of being accurate'' is not crisp. We make it crisp by a restriction:
every event under study records an \emph{observed value}, and observed $=$ true $+$ Gaussian
noise. Under that restriction, does a \emph{ledger entropy} make sense? If entropy is the
wrong functional, which functional is right? We define what kind of object such an entropy
would be, name three candidates, and test each. The verdict is a qualified yes: a single
absolute entropy of a ledger does not make sense, but a well-posed relative functional does.

\section{Setting}
Fix a valuation time and one conservation triple (unit, coordinate, agreement). Ten objects,
in dependency order.

\begin{enumerate}[leftmargin=1.4em,itemsep=0pt,topsep=2pt]
\item \textbf{Log} $\Log$ --- a finite append-only sequence of admitted events, ordered by admission.
\item \textbf{Event} $\ev_i$ --- one admitted record appended to the log at position $i$.
\item \textbf{Observed-value event} --- an event carrying a numerical reading $\obs_i\in\mathbb{R}^{d_i}$ together with an accuracy $\acc_i>0$.
\item \textbf{Noise model} --- the assumption $\obs_i=\Amap_i\st+\varepsilon_i$, with $\varepsilon_i\sim\Ncal(0,\Rcov_i)$, $\Rcov_i=\acc_i^{-1}I$, independent across events; $\Amap_i$ is a fixed observation map. Accuracy is precision: $\acc_i$ large means $\Rcov_i$ small.
\item \textbf{Candidate-log configuration} --- a finite log together with its fixed noise model and prior; the complete data on which any functional is evaluated.
\item \textbf{True state} $\st\in\mathbb{R}^n$ --- the unobserved vector of which the events are noisy readings.
\item \textbf{The fold} --- the measurable map $\fold$ sending a configuration to its posterior parameters $(\mean,\cov)$, measurable in the readings $(\obs_i)$.
\item \textbf{The posterior} --- the Gaussian law $\Ncal(\mean,\cov)$ on the true state induced by prior and noise model via $\fold$.
\item \textbf{Attestation} --- a refinement of an observed-value event that raises its accuracy $\acc_i$ (shrinks its noise), leaving $\Amap_i$ fixed.
\item \textbf{Constraint subspace} $\Vsub\subseteq\mathbb{R}^n$ --- the affine subspace $\{\st:C\st=c\}$, $\operatorname{rank}C=r$, on which conserved quantities are fixed; the posterior is supported on $\Vsub$, so its ambient covariance $\cov$ is rank-deficient.
\end{enumerate}

\noindent\textbf{Notation.} $\Log$ log, $\ev_i$ event, $\obs_i$ reading, $\acc_i$
accuracy/precision, $\Amap_i$ observation map, $\Rcov_i$ noise covariance, $\st$ true state,
$(\mean,\cov)$ posterior mean and covariance, $\Vsub$ constraint subspace, $\Ent(\cdot)$
differential entropy, $\KL(\cdot\,\|\,\cdot)$ relative entropy (Kullback--Leibler). No symbol
is introduced elsewhere.

\section{Candidate Definitions}
What is random is not the state --- it is a fixed unknown --- but our \emph{belief} about it,
the posterior. Any ledger entropy is a functional of that posterior. Three candidates.

\begin{itemize}[leftmargin=1.4em,itemsep=1pt,topsep=2pt]
\item[\textbf{E1}] \emph{Absolute differential entropy of the state posterior:}
      $\Ent=\tfrac12\log\det(2\pi e\,\cov)$.
\item[\textbf{E2}] \emph{Hybrid mixture total.} Keep also the discrete layer: a configuration
      $K$ records which candidate log is true, with weights $w_k$ built from the accuracies
      $\acc_i$. The posterior is the mixture $\pi=\sum_k w_k\,\Ncal(\mean_k,\cov_k)$, and E2 is
      its total entropy, combining discrete bits and continuous nats. This is the literal
      reading of ``probability of being accurate.''
\item[\textbf{E3}] \emph{Relative entropy / invariant information gain:} $\KL(\text{posterior}\,\|\,\text{reference})$,
      equal on the observation channel to the mutual information $I(\st;\obs)$, with closed form
      $\tfrac12\log\det\!\big(I+\cov\,\Amap^{\!\top}\Rcov^{-1}\Amap\big)$.
\end{itemize}

\begin{center}
\begin{tabular}{lcccccc}
\toprule
 & well-posed & invariant & compositional & attest.-mono. & conserv.-compat. & computable \\
\midrule
\textbf{E1} & \no & \no & \yes & \yes & \no & \yes \\
\textbf{E2} & \no & \no & \hf$^{a}$ & \yes & \no & \no \\
\textbf{E3} & \hf$^{b}$ & \yes & \yes & \yes & \yes & \yes \\
\bottomrule
\end{tabular}
\end{center}
\noindent{\footnotesize \yes\ holds unconditionally; \no\ fails, with a counterexample in
Results; \hf\ holds under a stated side-condition.
$^{a}$ the joint chain-rule total is additive over independent sub-logs, but the observable
marginal does not factor.
$^{b}$ well-posed once a reference is supplied; the ledger supplies it (see Verdict and
Declaration).}

\medskip
\noindent The two E1 failures are \emph{independent}: invariance fails for units, conservation
fails for degeneracy, and repairing one does not repair the other (Prop.~\ref{p:unit},
Prop.~\ref{p:cons}, and its remark). E2 inherits both and adds a third, sharper one. E3 fails
nothing structural; its only \hf\ is the reference, which the ledger itself resolves.

\section{Results}
Throughout, the prior is Gaussian and proper on the free directions of $\Vsub$; observations
are linear-Gaussian.

\begin{proposition}[attestation reduces differential entropy]\label{p:attest}
An observation $\obs=\Amap\st+\varepsilon$, $\varepsilon\sim\Ncal(0,\Rcov)$, $\Rcov\succ0$, sends
the posterior $\Ncal(\mean,\cov)\mapsto\Ncal(\mean',\cov')$ with
$\cov'^{-1}=\cov^{-1}+\Amap^{\!\top}\Rcov^{-1}\Amap$, hence $\cov'\preceq\cov$ and
$\Ent'\le\Ent$. The drop equals the information gained,
$\Ent-\Ent'=\tfrac12\log\det(I+\cov\,\Amap^{\!\top}\Rcov^{-1}\Amap)=I(\st;\obs)\ge0$, with
equality iff $\Amap$ loads on no non-degenerate direction.
\end{proposition}
\begin{proof}
The information form is standard conjugacy. Then
$\cov-\cov'=\cov\Amap^{\!\top}(\Amap\cov\Amap^{\!\top}+\Rcov)^{-1}\Amap\cov\succeq0$, of the form
$M^{\!\top}\!AM$ with $A\succ0$; determinant is monotone under the Loewner order on positive
definite matrices, so $\det\cov'\le\det\cov$ and $\Ent'\le\Ent$. The identity
$\det\cov/\det\cov'=\det(I+\cov\Amap^{\!\top}\Rcov^{-1}\Amap)$ follows from
$\cov'^{-1}=\cov^{-1}(I+\cov\Amap^{\!\top}\Rcov^{-1}\Amap)$; it is $>1$ iff $\Amap\ne0$ on a
free direction.
\end{proof}

\begin{proposition}[conservation forbids an absolute entropy]\label{p:cons}
Exact conservation $C\st=c$, $\operatorname{rank}C=r$, confines the posterior to the affine
subspace $\Vsub$, which is Lebesgue-null in $\mathbb{R}^n$. Then $C\cov=0$, $\det\cov=0$, and the
ambient $\Ent=\tfrac12\log\big((2\pi e)^n\det\cov\big)=-\infty$. On $\Vsub$ with its
$(n-r)$-dimensional Lebesgue measure, $\Ent_{\Vsub}=\tfrac12\log\big((2\pi e)^{n-r}\det{}_{+}\cov\big)$
is finite, where $\det_{+}$ is the product of the nonzero eigenvalues.
\end{proposition}
\begin{proof}
$\operatorname{Var}(C\st)=C\cov C^{\!\top}=0$ gives $C\cov=0$, so $\cov$ is singular and the ambient
formula is $-\infty$. Restricting to an orthonormal basis of the tangent space of $\Vsub$
replaces $\det\cov$ by $\det_{+}\cov$, which is finite.
\end{proof}
\begin{remark}[the two obstructions are independent]
$\Ent_{\Vsub}$ cures degeneracy but not units: under a basis change $B\mapsto BM$ of $\Vsub$,
$\Ent_{\Vsub}\mapsto\Ent_{\Vsub}+\log|\det M|$. Fixing the support does not fix the scale.
\end{remark}

\begin{proposition}[differential entropy is not reparameterization-invariant]\label{p:unit}
For an invertible smooth change of coordinates $\obs=g(\st)$,
$\Ent(\obs)=\Ent(\st)+\mathbb{E}\big[\log|\det J_g(\st)|\big]$; for a linear rescale $\obs=B\st$,
$\Ent(B\st)=\Ent(\st)+\log|\det B|$.
\end{proposition}
\begin{proof}
Change of variables in the density, $p_{\obs}(\obs)=p_{\st}(g^{-1}(\obs))\,|\det J_{g^{-1}}|$,
substituted into $-\int p\log p$ (Cover--Thomas, ch.~8).
\end{proof}
\noindent Consequence: recording the free coordinates in cents rather than dollars
($B=100\,I_{n-r}$) shifts $\Ent$ by $(n-r)\log100$. The absolute number, and even its sign, is a
unit artefact. Only differences, relative entropy, and mutual information survive --- which is
exactly the closed form of Prop.~\ref{p:attest}, invariant because the unit Jacobians cancel in
the density ratio, and finite on $\Vsub$ because the singular directions of posterior and
reference cancel there. This is E3.

\paragraph{Chain-rule identity (E2), and its unit caveat.}
For the discrete configuration $K$ and state $X$,
$\Ent(K,X)=\Ent(K)+\sum_k P(k)\,\Ent(X\mid K=k)$; this is an exact identity. But $\Ent(K)$ is
dimensionless Shannon (bits, $\ge0$, invariant) while $\Ent(X\mid K=k)$ is differential (carries
units, sign-indefinite, non-invariant). Their sum adds bits to nats-with-units and inherits the
reparameterization dependence of Prop.~\ref{p:unit}. This is the \emph{sharpest} obstruction to
E2: it ships only with a declared quantization scale $\delta$ (using
$\Ent_\delta(X)\approx\Ent(X)-(n-r)\log\delta$ throughout), or by keeping the pair
$(\Ent(K),\Ent(X\mid K))$ unsummed, or by going relative on both --- which is again E3.

\paragraph{Second law (Heuristic).}
The claim ``entropy is non-decreasing between attestations and drops at each'' is
\emph{false as stated}. With a deterministic fold and a static latent state, conservation-preserving
moves push measure forward with $|\det|=1$, so $\Ent$ is constant between attestations and falls
only at them (Prop.~\ref{p:attest}); an append-only observation ledger has entropy that only
ever ratchets \emph{down}, the opposite of a thermodynamic second law. A genuine non-decrease
requires an injected diffusion prior $Q\succeq0$ ($\cov\mapsto F\cov F^{\!\top}+Q$), a modelling
choice not forced by the framework. With it the picture is a sawtooth: rise by $Q$ between,
drop at each attestation. Status: heuristic, conditional on a declared $Q$.

\paragraph{Mixture marginal (cited fact).}
The marginal of the E2 mixture $\sum_k w_k\Ncal(\mean_k,\cov_k)$ has no closed-form differential
entropy (Huber et al., 2008); only bounds and approximations exist. So the single E2 total is
not closed-form computable, and summing its $2^{|\Log|}$ configuration terms is exponential.

\section{Worked Example}
This instantiates Prop.~\ref{p:attest} (attestation) and Prop.~\ref{p:unit} (units). One
instrument; wallets $A$, $B$, issuer $I$; latent quantity the true price $\st$. Accuracy is
precision, $\acc=1/\text{variance}$.

\vspace{-0.5em}\begin{center}\small
\begin{tabular}{llll}
\toprule
event & effect & reading & noise \\
\midrule
$\ev_1$ issue 100 to $A$ & $I:-100,\ A:+100$ (exact) & --- & none \\
$\ev_2$ observe price & valuation & $\obs_1=50.00$ & $\acc_1=100$ ($\sigma_1=0.10$) \\
$\ev_3$ transfer $A\!\to\!B$ qty 40 & $A:-40,\ B:+40$ (exact) & --- & none \\
$\ev_4$ attest $\ev_2$ & valuation & $\obs_2=50.04$ & $\acc_2=400$ ($\sigma_2=0.05$) \\
\bottomrule
\end{tabular}
\end{center}

\noindent\emph{Chair ruling (ratified): quantities are exact; only values are noisy.} A move is
an instruction, not a measurement. Conservation vanishes by paired-leg construction precisely
because the two legs are the same number posted with opposite signs. If a transfer quantity were
a noisy observation, either both legs share the draw (the noise cancels in the wallet sum and the
``noise'' is illusory) or the legs are independent draws (conservation is violated in
expectation, which the framework forbids). So after $\ev_3$, $A$ holds exactly $60$ units and $B$
exactly $40$; all uncertainty lives on the price.

\smallskip\noindent\textbf{Posterior, two independent ways.}
Sequential update (flat prior; $\ev_2$ sets $\Ncal(50.00,0.01)$; gain
$K=0.01/0.0125=0.8$): mean $=50.00+0.8(0.04)=\mathbf{50.0320000000}$, variance
$=0.2\times0.01=\mathbf{0.0020000000}$. Batch precision-weighting ($\acc_1=100,\acc_2=400$):
variance $=1/500=\mathbf{0.0020000000}$, mean $=0.002(100\cdot50.00+400\cdot50.04)=\mathbf{50.0320000000}$.
Identical to ten digits: $\st\sim\Ncal(50.032,\,0.002)$.

\smallskip\noindent\textbf{E1, two independent ways.} $\Ent=\tfrac12\ln(2\pi e\,\sigma^2)$
(closed form) versus $-\int f\ln f$ (numerical quadrature, $\pm12\sigma$):

\vspace{-0.5em}\begin{center}\small
\begin{tabular}{lccc}
\toprule
state & variance & $\Ent$ closed (nats) & $\Ent$ quadrature (nats) \\
\midrule
after $\ev_2$ & $0.01$  & $-0.883647$ & --- \\
after $\ev_4$ & $0.002$ & $\mathbf{-1.688366}$ & $\mathbf{-1.688366}$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent The negative sign and the unit-dependence of these levels are exactly
Prop.~\ref{p:unit}: read the price in cents and every $\Ent$ shifts by $\ln100$. The
\emph{change}, however, is invariant. \textbf{E3:} the attestation deposits
$\Ent-\Ent'=\tfrac12\ln(0.01/0.002)=\tfrac12\ln5=\mathbf{0.804719}$ nats
$=1.160964$ bits $=I(\st;\obs_2\mid\obs_1)$, matching
$\tfrac12\ln(1+\acc_2\sigma_{\text{before}}^2)=\tfrac12\ln(1+400\cdot0.01)=\tfrac12\ln5$; and
$\KL(\text{posterior}\,\|\,\text{prior})=0.4559$ nats. The two marks agree
($|\Delta|=0.04$ versus $\sigma_1=0.10$, about $0.4\sigma$): no attestation dispute. The
operator's reading is the per-position mark and its confidence interval --- $A$'s line
$60\times50.032=3001.9\pm2.7$, $B$'s $40\times50.032=2001.3\pm1.8$ --- and the per-attestation
gain in bits, both invariant; not $\Ent$ itself.

\section{Verdict}
\emph{Signed, GROTHENDIECK, Chair.}
The word ``entropy'' is the wrong word for a ledger, and the right functional is
\emph{relative entropy}, the invariant information gain E3.
An absolute entropy of a ledger state (E1) fails two independent tests: it is not
\emph{invariant} --- Prop.~\ref{p:unit}, dollars-to-cents moves the number and its sign --- and it
is not \emph{conservation-compatible} --- Prop.~\ref{p:cons}, the conservation law pins the state
to a subspace where the ambient entropy is $-\infty$. Curing one does not cure the other. The
hybrid total (E2), which takes the ``probability of being accurate'' literally, inherits both and
adds a third and sharper defect: it sums Shannon bits to differential nats, quantities of
different type, and has no closed form. What survives all six tests is E3: the relative entropy of
the current posterior against a reference, equal to the information the event deposits about the
truth. It is unit-invariant (the Jacobians cancel in a ratio); degeneracy-safe (posterior and
reference share the constraint subspace, so singular directions cancel); additive over
independent sub-logs, hence checkpoint-independent and order-insensitive; monotone under
attestation; and cheap, needing only the marginal covariance on the coordinates an observation
touches. The one thing E3 needs is a reference, and the ledger already keeps one: the
\emph{as-known-at} state before the event. So the honest object is not ``the entropy of the
ledger'' but ``the information this event added, measured against what was known before.''
Entropy does not make sense as a level. Information gain makes sense as a difference. That
difference is the functional the ledger admits.

\section{Open Questions and Related Work}
\textbf{Classification.} Under the restriction, this is a constrained Kalman / information filter
over a Gaussian field --- a linear-Gaussian estimation problem in an entropy costume. Every
candidate is a scalarization of the posterior covariance $\cov$: E1 is $\log\det\cov$
(D-optimality), $\tr\cov$ and $\lambda_{\max}(\cov)$ its A- and E-optimality siblings. E3 is the
Bayesian experimental-design criterion (Lindley), which turns the survivor into a policy: point
the scarce attestation budget at the observation of largest expected gain.

\smallskip\noindent\textbf{Related work} (contribution / bearing on this question).
{\small \emph{Shannon} 1948 --- coding rate of a source; fixes what this is not, the log is a record.
\emph{Jaynes} 1957 --- the prior is max-entropy given the conservation moments.
\emph{Kalman} 1960 --- the actual engine; the restricted ledger is an information filter.
\emph{Kullback--Leibler} 1951 --- relative entropy, the invariant object E3.
\emph{Amari} --- Fisher information $\cov^{-1}$ as metric; $\KL$ is its second-order form.
\emph{Baez--Fritz--Leinster} 2011 --- entropy as information loss along a map; the fold is lossless replay, so the analogue is gain, not loss.
\emph{Fritz} 2020 --- Markov categories; a second telling of the update only.
\emph{Pearl} --- belief propagation, the feasible computation on the sparse coordinate graph.
\emph{Valiant} (PAC) --- ``probability an event is accurate'' as a learnable reliability parameter, orthogonal to the functional.
\emph{Dempster--Shafer} --- the imprecise-probability layer the Gaussian restriction deletes; needed back only if $\acc_i$ were an interval.
\emph{Lindley} 1956 --- Bayesian experimental design, the operational keep.
\emph{Prior art, not a source:} the audit-risk model $AR=IR\times CR\times DR$ (AICPA; ISA/AU-C 200) is a heuristic product of subjective probabilities, with no derivation the ledger could reuse.}

\smallskip\noindent\textbf{Open.} (i) The diffusion prior $Q$ that would make a second law
non-vacuous --- what, if anything, in the framework supplies it. (ii) A discrete-layer attestation
calculus: E2 read as ``which events are admitted'' is a genuine object; its well-posed relative
form is unsettled. (iii) Computability at scale: for $\sim10^{6}$ live coordinates a global
$\log\det$ is governed by the treewidth of the coordinate-dependency graph, which must be
measured, not assumed; the local E3 gain is cheap regardless.

\begin{remark}[second telling; delete this box and nothing above changes]
In the Kleisli category of the Giry monad, the noisy log is one arrow $\Theta\to S$: a tensor of
Gaussian observation channels followed by the deterministic fold. The posterior is its Bayesian
inverse; attestation is post-composition of a further channel. Information gain is a monoid
homomorphism $([\text{Event}],{+}{+})\to(\mathbb{R}_{\ge0},+)$, which is why the total of E3 is
checkpoint-independent and order-insensitive by construction. No proof above rests on this
paragraph.
\end{remark}

\section{Declaration}
\textbf{Sessions.} Sessions 1--4 were used: nine independent Session-1 formalizations by distinct
routes (analytic, symmetry, honesty gate, categorical, unifying object, rigor veto, architecture,
classification, framing); a Session-2 synthesis into E1/E2/E3 against the six criteria; a
Session-3 worked example; this Session-4 write-up.

\smallskip\noindent\textbf{Chair rulings.} (1) \emph{The toy decision:} quantities are exact,
only values are noisy, on the conservation argument given in the Worked Example --- ratified. (2)
\emph{The reference gap:} the reference E3 requires is \emph{closed} for the incremental gain and
remains a \emph{conjecture} for any absolute total. The incremental reference is the previous
as-known-at posterior, which the bitemporal doctrine already maintains --- no new object is
posited, so the per-event gain is well-posed. A cumulative total would need a canonical origin
(the initial prior); that the max-entropy-given-conservation prior is that origin is natural but
not forced, hence a conjecture, not a proposition.

\smallskip\noindent\textbf{Labels.} The rigor constraints W1--W7 were enforced: every
differential quantity names its reference; only invariants are headline numbers; all statements
live on the constraint subspace of dimension $n-r$; bits and nats are never summed without a
declared $\delta$; each object is named; the second-law claim names the diffusion prior it needs;
the mixture marginal is invoked only with its no-closed-form citation. Only
Propositions~\ref{p:attest}--\ref{p:unit} carry proofs; the chain rule is an identity with a
caveat; the second law is a heuristic; the mixture fact is cited. The E3 closed form is stated as
a corollary of Proposition~\ref{p:attest}, not minted as a new theorem --- the one label decision
this chair made to stay within the approved set.

\smallskip\noindent\textbf{Dissents, independence, confidence.} No dissents. The nine notes
shared no cross-reading yet converged by distinct routes --- either strong evidence or a common
prior in the authors' training; recorded as convergence, not proof. Confidence: high on the
classification and the invariance argument; medium on global computability, which hinges on a
treewidth that must be measured.

\end{document}
