% temporalv16.tex — Temporal execution addendum for the Ledger Spec v16.0
% Produced by five independent temporal-engineer instances (TE-1..TE-5) under the
% convergence protocol; unanimous register, three rounds, no panel ruling required.
% Placement: LedgerManifesto/, the repository's convention for cross-cutting documents
% (the Manifesto, TuringReview, next_step_v16 all live here); not genuinely ambiguous,
% so no Decision Panel placement ruling was needed.
% This is a SUBSTRATE PROPOSAL, not an amendment. The ledger's event log remains the
% sole book of record; Temporal workflow histories are never the source of truth.
\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{microtype}
\setlength{\emergencystretch}{2em}

\title{\textbf{Temporal as Execution Substrate for the Ledger v16}\\
\large An addendum: five-engineer consensus, v1.0}
\author{TE-1 \ldots{} TE-5 (temporal-engineer $\times$5), converged}
\date{15 July 2026}

\begin{document}
\maketitle

\section{The Way}

Temporal runs the two \emph{untrusted} machines and nothing else. The Event Monitor and
the Events Executor become durable, replayable orchestration; the Transaction Executor
and the immutable log stay outside Temporal as the trusted single writer, fronted by one
idempotent propose-to-door activity --- the only write path. The ledger log is the sole
book of record. Temporal history is a disposable execution cache, and the acceptance
test is operational: wipe the Temporal cluster, and the entire workflow population
re-derives from the ledger --- UnitStatus re-read, watches re-armed, nothing lost.

\textbf{One workflow per unit.} Each unit runs one long-lived workflow, keyed by its
\texttt{unitId}, walking the unit's product graph. The workflow's pending timers and
signal-waits equal the current node's out-edge set --- the armed watch list of the
graph-consistency invariant. The in-memory node is a mirror, never authoritative; it is
rehydrated from the ledger on start, on ContinueAsNew, and on any doubt. Units the
ledger itself creates --- the settlement-obligation unit, a market-claim unit, the
partial-split successor legs --- start their own independent unit-workflows
(signal-with-start, no Temporal parent--child coupling: parental linkage is
ledger-absent relational state). Branches of a unit's own graph, such as the
market-claim branch at a node, stay inside that unit's single workflow.

\textbf{Side effects live in activities only.} Capture an arrival, fetch a projection,
evaluate a contract, propose to the door, emit a settlement instruction --- each is an
activity. Workflow code holds timers, signal-waits, selectors, and sequencing, and
nothing else. Smart contracts run in version-pinned, queue-routed activities (never
local activities), so ledger economics never enters Temporal's replay or versioning
surface: the ledger replays by re-reading admitted transactions, Temporal by re-reading
its own history, and the two never meet.

\textbf{Watches.} Date watches are in-workflow durable timers; observation and barrier
watches are signal-driven evaluation against recorded observations, never a timer poll.
Temporal Schedules serve system cadence only (end-of-day sweeps, overdue-watch
reconciliation) --- a 252-fixing swap is one workflow, not 252 schedules. The lifecycle
declared$\to$armed$\to$fired/expired is authoritative in the ledger: arming writes the
recorded acknowledgement through the door, and Temporal supplies only the liveness half.

\textbf{The boundary.} Capture precedes routing: a pushed arrival is recorded through
the door --- capture envelope, quarantine-or-route --- by the ingestion activity before
any workflow signal exists. Signals carry timing and reference (a cause identifier, a
log position), never economic payload, and every signal has a record-derived backstop:
a lost signal costs latency, never the record. The settlement interface remains a pure
projection with no write privilege; settlement failure returns as a recorded external
event and walks the settlement-obligation unit's graph --- never a Temporal-side
compensation or saga rollback.

\textbf{Idempotence.} The door's cause-derived transaction identifier, computed from
recorded inputs and never from Temporal run or attempt identifiers, absorbs Temporal's
at-least-once activity delivery into exactly-once commitment. Temporal's own
deduplication is tolerated as optimisation, never trusted for correctness.

\section{Decision Register}

Unanimous after three rounds. R-01..R-21 as amended in Round 2; R-22..R-27 added from
Round-1 missing-decision returns.

\begin{enumerate}[label=\textbf{R-\arabic*.}, leftmargin=*, itemsep=1pt]
\item Temporal runs only the two untrusted machines; the Transaction Executor is an
  external single-writer service fronted by one idempotent door activity (M6).
\item The ledger log is the sole book of record; Temporal history is disposable;
  wipe-and-rebuild from the ledger is the standing acceptance test (C-2.2).
\item One long-lived workflow per unit, keyed by plain \texttt{unitId} in the
  authoritative production namespace; lineage is a namespace-level fact, and
  \texttt{lineageId:unitId} is reserved for shared simulation namespaces where
  non-authoritative lineages deliberately co-locate.
\item The workflow's node and watch set mirror the ledger and are rehydrated from it;
  pending timers and signal-waits equal the current node's out-edge set.
\item Every side effect is an activity; workflow code is orchestration only.
\item Contracts run in version-pinned activities, never workflow code.
\item Idempotence is the door's cause-derived txid from recorded inputs; at-least-once
  delivery becomes exactly-once at the door; Temporal dedup is never load-bearing.
\item Date watches are durable timers; condition watches are signal-driven evaluation
  over recorded observations, never polls.
\item Schedules serve system cadence only; per-unit contractual dates never use them.
\item The watch lifecycle is authoritative in the ledger; arming is a recorded
  acknowledgement through the door; Temporal supplies liveness (M1, M3).
\item Signals carry timing and reference only, never payload (M4), with record-derived
  backstops.
\item Capture precedes routing: envelope-first through the door before any signal
  (M8, TA-ARRIVAL); a signal is never the sole record of an arrival.
\item Updates serve human gates only (forward repair, quarantine disposition,
  authorised fork), returning the door's admit/refuse; boundary capture never rides an
  Update; Queries are a non-authoritative liveness view.
\item Ledger-created units start their own independent unit-workflows
  (signal-with-start, no parent--child coupling); graph branches stay in the owning
  unit's workflow; corporate-action cascades fan out as signals to existing units, each
  firing proposing its own txid. \emph{Round-3 note: TE-2's dispute withdrawn on the
  certified partial-edge text --- the split legs are genuinely new units.}
\item ContinueAsNew at node transitions and every K fixings, carrying only
  \{unitId, nodeId, log-cursor\}; rehydration and disaster recovery are the same
  re-derivation.
\item No run/execution timeouts on unit workflows; liveness is bounded by watch windows
  and obligation deadlines surfaced as overdue items. Door activity: short
  start-to-close, unlimited retry for transient failures only (see R-22). Long waits are
  timer-plus-signal; only egress and batch activities heartbeat.
\item Two versioning axes stay separate: Temporal Build-IDs/GetVersion for
  orchestration (ContinueAsNew boundaries as cutover points); ProductTerms versions on
  the log for economics, which alone bear C-2.2.
\item Task queues split by trust and scaling: door (sole write credential, serialized,
  backpressure-alarmed), contracts (read-only, versioned), ingestion (must-not-lose),
  unit-orchestration (idle-many), settlement (egress-only).
\item Search attributes (unit, node, next watch due, obligation status, overdue and
  quarantine flags) are an ops projection reconciled to the ledger, never authoritative.
\item Authorised forks and simulated ledgers run in isolated namespaces against their
  own lineage's door, never the production door (C-12.5, C-2.8).
\item Economic errors are repaired forward through the door; a Temporal
  non-determinism bug degrades to a liveness incident, never wrong ledger state.
\item A door \emph{refusal} is a returned value, not a retryable error: it breaks the
  retry loop and stands as a visible blocked/overdue item awaiting declared precedence
  or a corrected proposal --- never retried to infinity, never a crash.
\item A late or back-dated observation is an ordinary capture-through-the-door arrival;
  affected workflows re-fire on the correction signal under a new txid; the valid-time
  re-derivation is a pure ledger fold --- Temporal never rewinds workflow state.
\item Settlement failure is a recorded external event walking the unit's graph, never a
  Temporal compensation; the settlement projection proposes nothing.
\item Signal-arrival order is orchestration sequencing only; the committed order is
  decided at the door alone (C-2.7).
\item EvaluateContract is a normal queue-routed activity, never a local activity.
\item Fan-out to possibly-not-running workflows uses signal-with-start; record-first
  capture remains the correctness backstop.
\end{enumerate}

\section{What Temporal Must Never Own}

The book of record --- the immutable, hash-chained log and every fact folded from it.
Balances, the three homes, PnL, NAV, every projection. Invariant enforcement:
conservation, coverage and its right-of-use gate, consistency of reference, the
admission decision itself. The total order of transactions. Idempotence keys. The watch
lifecycle's recorded facts (declared, armed, fired, expired). Quarantined events and
capture envelopes. The close-out net and CLM-CO. Anything the ledger must reconstruct
alone: if losing Temporal would lose it, the design is wrong --- Temporal owns
schedules-in-flight and orchestration position, and nothing else.

\section{Operational Constraints}

\textbf{Determinism rules for workflow code.} No wall clock (\texttt{workflow.Now} only,
and never feeding a contract), no randomness, no iteration over unordered maps, no
direct I/O, no local activities for contract evaluation; all non-determinism lives in
activities whose results Temporal records.

\textbf{Versioning plan.} Orchestration changes ride Worker Versioning; old runs drain
to their next ContinueAsNew boundary, which is the standing cutover point, so GetVersion
patches are rare. Economics changes are new ProductTerms versions plus new version-pinned
pure functions on the record; they never touch workflow code.

\textbf{Timeouts.} Door activity: start-to-close of seconds, unlimited retry on
transient failure, refusal returned as a value (R-22). Contract evaluation:
start-to-close bounded by pricing cost, heartbeat none. Ingestion capture:
schedule-to-close tight, since capture loss is the one unacceptable latency. Egress and
batch reconciliation: heartbeated. No workflow run or execution timeouts (R-16).

\textbf{ContinueAsNew policy.} At every product-graph node transition and every K
fixings (K sized to keep history well under the server limit); carried state is the
identifier triple only; everything else rehydrates from the ledger.

\textbf{Worker topology.} Five queue families (door, contracts, ingestion,
unit-orchestration, settlement) with separate worker pools, credentials, and scaling;
the door pool alone holds ledger-write credentials, sized to the single writer, with
schedule-to-start latency alarmed as the backpressure signal.

\section{Declaration}

\textbf{Rounds used.} Round 0 (five independent papers), Round 1 (21-entry consolidated
register: 18 unanimous, 3 disputed, 6 missing decisions raised), Round 2 (amended
register: 49 of 50 rulings AGREE), Round 3 (single scoped fact-check on R-14 against the
certified v16.0 partial-edge text; dispute withdrawn). \textbf{Final register status:
UNANIMOUS on all 27 entries. No Decision Panel ruling was required.}

\textbf{Decision Log.} DL-01 $|$ Phase 2 $|$ authoritative v16 artifact $|$ single
candidate (\texttt{ledger\_v16\_0.tex}, certified) $|$ no panel convened $|$ n/a.
DL-02 $|$ Phase 2 $|$ R-14 dispute (split legs: new units or branches?) $|$ resolved by
the certified specification text, not by vote; TE-2 withdrew $|$ no panel convened $|$
5--0. DL-03 $|$ Phase 3 $|$ placement $|$ repository convention (all cross-cutting
documents live beside the Manifesto: TuringReview precedent) $|$ not genuinely
ambiguous; no panel convened $|$ n/a. \textbf{DEFERRED-TO-OWNER: none} --- the substrate
proposal demands nothing of the Constitution; every entry keeps the ledger as sole
truth.

\textbf{Independence caveat (recorded, not absorbed).} The five instances share the
temporal-engineer agent type's persistent memory; TE-5 disclosed consulting it. Round-0
independence is therefore partial. Mitigation: the adversarial round structure, the
register-level dispute discipline, and the Round-3 fact-check against the specification
text rather than any instance's recollection.

\textbf{Confidence: HIGH} on the architecture mapping and the negative space (they
restate the specification's own trust boundary); \textbf{MEDIUM} on the operational
constants (K for ContinueAsNew, timeout values), which are estimates pending a load
model --- the single biggest unknown is production event volume per unit per day, which
sizes history growth, the door pool, and the ContinueAsNew cadence.

\end{document}
