\section*{Preface}
\addcontentsline{toc}{section}{Preface}

A trading business records the same activity many times over. A trade is captured in the
trading system, again in the risk system, again in the general ledger, and again in
regulatory reports. Each is a separate copy of one economic fact. Because the copies are
kept apart, they drift apart, and the work of checking that they still agree---reconciliation---becomes
a dominant operational cost and a primary source of error. When two records disagree,
someone must decide which is right. Sometimes no one notices in time.

This specification removes the cause rather than managing the symptom. It keeps one record,
not many. Every event in scope is written once, as a single kind of entry: the movement of
a quantity from one place to another. Nothing is ever overwritten or deleted; entries are
only added. A position, a profit-and-loss figure (P\&L, the gain or loss on a book), a
balance sheet, or a regulatory report is not a second copy to be reconciled against the
first. It is a view computed from the one record, as a total is computed from a list of
numbers. Among the firm's own records there is nothing to reconcile, because there is only
one. (Checking that record against outside authorities such as custodians and depositories
remains a separate activity at the system's boundary; it is the internal reconciliation,
between the firm's own copies, that this design removes.)

This is what ``conservation by construction'' means, and why it matters in daily
operation. Conservation is the rule that a quantity is never created or lost inside the
system: whatever leaves one place arrives somewhere else, exactly. The rule is not a check
run afterward and hoped to pass; it is built into the only operation the system permits, so
a record that violated it could not be written in the first place. Three consequences
follow directly.

\begin{itemize}
\item The books cannot silently fail to balance. An entry that did not balance is not a
warning to investigate later---it cannot exist.
\item Value cannot be created or lost inside the system. Quantities are only moved between
places. Gains and losses arise from price changes, which are recorded separately and
plainly, never from a quantity quietly appearing or vanishing.
\item Any past state can be reproduced exactly. Because the record is only ever added to,
the system can be rewound to any earlier moment and will show precisely what was known
then. Audit and investigation work from fact, not from reconstruction.
\end{itemize}

The same single record serves the wider market context. Where the activity must be
expressed in the shared industry vocabulary for derivatives---the Common Domain Model
(CDM), a standard maintained by the International Swaps and Derivatives Association---the
record maps onto it directly, without a separate translation system to keep in step.

A reader wanting the technical argument should continue to the Abstract and to the Roadmap
and Reading Guidance, which give the shortest path through the document for each audience.
