\section{Regulatory Obligations and the Direction of Travel}
\label{sec:sec17}
\label{sec:regulatory}
%==============================================================================

Post-trade regulatory reporting is the dominant compliance cost for derivatives market participants. The same economic event must be reported to multiple trade repositories in jurisdiction-specific formats: EMIR Refit (EU~2024/1856, 203 reportable fields), MiFIR (RTS~22, 65 fields), SFTR (Regulation~EU~2015/2365, 155 fields), and CFTC Parts~43 and~45 (rewritten 2022--2024). Dual-sided reporting under EMIR and SFTR makes the reconciliation burden structural: two counterparties independently produce matching reports for every lifecycle event, and mismatches trigger regulatory queries. Firms therefore maintain parallel pipelines per regime, each with bespoke field mappings, validation, and exception management.

\paragraph{CDM and Digital Regulatory Reporting.}
The ISDA Common Domain Model, governed by FINOS, supplies a single machine-readable vocabulary for trade description. ISDA's Digital Regulatory Reporting (DRR) encodes each regime's reporting rules as executable code over the CDM, so the interpretation of a regulation is a shared, testable artefact rather than a firm-specific implementation. DRR is in production, not sandbox: as of the source below, four institutions file DRR-generated reports with live trade repositories, with further institutions in proof-of-concept, at reported acknowledgement rates above 98\% under EMIR Refit.\footnote{ISDA, ``Digital Regulatory Reporting: 2025 Year in Review,'' January 2026. Figures are pinned to that publication and not restated as current.} The convergence point is a unified post-trade stack: a trade described once in CDM, reports generated mechanically by DRR, settlement executed by CDM-native contracts.

\paragraph{Framework alignment.}
The architecture derives this alignment from its own primitives rather than adopting CDM because a standard mandates it. Transactions are expressed in CDM (Section~\ref{sec:cdm}); instruments are deterministic move-generating contracts (Section~\ref{sec:contracts}); the canonical internal record is the immutable event log (Section~\ref{sec:ledger}). The economic content of a DRR report---the lifecycle event, its quantities, and its timing---is therefore already present in the move stream and is generated from it by projection, with no separate trade-data pipeline.

Two boundaries bound this claim and are stated once. First, a complete report (e.g.\ EMIR Refit's 203 fields) also requires reference-data enrichment---UTI generation, LEI validation, counterparty classification, collateral detail---that does not reside in the move stream; this enrichment is supplied by the reference-data authority at the ledger boundary, specified in the standalone data spec (pointer per Section~\ref{sec:scope-limitations}), not by this framework. Second, the framework supplies technical substrate, not organisational compliance.

\paragraph{Conformance checkpoints.}
The event log satisfies the audit-trail requirements of BCBS~239 Principle~6 (accuracy and integrity) and Principle~11 (distribution): every report projects from one immutable record. The property-based testing strategy (Section~\ref{sec:invariants}), deterministic lifecycle functions, and immutable audit trail are the technical substrate for the testing and traceability requirements of the EU Digital Operational Resilience Act (DORA, Regulation~EU~2022/2554), Article~8 (ICT risk management) and Chapter~IV (digital operational resilience testing). Full DORA conformance requires operational and organisational measures outside the scope of a design specification.

\paragraph{Mandate issuance reporting.}
Representing a managed-account mandate as a unit creates an SFTR/EMIR reporting surface for mandate issuance. The framework supplies the mechanism (flag~F5): the \texttt{reportable} predicate on \texttt{ProductTerms} gates whether a unit's issuance requires a UTI/LEI pair and which units project into the reporting layer; the managed-account treatment (Section~\ref{sec:managed}) carries the construction. F5 remains flagged pending external regulatory sign-off (Section~\ref{sec:conclusion}).

\clearpage
%==============================================================================
