\section{Conclusion and Open Problems}
\label{sec:sec19}
\label{sec:conclusion}
%==============================================================================

\subsection{Summary}

All economic activity within scope uses one primitive: the atomic wallet-to-wallet move. Every instrument---equity, bond, derivative, swap, managed account, QIS strategy---is a deterministic contract that emits moves. One law (conservation) governs the system; one immutable event log records it; pure lifecycle functions extend it, giving idempotent processing and replay to any past state.

The scalar position---one number per $(\text{wallet},\text{unit})$ pair---suffices for instruments where ownership and possession coincide. Where they diverge (securities lending), the generalised position vector of \S\ref{sec:sec16} extends the balance to six stored coordinates, with $\mathrm{own}$ giving economic ownership and $\mathrm{avail}=\mathrm{own}-\mathrm{onloan}+\mathrm{borr}$ computed on read; the non-lendable case degenerates to the scalar. The Single-Coordinate Move Principle governs every state change: one move, one coordinate, one unit, two entities; multi-coordinate operations are transactions.

The architecture rests on five concepts---wallet, unit, move, transaction, contract---and one invariant. Every financial event is a set of quantity transfers between wallets; the total quantity of each unit is conserved. PnL, valuation, dual pricing, time travel, lifecycle management, CDM integration, property-based testing, and securities lending all follow. Product complexity is isolated in contracts; the ledger engine is invariant under what is traded.

\subsection{The Four Design Goals}

\begin{description}[nosep]
\item[Auditability.] Every balance change carries a provenance chain. The move stream is the complete audit trail; balance-sheet substantiation at any date is a deterministic computation.
\item[Reproducibility.] Pure lifecycle functions and the immutable log reconstruct any historical state and re-execute any lifecycle logic to identical results, under identical versions of market data, reference data, and pricing models. Reproducibility requires version-pinning of these external dependencies (\S\ref{sec:scope-limitations}).
\item[Time travel.] \texttt{clone\_at($t$)} reconstructs the ledger at any past date. With pure lifecycle functions, this is a strict notion of historical truth.
\item[Encoded correctness.] Conservation is algebraic, not checked. Typed unit state makes illegal lifecycle transitions structurally unreachable. Invariants are the specification language; property-based tests verify them.
\end{description}

\subsection{Self-Consistency as the Standard}

Internal reconciliation---comparing two independent records of one reality within the firm---is unnecessary for activity in the ledger. Balance sheet, PnL, position reports, and regulatory filings are projections of one canonical record. The most damaging failures---trading/GL divergence, PnL/BS mismatch, inter-entity breaks---are structurally unreachable within scope.

The standard is self-consistency: internally inconsistent states are unreachable by construction, not detectable after the fact. Where a failure depends on an external actor and cannot be eliminated, it is made detectable and narrowly scoped. The reconciliation-failure taxonomy in Appendix~\ref{app:reconciliation} records this assessment per failure mode. External reconciliation---against custodians, counterparties, and regulatory authorities---remains necessary at the ledger boundary.

\subsection{Resolved Design Problems}

Three problems intrinsic to an event-sourced post-trade ledger are resolved; they are stated here only to close them.

\begin{description}[nosep]
\item[Concurrency.] Resolved by the single-writer workflow model (\S\ref{sec:sec14}). Each lifecycle workflow instance serialises events for its unit; cross-unit coordination is handled by parent workflows; the executor never receives concurrent mutations for one unit.
\item[Liveness.] Resolved in two stages (\S\ref{sec:sec14}): deterministic-date obligations by durable timers, event-triggered obligations (SBL recalls, collateral substitution, CSA margin, close-out deadlines) by the obligation as a first-class object with invariants P21--P23. The guarantee covers all obligation types. Durable timers (liveness) composed with the idempotency chain (safety) yield exactly-once lifecycle processing.
\item[State attachment.] The unit-versus-wallet question of where unit state lives is resolved by the three-home state model (\S\ref{sec:sec04}): unit state is a materialised projection of the immutable event log---a read cache the log always rebuilds---not an authoritative mutable store.
\end{description}

\subsection{Open Problems}

The following are design-level questions, not implementation details, and must be resolved before production. Repos and access control can be addressed incrementally.

\begin{description}[nosep]
\item[Netting and close-out.] Atomic moves are gross. Netting is a higher-level operation that inspects pending gross moves, computes net amounts per $(\text{counterparty},\text{unit})$ pair, and emits reduced net moves; conservation holds by construction, the net moves forming a transaction whose signed edge-sum is zero. The algebra of netting---commutativity, associativity, interaction with CCP default management---is unformalised, particularly for multilateral netting through CCPs. Close-out netting under the ISDA Master Agreement is a mass lifecycle event over many contracts; the per-contract model requires a cross-contract close-out mechanism that terminates multiple contracts, computes close-out values, and emits one net settlement move.
\item[Correction algebra.] Compensating transactions are described informally. A formal algebra of corrections (anti-moves, amendment chains, composition rules) would make corrections first-class objects.
\item[Multi-entity federation.] The design assumes one shared ledger. Multi-entity deployments require a federation protocol preserving conservation across ledger boundaries; inter-company eliminations for consolidated reporting are unaddressed.
\item[XVA.] CVA, DVA, and FVA are non-transfer value changes that do not decompose into wallet-to-wallet moves and may require a dedicated adjustment layer (valuation mechanics: valuation specification).
\item[Bitemporal semantics.] Economic and booking timestamps are distinguished informally. A full bitemporal model would handle market-data restatements without corrupting the historical record (market-data: market-data specification).
\item[Tax-lot tracking.] Wallets store net quantities. US wash-sale and specific-identification reporting require per-lot tracking (date, quantity, cost basis) as move-stream metadata, with a lot-selection method (FIFO, specific identification). Its interaction with the futures accumulated-cost method (\S\ref{sec:sec08}), which deliberately avoids per-trade tracking, is unresolved.
\item[Impairment and ECL.] IFRS~9 expected credit losses and IAS~36 impairment reduce carrying value without a counterparty transfer; these may require a virtual impairment-reserve wallet or an extension of the move primitive.
\item[Data privacy and GDPR.] The append-only log conflicts with GDPR Article~17 (erasure) when moves carry personal data. Candidate resolutions: pseudonymous identifiers linked to a separate mutable store; crypto-shredding; or a schema in which personal data never enters the log. Crypto-shredding trades against time travel---\allowbreak\texttt{clone\_at(\allowbreak$t$)} returns anony\-mised counterparty references once keys are destroyed, acceptable for financial replay, incomplete for audit requiring counterparty identification. The choice fixes the move schema and must be made at design time (data: data specification).
\item[Access control and actor attribution.] The move primitive carries no authenticated actor identity. Production requires per-move attribution to an authorised actor, maker-checker workflows, cryptographic non-repudiation, and wallet-level read/write permissions scoped to role and organisational boundary.
\item[Repos.] Repurchase agreements (GMRA) share collateral and settlement primitives with SBL (\S\ref{sec:sec16}) but differ in title-transfer semantics. Extension requires the GMRA-specific lifecycle events (open leg, close leg, repricing, substitution, classic repo versus buy-sell back).
\item[Default management.] Counterparty or clearing-member default requires close-out, portfolio auction, and loss allocation. Conservation must accommodate loss mutualisation---transfer of losses to surviving participants---which needs governance and waterfall rules the framework does not provide.
\item[Pre-trade validation.] The framework is silent on pre-trade checks (credit limits, position limits, margin sufficiency, regulatory constraints). A pre-condition gate on the move primitive is a natural extension point; its interaction with the conservation law is unspecified.
\item[Migration from legacy architectures.] Adoption requires migrating from separate trading systems, risk engines, and general ledgers---parallel running with reconciliation during transition, progressive routing of new trades, back-loading of historical positions. The political economy of the transition is as consequential as the technical path.
\item[Tokenised securities.] Appendix~\ref{app:cdm-walkthrough} establishes the tokenised-equity model, the double-counting resolution, and the custodian-is-flat principle, and identifies four CDM v6.0.0 gaps. Open: cross-chain interoperability; partial backing and proof-of-reserves; regulatory classification under MiFID~II and SEC frameworks; tokenised-collateral eligibility; and the interaction of on-chain governance with custodian-mediated shareholder voting.
\end{description}

\subsection{Flagged Open-Items Register}

The items below are honest open items---the next-version agenda---kept distinct from the proven quantity algebra. They are migration, governance, operational, and conformance risks; none invalidates the design on correctness grounds. Each escalation is also cited at its home section: futures in \S\ref{sec:sec08}, managed-account in \S\ref{sec:sec09}.

\subsubsection{State-Model Adoption Risk Register (F1--F8)}

These accompany the three-home state model (\S\ref{sec:sec04}). F2, F5, F6 require external stakeholders before any code ships; F1, F3, F4, F7, F8 are internal engineering risks manageable within the delivery programme.

\begin{longtable}{@{}p{0.5cm}p{6.3cm}p{6.3cm}@{}}
\toprule
& \textbf{Risk} & \textbf{Mitigation} \\
\midrule
\endhead
F1 & Migration scope of retrofitting every downstream consumer that reads unit state as a single combined field. & Ship a read-compatibility accessor exposing the composed single-field view; split additively before cutover; parallel-run $\geq$ one quarter with reconciliation. \\
F2 & Ownership of the C8 fungibility predicate is ungoverned. & Per-product RACI in the Unit Store registry; Legal/Product/Risk sign-off for predicate changes; version the predicate. \\
F3 & Monotone carrier key-space growth at $10^7\times10^6$ scale is unbenchmarked. & Benchmark Option-accessor cost and cache behaviour before merge; tombstone-compaction policy for rows \texttt{Some(zero)} beyond the retention horizon. \\
F4 & C1--C12 is a multi-sprint test programme, not a release. & Land the schema with the type-enforced conditions first; stage C2/C3/C8/C11 over later releases; never merge below C1+C2+C3. \\
F5 & Mandate-as-unit creates an SFTR/EMIR reporting surface for mandate issuance. & Pre-flight with Regulatory; determine whether $u_{\mathrm{MA}}$ issuance needs a UTI/LEI pair; possible \texttt{reportable} opt-out flag. \\
F6 & CDM alignment (\texttt{TradeState} per \texttt{Trade} vs $\texttt{PositionState}[w,u]$) is asserted, not verified. & Rerun the Rosetta NS1--7 mapping against the three-map schema before any CDM adapter work; publish a delta document. \\
F7 & Onboarding cost of three maps, Option/Monotone carriers, and twelve conditions is real. & Ship a decision tree, runbook, and cheat sheet; mandatory training before handler work. \\
F8 & Forward migration is irreversible once the pre-migration combined wallet-state representation is collapsed into the three maps. & Keep a read-only mirror of that pre-migration representation $\geq$ four quarters post-cutover; document the inverse mapping explicitly. \\
\bottomrule
\end{longtable}

\subsubsection{Testing Commitments (Mutation Score)}

Concrete targets, to be met before production. The generator universe---CDM enum $\times$ product-type cross-product---is finite and enumerable, so coverage is structurally bounded.

\begin{itemize}[nosep]
\item Event handlers (decimal arithmetic): 85--90\,\%. Conservation kills most sign and coefficient mutants.
\item Lifecycle guards: 70--80\,\%. Boundary-comparison mutants ($>$ versus $\geq$) require targeted tests.
\item Overall core: $\geq 80\,\%$ (the Feathers change-safety threshold).
\item State-machine model at $|W|=3$, $|U|=2$, depth $\leq 6$: $10^5$--$10^6$ reachable states (TLC-tractable); conservation-violating handlers are caught at depth~1.
\end{itemize}

\subsubsection{Managed-Account Escalations (ME1--ME5)}

Open items from \S\ref{sec:sec09}, distinct from the segregation theorem (CONS $\wedge$ LOC $\wedge$ C4) proved there.

\begin{longtable}{@{}p{0.5cm}p{12.6cm}@{}}
\toprule
& \textbf{Open item} \\
\midrule
\endhead
ME1 & Store-versus-derive for account-level scalars (NAV, high-water mark): which are stored and which recomputed on read is unsettled. \\
ME2 & The observation surface is unattested (Nazarov attestation-envelope finding): the external observations driving mandate evaluation carry no cryptographic attestation; an attestation envelope is required. \\
ME3 & Counterparty legal identity: the mapping from client to a governed LEI is outside the ledger and must be bound. \\
ME4 & Localisation condition C4 is asserted in the segregation argument, not yet discharged by a type. \\
ME5 & No in-ledger solvency liveness: continuous solvency monitoring of the managed account is not a framework guarantee. \\
\bottomrule
\end{longtable}

\subsubsection{Futures Escalations (FE1--FE2)}

Open items from \S\ref{sec:sec08}, distinct from the futures lifecycle and its closing identity (cumulative VM $=$ economic PnL) proved there.

\begin{longtable}{@{}p{0.5cm}p{12.6cm}@{}}
\toprule
& \textbf{Open item} \\
\midrule
\endhead
FE1 & Variation-margin settlement fan-out cost at scale is unbenchmarked. \\
FE2 & The derived-consequence alternative to storing accumulated cost per position (C11) was declined; the trade-off is recorded, not closed. \\
\bottomrule
\end{longtable}

\noindent Concurrency, liveness, and state attachment are resolved (\S\ref{sec:sec14}, \S\ref{sec:sec04}). The items in this register and in Open Problems are not. The quantity algebra is sound within the boundary of \S\ref{sec:scope-limitations}; the open items bound what the framework does not yet establish.

\clearpage
%==============================================================================
