\documentclass[11pt,a4paper]{article}

% --- encoding + clean PDF text extraction (ligature fix, proven) ---
\usepackage{cmap}                 % must precede fontenc
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}              % must follow fontenc
\input{glyphtounicode}\pdfgentounicode=1
% small-caps has no italic shape in Computer Modern; substitute upright small-caps
% silently (renders identically) so \textsc inside italic theorem/formula contexts
% raises no font-substitution warning. Deferred so the lmodern family is declared first.
\AtBeginDocument{%
  \DeclareFontShape{T1}{lmr}{m}{scit}{<->ssub*lmr/m/sc}{}%
  \DeclareFontShape{T1}{lmr}{bx}{scit}{<->ssub*lmr/bx/sc}{}%
}

% --- math, tables, code, graphics ---
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{enumitem}
\usepackage{listings}
\usepackage{longtable}
\usepackage{textcomp}
\usepackage{array}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,calc}
\usepackage[hidelinks]{hyperref}

\geometry{margin=1in}
\setcounter{tocdepth}{2}
% last-resort stretch so tough paragraphs (long typewriter identifier runs) break
% without running into the margin, instead of an overfull box.
\setlength{\emergencystretch}{3em}
\newcommand{\EUR}[1]{\texteuro#1}

% --- theorem-like environments ---
\newtheorem{definition}{Definition}[section]
\newtheorem{principle}[definition]{Principle}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{invariant}[definition]{Invariant}
\newtheorem{remark}[definition]{Remark}

% --- Haskell listings ---
\lstset{
  basicstyle=\ttfamily\small,
  columns=fullflexible,
  keepspaces=true,
  xleftmargin=2em,
  aboveskip=0.5em,
  belowskip=0.5em,
  language=Haskell,
  showstringspaces=false,
  breaklines=true
}

\title{The Ledger: A Closed-System Framework for Financial Settlement\\[6pt]
\large Unified Design Specification: Architecture, Lifecycle, and CDM Integration}
\author{R.\ Delloye}
\date{June 2026}

\begin{document}
\maketitle

\input{drafts/front_preface}

\begin{abstract}
Within its scope, the post-trade activity addressed by this framework reduces to one operation: the
atomic move of a quantity between wallets. From this single primitive the Ledger builds a unified system
in which financial instruments---equities, bonds, derivatives, swaps, managed accounts, Quantitative
Investment Strategies, listed futures---are deterministic contracts that emit sequences of moves. One law
governs the system: conservation, which holds by construction. From it follow path-independent profit and
loss, complete time travel over an immutable event log, and idempotent lifecycle processing through pure
functions, and several classes of internal reconciliation failure become structurally unreachable. Unit
state lives in three homes---immutable \texttt{ProductTerms}, the shared \texttt{UnitStatus}, and
per-position \texttt{PositionState}; \texttt{UnitStatus} is a materialised projection of the event log,
not an authoritative mutable store. The scalar balance suffices for buy/sell/exercise workflows; for
securities lending it extends to a six-coordinate position vector. Throughout, the design is illustrated
in Haskell, with types chosen so that illegal states are unrepresentable. This design
specification presents the three-home state model, the \texttt{UnitStatus} characterisation,
the futures lifecycle, and the managed-account workflow as one self-contained whole.
\end{abstract}

\tableofcontents
\clearpage

%==============================================================================
% FRONT MATTER — navigation aids (after the ToC, before Part I)
%==============================================================================
\input{drafts/front_roadmap}
\input{drafts/front_concepts}
\clearpage

%==============================================================================
% BODY — organised into parts (see coverage_map.md)
%==============================================================================
\part{Foundations: The Closed System and Conservation}
\input{drafts/sec01}
\input{drafts/sec02}
\input{drafts/sec03}
\input{drafts/part1_summary}

\part{The Three-Home State Model and \texorpdfstring{\texttt{UnitStatus}}{UnitStatus}}
\input{drafts/sec04}
\input{drafts/part2_summary}

\part{Valuation, Contracts, and Instrument Lifecycles}
\input{drafts/sec05}
\input{drafts/sec06}
\input{drafts/sec07}
\input{drafts/sec08}
\input{drafts/part3_summary}

\part{Accounts, Reporting, Settlement, and CDM Integration}
\input{drafts/sec09}
\input{drafts/sec10}
\input{drafts/sec11}
\input{drafts/sec12}
\input{drafts/sec13}

\part{Obligations, Generalised Positions, and Securities Lending}
\input{drafts/sec14}
\input{drafts/sec16}
\input{drafts/part5_summary}

\part{Formal Properties, Limitations, and Context}
\input{drafts/sec15}
\input{drafts/sec17}
\input{drafts/sec18}
\input{drafts/sec19}
\input{drafts/sec20}

%==============================================================================
% APPENDICES
%==============================================================================
\appendix
\part{Appendices}
\input{drafts/appA}
\input{drafts/appB}
\input{drafts/appC}
\input{drafts/appD}
\input{drafts/appE}
\input{drafts/appF}
\input{drafts/appG}
\input{drafts/appH}
\input{drafts/appI}

%==============================================================================
% REFERENCES (only the externally-cited regulatory sources; other external
% mentions are pinned inline as footnotes in house style)
%==============================================================================
\begin{thebibliography}{9}
\bibitem{isla-ibp} International Securities Lending Association, ``ISLA Best Practice Handbook,'' various editions.
\bibitem{sec15c33} U.S.\ Securities and Exchange Commission, ``Rule 15c3-3: Customer Protection---Reserves and Custody of Securities,'' 17~CFR \S240.15c3-3.
\bibitem{regsho} U.S.\ Securities and Exchange Commission, ``Regulation SHO, Rule 203(b)(1): Locate Requirement,'' 17~CFR \S242.203(b)(1).
\bibitem{sec10c1a} U.S.\ Securities and Exchange Commission, ``Rule 10c-1a: Reporting of Securities Loans,'' 17~CFR \S240.10c-1a, adopted January 2024.
\bibitem{finra6500} Financial Industry Regulatory Authority, ``FINRA Rule 6500 Series: Securities Lending and Transparency Engine (SLATE),'' 2024.
\end{thebibliography}

\end{document}
