% Chapter 2 — The Picture: Map, Then Fold. Author: KLEPPMANN. Budget 5 pp.
\section{The Picture: Map, Then Fold}\label{ch:picture}

This chapter discharges \S3 of the constitution.

Everything this framework does is a \emph{map} followed by a \emph{fold} --- two words
borrowed from programming, worth a line each. A \emph{map} processes a list element by
element: given a way to turn an \texttt{a} into a \texttt{b}, it carries \texttt{[a]} to
\texttt{[b]}, one element at a time. A \emph{fold} accumulates a list into a single
result: each element is absorbed into a running value, \texttt{[a]} to \texttt{b}. The
whole architecture is these two steps over one list. The world presents a list of
events, $E = [e]$; a map carries it to a list of transactions, $T = [tx]$; a fold
reduces that list to the ledger. Four terms complete the picture, each defined in
Chapter~\ref{ch:objects} and used here only at a gloss: a \textbf{unit} is anything that
can be held or owed; a \textbf{wallet} is where units are held; a \textbf{transaction} is
the atomic bundle of changes the ledger records; and a \textbf{projection} is any view
computed from the record. Every number the ledger reports --- a balance, a valuation, a
profit and loss, a report --- is a projection: a further fold of that same one list.

\subsection{The six threads}

Six instruments recur through this specification as threads. Their terms, parties, and
quantities are fixed here once and never vary, so a number met in one chapter is the same
number met in another. This chapter follows one of them --- the one-touch OT-1 ---
through the whole pipeline; the rest are worked where they bite.

\medskip
{\small
\noindent\begin{tabular}{@{}lp{4.1cm}p{9.4cm}@{}}
\toprule
 & Instrument, kind, parties & Fixed terms and headline numbers \\
\midrule
T1 & FUT-IDX --- cash-settled future on IDX; W-ALPHA / the CCP & multiplier 10, USD,
  settled-to-market; buy 1 at 995.00; settlement prints 1{,}000.00 / 990.00 /
  1{,}005.00; cumulative VM \USD{100} $=$ PnL \\
T2 & OT-1 --- one-touch on stock OMEGA; W-BANK writes, W-ALPHA holds & barrier 80.00
  touch-at-or-below, payout \USD{1{,}000{,}000}; spot 95.00 at inception; knock at close
  79.00 \\
T3 & ACME --- cash dividend; issuer / W-ALPHA & 1.50 per share; W-ALPHA holds 10{,}000
  shares; entitlement \USD{15{,}000} \\
T4 & ACME --- 2-for-1 split; issuer / W-ALPHA & 10{,}000 $\to$ 20{,}000 shares; price
  frame 100.00 $\to$ 50.00; declared dividend figure 1.50 $\to$ 0.75 \\
T5 & VS-1 --- OTC variance swap on IDX; bilateral & one year, 252 fixings, strike 400
  variance points, \USD{1{,}000} per point; realised variance 441; settlement
  \USD{41{,}000} \\
T6 & TRS-1 --- total return swap; reference W-QIS in virtual ledger V-1 (strategy S-QIS) &
  notional \USD{10{,}000{,}000}; financing 4\% p.a., quarterly; first reset NAV
  10{,}300{,}000; net move \USD{200{,}000} \\
\bottomrule
\end{tabular}
}
\medskip

\subsection{The world presents events}

The world presents the system with a list of events. Some arrive from outside: trades
executed in the market, corporate-action notices, fixings and other observations,
captured and recorded at the boundary. Others are not received but noticed. Every unit
declares, from its terms, the conditions its life requires watching --- its
\textbf{watches} (Chapter~\ref{ch:objects}) --- and the Event Monitor evaluates those
watches against the clock and the recorded observations, emitting an event onto the
record the moment a condition is met. The future FUT-IDX declares such watches at
registration: one for each daily settlement print, one for its expiry; the one-touch OT-1
declares one, its underlying's recorded close against the barrier. A coupon date
arriving, an expiry falling due, a close printing below a barrier --- each becomes an
event the same way a trade does, and once emitted the two kinds are indistinguishable:
both are on the record.

\subsection{The map: each event becomes a transaction}

Each event is mapped to a transaction, and the responsibility for the map divides in two.
The \textbf{smart contracts} are the mapping functions: each is a stateless piece of code
that, given one event and the current state of the ledger, computes the transaction the
event requires (Chapter~\ref{ch:contracts}). The \textbf{Events Executor} applies the map
over the list, $E=[e] \to T=[tx]$: for each event it fires the responsible contract,
hands it the event together with the current projection of the ledger, and collects the
proposed transaction. Many contracts, one Events Executor; the contracts carry all the
product knowledge, the Events Executor carries none (Chapter~\ref{ch:machines}).

The map is not blind to state. To turn a dividend event into cash moves, the contract
must read who holds the stock at that moment --- the fold of everything admitted so far.
Map and fold therefore advance together, one event at a time, in a single total order:
the transaction for event $e$ is computed against the ledger that the transactions for
all earlier events have already built.

\subsection{The fold: the transactions become the ledger}

The resulting list of transactions is folded --- reduced, one transaction at a time ---
into the state of the system: balances, and the three homes of unit and position state
(Chapter~6, \emph{State: The Three Homes}). The \textbf{Transaction Executor} performs the
fold, $T=[tx] \to \mathrm{Ledger}$: it serialises, validates, and applies each
transaction, and the immutable log is precisely the list being folded. Nothing else
writes. And because the log is one list, every other view is a further fold of it: a
balance sums the moves that touch a wallet, a valuation folds the composition against a
price vector (Chapter~\ref{ch:valuation}), a report folds it into a presentation
(Chapter~\ref{ch:reporting}). One record, and many folds of it --- never many records.

\subsection{Data, functions, and machines}

Three kinds of thing appear in this picture, and they are never confused: data,
functions, and machines. The data are Event, Transaction, and Ledger. The functions are
the arrows between the data. The machines are what evaluate the arrows --- one machine per
arrow, and they are the subject of Chapter~\ref{ch:machines}. Typed:

\begin{lstlisting}
watch    : (Record, Clock)       -> [Event]      -- the Event Monitor
contract : (Event, Ledger)       -> Transaction  -- the Events Executor
apply    : (Transaction, Ledger) -> Ledger       -- the Transaction Executor

step (ledger, e) = apply (contract (e, ledger), ledger)  -- map, then fold
final ledger     = fold step over E = [e]
\end{lstlisting}

The three arrows differ in what they read. \texttt{watch} reads the Record and the clock
and yields events; \texttt{contract} and \texttt{apply} consume the Ledger --- the folded
state --- and never the clock. One \texttt{step} is a map (fire the contract) immediately
followed by a fold (apply the result), and the whole run is that step folded over
$E=[e]$. Map, then fold: the phrase is the architecture, not a metaphor for it.

\subsection{The clock, and proposal versus authority}

Two properties of the picture carry most of its weight.

\emph{The clock is the single input that is not on the record}, and the Event Monitor is
the single place it is consulted. The event the Monitor emits is itself recorded, so
everything downstream of the emission is a function of the record alone --- which is
exactly why a late firing produces the identical transaction, and why, given the record
and the clock, any party can recompute which events should have been emitted, and when
(Chapter~\ref{ch:machines}).

\emph{The map is proposed; the fold is authoritative.} A contract's output is a proposal,
not a fact: a transaction joins the list only when the Transaction Executor admits it. The
smart contracts, the Event Monitor, and the Events Executor sit outside the trust boundary
and may only propose; the one Transaction Executor writes. Either arrow may refuse --- a
contract may find nothing due; the door rejects a malformed transaction and writes nothing
--- and a refusal is a returned value, never a half-applied state. Because both the events
and the transactions are on the record, economic correctness is checkable after the fact:
re-run the map on the recorded event and compare the result against the recorded
transaction (Chapter~15, \emph{Testability and the Executable-Check Regime}). Verification
lives with the events; authority lives with the transactions.

\subsection{One instrument walks the pipeline}

The one-touch OT-1 pays \USD{1{,}000{,}000} if stock OMEGA ever trades at or below the
barrier 80.00; its spot at inception is 95.00. It is written by \mbox{W-BANK} and held by
\mbox{W-ALPHA}, and at registration its terms declared one watch: OMEGA's recorded
official closes against the barrier.

\textbf{Trigger (watch).} OMEGA's official close of 79.00 is recorded. The declared
condition --- at or below 80.00 --- is met, and the Event Monitor emits the touch event
onto the record. No ledger state has changed: the Monitor emits events, and an event is an
input to the map, not a write.

\textbf{Contract (map).} The Events Executor fires OT-1's contract, handing it the touch
event and the current projection of the ledger. The contract reads the terms (payout
\USD{1{,}000{,}000}), the holder (\mbox{W-ALPHA}), and the unit's state (\emph{live}), and
returns one transaction:

\begin{center}
\begin{tabular}{@{}rll@{}}
\toprule
 & Kind & Content \\
\midrule
1 & unit-state change & OT-1: UnitStatus \emph{live} $\to$ \emph{triggered} \\
2 & move & owned(USD) 1{,}000{,}000: W-BANK $\to$ W-ALPHA \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Apply (fold).} The Transaction Executor admits it --- the move's paired legs leave
conservation undisturbed, the cause-derived identifier names the touch event, and the
writer is OT-1's own contract --- and the fold advances: the ledger now shows the cash with
\mbox{W-ALPHA} and the option \emph{triggered}. Where OT-1 stands encumbered under a
collateral agreement the payment leg routes differently, and Chapter~\ref{ch:collateral}
works that same knock start to finish; the pipeline is the one drawn here either way.

\emph{One list of events, one map, one fold: the option's whole knock is a single event
carried across the three arrows, and nothing off the record ever touches it.}

\subsection{Not event sourcing}

The resemblance to event sourcing is real, and the difference matters. The Ledger is
event-driven because finance is: trades execute, prices print, coupons mature, and
barriers are touched whether or not the software cares, so events are recorded because
reality presents them, not because a pattern was chosen. The difference is where authority
sits. In event sourcing the event stream is the canonical history, state is whatever the
current code derives from it, and reinterpreting history through new code is a feature.
Here events are inputs to the map, never the accounting record: contracts transform them
into proposed transactions, the Transaction Executor admits or refuses --- an admission
door event sourcing has no equivalent of --- and the admitted transaction log alone is the
canonical history. Replay re-reads transactions and never re-executes contracts
(Chapter~\ref{ch:machines}), so history cannot be silently reinterpreted; a wrong
interpretation is repaired by a compensating transaction, in the open (Chapter~15). And
every view is a projection of that one log, not a fold over per-aggregate streams stitched
back together: there is a single total order, and two projections of it cannot disagree.
The events stay on the record for one purpose --- so the map can be re-run and verified.

\begin{secondtelling}
A fold with a fixed step is a \emph{catamorphism}: the unique map out of the list $E=[e]$
determined by one rule for the empty list and one rule for prepending an element --- here
\texttt{step}. Because the step is fixed and the list carries its own order, the ledger is
that unique image, and every projection --- balance, valuation, report --- is another
catamorphism over the same list. ``Map, then fold'' names the composite; that the fold is
a catamorphism is why one record admits many views and no two of them can disagree.
\end{secondtelling}
