% Chapter 11 — The Settlement Interface. Author: KLEPPMANN. Budget 5 pp.
% Folds MINSKY's D2 insert (settlement state and the market claim) into KLEPPMANN's voice.
% Seam (for FORMALIS): MC-1 is created by the record-date contract firing, not by the
% settlement projection, which writes nothing. Numbers per the frozen thread map.

\section{The Settlement Interface}\label{ch:settlement}

This chapter discharges the constitution's Scope commitment to the
settlement-projection layer; it cites \S3, which makes every view a projection of
the record, and \S4, which fixes what the record is.

\subsection{Settlement as a projection}

The settlement interface is where the closed ledger meets the open infrastructure
of settlement. It is a projection like any other (\S3): a function of the committed
log that reads the moves the ledger has admitted and emits \emph{settlement
instructions}. The division of labour is exact --- the ledger records \emph{what}
settles; the external infrastructure decides \emph{how}, and how is out of scope
(settlement finality, payment-system access, and central-securities-depository
connectivity are external authorities the ledger reconciles against, never
functions it performs). An instruction is the boundary object: it names the
transfer to be performed and carries none of the machinery that performs it.

What the interface never does is write. It is not the Transaction Executor and
holds none of its privilege; it proposes nothing to the door and changes no
balance. The single-writer discipline (\S5) is not bent here --- the interface
reads and emits, and only the Transaction Executor writes. Every fact the interface
consumes is already on the record, and every fact it produces is an instruction,
never a move. Four properties follow, and each is load-bearing.

\begin{itemize}[nosep]
\item \emph{Pure.} It reads the recorded log alone --- no clock, no call to the
outside world --- so any party recomputes the same instructions from the record
(\S2, \S3).
\item \emph{Total.} It is defined on every committed transaction. A transaction
that carries no boundary-crossing move --- a moveless corporate-action component, a
holding in a virtual ledger (\S10) --- has the empty instruction list as its image,
by construction, not by a special rule.
\item \emph{Deterministic.} The same log yields the same instructions, always.
\item \emph{Idempotent.} Because it is a function of the recorded log and the
recorded settlement state, re-running it produces the same instructions, and an
instruction whose transfer the record already shows as settled is recognised and
not performed twice.
\end{itemize}

\begin{lstlisting}
settle :: Ledger -> [Instruction]   -- pure, total, deterministic, idempotent; writes nothing

data Instruction                    -- the boundary object: what settles, never how
  = Cash     WalletId WalletId UnitId MinorUnits  -- from, to, currency, amount
  | Delivery WalletId WalletId UnitId Quantity     -- from, to, security, quantity
\end{lstlisting}

One consequence is stated once and relied on throughout: \emph{settlement failure
is a recorded event, never a reversal}. A settlement that does not complete returns
as an external event, recorded as a new fact; the instruction's underlying move is
not undone. History is never rewritten (\S2): a trade ultimately unwound is unwound
by a compensating transaction, through the same door, never by deleting the booking
that instructed it.

\subsection{Three episodes}

\textbf{Dividend: cash projects to one instruction.} W-ALPHA holds $10{,}000$ ACME;
the declared dividend is $1.50$ per share, and nothing encumbers the holding. The
payment-date firing has already produced its transaction
(Chapter~\ref{ch:collateral}): the conditional payment obligation discharges
vacuously, and one move crosses the boundary --- owned(USD) \USD{15{,}000}, ACME's
issuer wallet $\to$ W-ALPHA.
\emph{The projection} reads that committed move.
\emph{The instruction:}
\begin{enumerate}[label=(\arabic*),nosep]
\item Cash \USD{15{,}000}, ACME issuer $\to$ W-ALPHA.
\end{enumerate}
The moveless component of the same transaction --- the obligation's creation and
vacuous discharge --- crosses nothing and projects to the empty list.
\emph{The door} is not the interface's: the move already passed the Transaction
Executor, and the projection only reads what was admitted.
\emph{The ledger determined and recorded the entitlement; the interface instructs
only the cash that leaves the ledger's edge.}

\medskip\noindent\textbf{Future: each day's margin, and the expiry unwind.} FUT-IDX
is settled-to-market, multiplier $10$; W-ALPHA is long one contract bought at
$995.00$. On each settlement print the future's contract fires and emits the day's
variation-margin move (Chapter~\ref{ch:collateral}); the interface projects each to
one instruction, and the expiry's final settlement to a settlement instruction that
also closes the contract.

\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
Recorded print & Variation margin & Committed move & Instruction \\
\midrule
day 1 \ $1{,}000.00$ & $+50$ & owned(USD) $50$, W-CCP $\to$ W-ALPHA & Cash $50$ to W-ALPHA \\
day 2 \ $990.00$ & $-100$ & owned(USD) $100$, W-ALPHA $\to$ W-CCP & Cash $100$ to W-CCP \\
day 3 \ $1{,}005.00$ & $+150$ & owned(USD) $150$, W-CCP $\to$ W-ALPHA & Settlement $150$ to W-ALPHA \\
\bottomrule
\end{tabular}
\end{center}

\noindent At expiry the position unwinds: the final settlement of \USD{150}
projects to the settlement instruction above, while the unit-state change that
retires the contract carries no move and projects to nothing.
\emph{The door} again belongs to the Transaction Executor; the interface reads
three admitted moves and emits three instructions.
Cumulative variation margin is $50-100+150 = 100$, the total profit and loss
(Chapter~\ref{ch:valuation}) --- but the interface never sees the profit and loss,
only the cash that crosses each day.
\emph{The ledger nets the day to a single move; the interface settles exactly that
move, and no valuation reaches the boundary.}

\medskip\noindent\textbf{Split: the moveless corporate action projects to nothing.}
ACME splits two-for-one, effective 2026-06-01. One transaction carries the
re-denomination and the new terms (Chapters~\ref{ch:objects} and~\ref{ch:marketdata}):
\begin{enumerate}[label=(\arabic*),nosep]
\item move owned(ACME) $10{,}000$, ACME issuer $\to$ W-ALPHA --- W-ALPHA's
$10{,}000$ shares become $20{,}000$;
\item a new ProductTerms version: price frame $100.00 \to 50.00$, with the market
data operators declared (spot scales by $\tfrac12$; the cash dividend figure scales
$1.50 \to 0.75$; proportional figures stand).
\end{enumerate}
\emph{The projection.} Component~(2) carries no move; its image is the empty list by
the projection's totality. Component~(1) is the ledger mirroring a re-denomination
the external register performs of its own accord as the corporate action; the
interface issues no delivery or payment instruction for it, because a split settles
nothing between counterparties.
\emph{The door} admitted a well-formed transaction whose consistency of reference
is checked in Chapter~\ref{ch:invariants} ($20{,}000$ against the fresh $50.00$,
never the stale $100.00$); the interface adds no check and emits nothing.
\emph{What crosses the boundary is a transfer; a corporate action that
re-denominates terms crosses nothing, so it projects to nothing --- by
construction.}

\subsection{Settlement state and the market claim}

One recorded fact upstream of the projection is load-bearing for its correctness:
the settlement state of a trade. A determination turns on it --- who receives a
distribution when a trade is instructed but unsettled across a record date --- and
the fact must be read, not assumed away.

\medskip\noindent\textbf{Owned re-books at instruction.} Title moves on the
instruction event, on the record (\S4), not on settlement confirmation --- the
trade-date booking of Chapter~\ref{ch:collateral}. The instant a trade is
instructed the owned plane shows the buyer; the settlement instruction the
interface projects is the promise that the external register will catch up. Between
the two lies the instruction-to-settlement gap, and the gap is carried explicitly
--- never elided, never back-filled at read time.

\medskip\noindent\textbf{A record date inside the gap.} Entitlement determination
reads the owned plane (Chapter~\ref{ch:collateral}): who owns a unit at the record
date is who is owed its distribution. But the paying agent credits the \emph{record
holder} at the external register, and while a sale is unsettled the record holder is
still the deliverer --- the seller --- not the buyer the owned plane already shows.
The two disagree by construction, and the disagreement is not a break to be
reconciled at report time: it is a recorded obligation. The \textbf{market claim} is
that obligation --- a unit (\S6) whose obligee holds it positive and whose obligor
holds it negative, the buyer's claim-for-equivalent on one side and the deliverer's
duty to remit on the other, created over the gap by the record-date firing and
discharged when the deliverer receives the distribution. Here is the seam that keeps
the projection pure: the market claim is created not by the projection, which writes
nothing, but by the contract that fires at the record date. That contract reads the
owned plane and the recorded settlement state and proposes the reconciling leg as
one component of the determination transaction, which the Transaction Executor
admits like any other. The projection then reads the resulting committed moves; it
originates none of them.

\begin{lstlisting}
data SettlementState = Instructed | Settled | Failed  -- a recorded fact, read by routing

-- The market claim is a unit (S6): the obligee holds it +, the obligor -.
-- Conservation makes the two signed sides one fact.
data MarketClaim = MarketClaim
  { distribution  :: UnitId        -- what the deliverer receives and owes onward
  , parAmount     :: MinorUnits    -- valued at par; the deliverer's owned value nets to zero
  , dischargePred :: Predicate     -- satisfied by the deliverer's receipt of the distribution
  }
\end{lstlisting}

\begin{proposition}[Entitlement routing]
The recipient of a distribution is a total function of the recorded owned plane and
the recorded settlement state at the record date. Determination reads owned; where
the settlement state of an owned re-booking is unsettled across the record date, the
paying agent credits the deliverer and a market-claim unit carries the reconciling
leg from the deliverer to the owner. No distribution reaches any wallet except by a
recorded move against a determined entitlement or the discharge of a market-claim
unit.
\end{proposition}

\medskip\noindent\textbf{Worked micro-example (dividend).} Stock ACME pays a cash
dividend of $1.50$ per share; the record date is 2026-05-15. Wallet W-ALPHA holds
$10{,}000$ ACME and sells them to a buyer, wallet W-BETA: the sale is instructed on
2026-05-14 for settlement on 2026-05-16, so the record date falls in the gap. (The
cash consideration settles symmetrically and does not bear on the entitlement.)

\emph{2026-05-14 --- the sale; owned re-books at instruction.} One transaction:
\begin{enumerate}[label=(\arabic*),nosep]
\item move owned(ACME) $10{,}000$, W-ALPHA $\to$ W-BETA --- re-booked at
instruction; the settlement instruction, due 2026-05-16, is recorded as unsettled.
\end{enumerate}
The owned plane now shows W-BETA holding $10{,}000$ ACME and W-ALPHA holding none.

\emph{2026-05-15, record date --- determination and the market claim.} The
record-date watch fires; the stock's contract determines the entitlement and,
reading the recorded settlement state, proposes the reconciling leg. One
transaction:
\begin{enumerate}[label=(\arabic*),nosep]
\item determination reads the owned plane --- W-BETA owns $10{,}000$ --- and records
the entitlement, $10{,}000 \times 1.50 = \USD{15{,}000}$, to W-BETA in PositionState;
\item settlement state reads unsettled across the record date: the paying agent will
credit the record holder, the deliverer W-ALPHA. Issue the market-claim unit MC-1,
valued at par: move owned(MC-1) $1{,}500{,}000$ (minor units, $=\USD{15{,}000}$),
W-ALPHA $\to$ W-BETA --- W-ALPHA holds $-1{,}500{,}000$ (the duty to remit once
received), W-BETA holds $+1{,}500{,}000$ (the claim-for-equivalent). MC-1's discharge
predicate is the deliverer's receipt; the obligation is contingent on it and is
matched by the deliverer's record-holder receivable, so the conduit's owned value is
neutral. While W-ALPHA performs the claim is valued at par, and it decouples to a
recovery claim on W-ALPHA's default.
\end{enumerate}
This transaction is admitted through the door; MC-1 exists because the record-date
firing wrote it, not because the projection did.

The sale settles on 2026-05-16, recorded as an external confirmation; this updates
the register but not the record-date snapshot that governs the distribution --- the
paying agent still credits the record holder as of 2026-05-15.

\emph{2026-05-22, payment date --- the distribution and the discharge.} The
payment-date watch fires. One transaction, two moves:
\begin{enumerate}[label=(\arabic*),nosep]
\item move owned(USD) \USD{15{,}000}, ACME issuer wallet $\to$ W-ALPHA --- the paying
agent credits the record holder;
\item move owned(USD) \USD{15{,}000}, W-ALPHA $\to$ W-BETA --- the market claim
discharges under its predicate, and MC-1 retires at the zero vector.
\end{enumerate}
W-ALPHA nets to zero across the two moves --- a conduit, not an owner; W-BETA ends
with the \USD{15{,}000} it economically owns. \emph{The door}: conservation per
(unit, coordinate) on every paired-leg move; MC-1 retires only from the zero vector;
idempotence --- a duplicate payment-date event finds MC-1 discharged and proposes
nothing. Each of these two cash moves is what the interface then projects to a cash
instruction; the market claim itself never crosses the boundary --- it is the
internal record that keeps the two boundary-crossing payments consistent.

\medskip\noindent\textbf{Settlement failure is a recorded event, never a reversal.}
Should the sale fail to settle on its due date, the failure returns as an external
event and is recorded as a new fact; the instruction's owned re-booking is not undone
and the market claim is not withdrawn. The market claim, an obligation unit, carries
a deadline, a discharge predicate, and a fallback (\S14.1); by its deadline it is
discharged, compensated, or declared defaulted --- a failed distribution is a
visible, overdue item, never a silent balance edit.

\medskip\noindent\textbf{Load-bearing, not a reporting axis.} Settlement state is
therefore a correctness input, not a presentation dimension. A design that read owned
alone would instruct the venue to pay the buyer directly --- an instruction it cannot
honour, since the buyer is not yet the record holder --- and the cash would arrive at
the deliverer with nowhere on the record to go, resurfacing as exactly the
reconciliation break the ledger exists to abolish. Reading settlement state, and
recording the market claim as a first-class unit, makes the routing both faithful to
what the venue will do and reproducible by any party from the record alone (\S2,
\S3). The construction is symmetric: where an owned re-booking runs the other way
across the record date, the same unit carries the reverse claim with its signs
reversed --- one mechanism, total over the gap.

\emph{The ledger says what settles and, across the gap, who is owed; the interface
projects that settled activity into instructions, writing nothing and inventing
nothing.}
